---
title: "Are these draws calibrated, and do the engines agree?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Are these draws calibrated, and do the engines agree?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. Three wild-bootstrap goodness-of-fit tests at 199 draws and two
     joint-permutation concordance tests at 199 draws, all on samples of a
     few hundred points. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A statistician has a pipeline that produces posterior draws, and no way
to audit it from the inside. *Three inference engines were pointed at the
same model and each returned a posterior sample. I want two things before
I use any of them. Do these draws actually follow the distribution they
claim to represent, when all I can evaluate is an unnormalised log
density? And do the three engines agree with each other, in a way that
tells me which one departed if they do not?* This vignette answers both,
on synthetic samples where the answer is known: one goodness-of-fit test
against a target density, and one joint test across several samples.

## What

`ksd_test()` is the calibration half. It computes the kernel Stein
discrepancy between a sample and a target distribution, and its defining
property is that it needs only the target's *score* -- the gradient of
its log density -- so the target may be unnormalised and no reference
sample is required. `gaussian_score()` supplies that gradient in closed
form for a multivariate normal target; `numeric_score()` takes it by
finite differences from any log-density function, which is the adapter
that lets an externally supplied log-posterior evaluator be tested
directly. The null is a wild bootstrap rather than a permutation, because
there is only one sample to resample. The default base kernel is the
inverse multi-quadric, which keeps power in higher dimensions where a
Gaussian kernel loses it; `kernel = "rbf"` selects the Gaussian
alternative.

`concordance_test()` is the agreement half. Its input is a list of
samples, one per source, and it asks whether all of them could have come
from one common distribution. It returns a single p-value calibrated by
one joint permutation across all sources, so it is one test with one
verdict rather than a sweep of pairwise comparisons, and alongside it the
full `pairwise` matrix of discrepancies, so a rejection can be read down
to the source responsible.

The two are complementary and neither substitutes for the other.
Calibration compares one sample against a *density*; concordance compares
samples against *each other*. Agreement among independent sources is
corroborating evidence and is hard to fake; calibration against the
target is the absolute check that the agreed answer is also the right
one.

## Do

### A sample that follows its target

A synthetic sample of 200 points from a standard bivariate normal, tested
against the standard bivariate normal target that `ksd_test()` assumes by
default.

```{r ksd-good}
set.seed(1L)
x_ok <- matrix(rnorm(400L), ncol = 2L)
fit_ok <- ksd_test(x_ok, n_boot = 199L, seed = 1L)
fit_ok
```

### The same sample, shifted

The identical points moved 0.7 in both coordinates, so the sample is
mis-specified against the same target by a known amount.

```{r ksd-shift}
x_bad <- x_ok + 0.7
fit_bad <- ksd_test(x_bad, n_boot = 199L, seed = 1L)
fit_bad
```

### A target supplied as an unnormalised log density

When no closed-form score is available, hand `numeric_score()` a log
density and it differentiates numerically. Any additive normalising
constant cancels in the gradient, so an unnormalised log density is
enough.

```{r ksd-numeric}
log_density <- function(z) -0.5 * rowSums(z^2)
fit_numeric <- ksd_test(
  x_ok, score = numeric_score(log_density), n_boot = 199L, seed = 1L
)
fit_numeric
```

### Three sources that agree

Three synthetic samples of 200 points each, all drawn from the same
standard bivariate normal, standing in for three inference engines
pointed at one model.

```{r concordance-ok}
set.seed(2L)
draws <- list(
  engine_a = matrix(rnorm(400L), ncol = 2L),
  engine_b = matrix(rnorm(400L), ncol = 2L),
  engine_c = matrix(rnorm(400L), ncol = 2L)
)
fit_agree <- concordance_test(draws, n_permutations = 199L, seed = 1L)
fit_agree
```

### One source departs

The same three samples with the third moved by 1 in both coordinates.

```{r concordance-bad}
draws_shifted <- draws
draws_shifted$engine_c <- draws_shifted$engine_c + 1
fit_disagree <- concordance_test(
  draws_shifted, n_permutations = 199L, seed = 1L
)
fit_disagree$p_value
```

```{r pairwise-table, echo = FALSE}
knitr::kable(
  fit_disagree$pairwise, digits = 4L,
  caption = paste0(
    "Pairwise discrepancy matrix after engine_c is shifted. The overall ",
    "verdict is a single p-value; this matrix says which pair carries it."
  )
)
```

### The figure: what the pairwise matrix is measuring

```{r fig-clouds, eval = has_ggplot2, fig.cap = "The three sources' samples before and after engine_c is shifted. In the left panel the three clouds are interleaved and the concordance test does not reject; in the right panel engine_c has moved as a body while the other two have not, which is the geometry the pairwise discrepancy matrix summarises to a number."}
to_frame <- function(lst, state) {
  do.call(rbind, lapply(names(lst), function(nm) {
    data.frame(
      dim1 = lst[[nm]][, 1L], dim2 = lst[[nm]][, 2L],
      source = nm, state = state, stringsAsFactors = FALSE
    )
  }))
}
cloud_df <- rbind(
  to_frame(draws, "as drawn"),
  to_frame(draws_shifted, "engine_c shifted")
)
ggplot2::ggplot(
  cloud_df, ggplot2::aes(x = dim1, y = dim2, colour = source)
) +
  ggplot2::geom_point(alpha = 0.6, size = 1.4) +
  ggplot2::facet_wrap(~state) +
  ggplot2::scale_colour_viridis_d(name = "source", end = 0.85) +
  ggplot2::labs(
    x = "coordinate 1", y = "coordinate 2",
    title = "Three sources, before and after one of them departs"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-clouds-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the sample-cloud figure",
    "is skipped.\n")
```

```{r verdict-table, echo = FALSE}
verdicts <- data.frame(
  check = c("draws follow the target", "draws shifted by 0.7",
            "target as an unnormalised log density",
            "three sources agree", "engine_c shifted by 1"),
  test = c("ksd_test()", "ksd_test()", "ksd_test(numeric_score)",
           "concordance_test()", "concordance_test()"),
  statistic = c(fit_ok$statistic, fit_bad$statistic,
                fit_numeric$statistic, fit_agree$statistic,
                fit_disagree$statistic),
  p_value = c(fit_ok$p_value, fit_bad$p_value, fit_numeric$p_value,
              fit_agree$p_value, fit_disagree$p_value),
  reject = c(fit_ok$reject, fit_bad$reject, fit_numeric$reject,
             fit_agree$reject, fit_disagree$reject),
  stringsAsFactors = FALSE
)
knitr::kable(
  verdicts, digits = c(0L, 0L, 5L, 4L, 0L),
  col.names = c("Check", "Test", "Statistic", "p-value",
                paste0("Reject at ", fit_ok$alpha)),
  caption = "The five verdicts in this vignette."
)
```

## Read

The well-specified sample gives a kernel Stein discrepancy of
`r signif(fit_ok$statistic, 3)` with a p-value of
`r round(fit_ok$p_value, 3)`, so the test does not reject: the sample is
consistent with the target it was drawn from, which is the answer the
construction demands. The statistic is a U-statistic and takes small
negative values under the null, which is why a number below zero is not a
sign of trouble; what matters is where it falls in its own bootstrap
null. Moving the same points by 0.7 lifts it to
`r signif(fit_bad$statistic, 3)`, larger than the well-specified value by
`r signif(fit_bad$statistic - fit_ok$statistic, 3)` in absolute terms and
exceeded by `r sum(fit_bad$null_distribution >= fit_bad$statistic)` of
the `r fit_bad$n_permutations` bootstrap draws, so the p-value falls to
`r round(fit_bad$p_value, 4)`. The numerically
differentiated score reproduces the well-specified verdict: statistic
`r signif(fit_numeric$statistic, 3)` against
`r signif(fit_ok$statistic, 3)` from the closed-form score, the two
differing by
`r formatC(abs(fit_numeric$statistic - fit_ok$statistic), format = "e",
   digits = 1L)`.
The finite-difference gradient is an approximation, so exact agreement is
not expected and the two verdicts matching is what the adapter has to
deliver.

The three concordant sources give a p-value of
`r round(fit_agree$p_value, 3)`, so the joint test does not reject and
the three may be treated as mutually consistent. Shifting the third drops
the p-value to `r round(fit_disagree$p_value, 4)`. The pairwise matrix
localises it: the two discrepancies involving `engine_c` are
`r signif(fit_disagree$pairwise["engine_a", "engine_c"], 3)` and
`r signif(fit_disagree$pairwise["engine_b", "engine_c"], 3)`, while the
`engine_a` to `engine_b` discrepancy stays at
`r signif(fit_disagree$pairwise["engine_a", "engine_b"], 3)`. The figure
shows the same fact as geometry: one cloud has moved as a body and the
other two have not.

## Limits

Every sample here is synthetic and drawn independently, which is the
condition both tests assume; posterior draws from a Markov chain are
autocorrelated, and neither the wild bootstrap nor the joint permutation
corrects for that, so a chain should be thinned to something close to
independent before either test is applied. The kernel Stein discrepancy
tests a sample against a *stated* target: it can tell you the draws do
not follow the density you supplied, and it cannot tell you the density
you supplied is the right one. A concordance verdict is symmetric in the
sources, so three engines that share a defect will agree and pass;
agreement is evidence only to the extent that the sources are genuinely
independent, and nothing in this vignette establishes that they are.
Both p-values are bounded below by `1 / (B + 1)` for the bootstrap or
permutation budget `B` used.

## What to read next

*Is the calibrated model right about more than the mean?* takes the same
distributional idea to the posterior-predictive question, comparing a
calibrated ensemble against held-out observations. *Getting started with
kernR* introduces the kernel and two-sample machinery both tests here are
built on. *Scaling HSIC to large ensembles* covers the low-rank
approximations that the Nystrom variants of both tests,
`ksd_test_nystrom()` and `concordance_test_nystrom()`, are built from.

## Reproduce

Seed `1` for the goodness-of-fit samples and `2` for the three sources;
`seed = 1L` passed to every test so the wild-bootstrap and permutation
draws are fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
