---
title: "Choosing a density-ratio backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a density-ratio backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render() with ranger, xgboost and
     proxymix installed; a build missing any of the three renders faster,
     with fewer rows in the comparison. macOS aarch64, R 4.5.2, one core. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

An analyst has run the backdoor-adjusted test and does not like the
diagnostics. *The causal test reweights my observational sample to look
like an intervention, and it told me the effective sample size of those
weights had collapsed. There are four backends behind that step and I
have no idea which one to pick. What actually differs between them, and
how would I tell on my own data which one is behaving?* This vignette
puts all four on one confounded design and reads the weights they
produce, then runs the causal test under each.

## What

`estimate_density_ratio()` is the entry point for every kernR estimator
that reweights an observational sample towards an interventional one,
`bd_hsic_test()` chief among them. It offers four backends behind one
signature, and `fit_density_ratio()` plus `predict_density_ratio()` split
the same work into a fit step and a predict step for callers who need to
apply the ratio to new points.

Three of the four are noise-contrastive classifiers. A classifier is
trained to tell the joint sample `(x, z)` from a product-of-marginals
sample `(x_permuted, z)`, and the density ratio is recovered from its
calibrated class probabilities. `logistic` is the default and fits a
linear logit, which is robust on smooth unimodal densities and cheap.
`ranger` fits a random forest, and `xgboost` gradient-boosted trees; both
are flexible non-linear alternatives, and both need their own package.

The fourth, `proxymix`, is parametric. It fits one Gaussian mixture to
the joint cloud and a second to the product-of-marginals cloud, then
evaluates the analytic ratio of the two mixture densities at each
observation. There is no classifier calibration step, the ratio is closed
form in the fitted parameters, and the fit carries its own convergence
diagnostics; `bd_hsic_test()` gates its verdict on them. The proxymix
package is a separate MIT-licensed package, consumed here as a soft
dependency through `requireNamespace()`, so the binding is one way and
kernR installs without it.

`effective_sample_size()` is the diagnostic that decides whether any of
this worked, and `plot_weights()` draws the weight distribution behind
it.

## Do

### Which backends this build carries

```{r backends}
has_ranger <- requireNamespace("ranger", quietly = TRUE)
has_xgboost <- requireNamespace("xgboost", quietly = TRUE)
has_proxymix <- requireNamespace("proxymix", quietly = TRUE)
c(ranger = has_ranger, xgboost = has_xgboost, proxymix = has_proxymix)
```

```{r backend-note, echo = FALSE, results = "asis"}
missing_backends <- c("ranger", "xgboost", "proxymix")[
  !c(has_ranger, has_xgboost, has_proxymix)
]
if (length(missing_backends) > 0L) {
  cat("Not installed on this build, and therefore absent from every ",
      "comparison below: ", paste(missing_backends, collapse = ", "),
      ".\n\n", sep = "")
}
```

### A confounded design

A synthetic design of 200 observations: a two-column Gaussian confounder,
a treatment that is a linear-Gaussian function of its first column, and
an outcome carrying a genuine causal effect of 0.7 per unit of treatment
plus a confounded path through the second column.

```{r data}
set.seed(2026L)
n_obs <- 200L
z_conf <- matrix(rnorm(n_obs * 2L), n_obs, 2L)
x_treat <- z_conf[, 1L] + rnorm(n_obs, sd = 0.5)
y_out <- 0.7 * x_treat + z_conf[, 2L] + rnorm(n_obs, sd = 0.4)
```

### The four ratios

```{r fits}
dr_fits <- list(
  logistic = estimate_density_ratio(x_treat, z_conf,
                                    method = "logistic", seed = 1L)
)
if (has_ranger) {
  dr_fits$ranger <- estimate_density_ratio(x_treat, z_conf,
                                           method = "ranger", seed = 1L)
}
if (has_xgboost) {
  dr_fits$xgboost <- estimate_density_ratio(x_treat, z_conf,
                                            method = "xgboost", seed = 1L)
}
if (has_proxymix) {
  dr_fits$proxymix <- estimate_density_ratio(
    x_treat, z_conf, method = "proxymix", proxymix_components = 2L,
    seed = 1L
  )
}
names(dr_fits)
```

```{r fits-table, echo = FALSE}
weight_tab <- do.call(rbind, lapply(names(dr_fits), function(nm) {
  w1 <- dr_fits[[nm]]$weights
  data.frame(
    backend = nm, ess = dr_fits[[nm]]$ess,
    ess_fraction = dr_fits[[nm]]$ess / length(w1),
    min_weight = min(w1), max_weight = max(w1),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  weight_tab, digits = c(0L, 1L, 3L, 3L, 3L), row.names = FALSE,
  col.names = c("Backend", "ESS", "ESS / n", "Smallest weight",
                "Largest weight"),
  caption = paste0(
    "Density-ratio weights on the same ", n_obs,
    "-observation confounded design, one row per available backend."
  )
)
```

### The figure: the weights behind the diagnostic

```{r fig-weights, eval = has_ggplot2, fig.cap = "Distribution of the density-ratio weights produced by each available backend on the same design, on a logarithmic scale, with the unit weight that a perfectly balanced sample would carry marked. The effective sample size in the table above is a summary of exactly this spread: the wider the distribution, the fewer observations are really contributing."}
weights_long <- do.call(rbind, lapply(names(dr_fits), function(nm) {
  data.frame(backend = nm, weight = dr_fits[[nm]]$weights,
             stringsAsFactors = FALSE)
}))
ggplot2::ggplot(
  weights_long, ggplot2::aes(x = weight, y = backend, colour = backend)
) +
  ggplot2::geom_vline(xintercept = 1, colour = "grey50",
                      linetype = "dashed") +
  ggplot2::geom_jitter(height = 0.18, alpha = 0.5, size = 1.3) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.85,
                                  guide = "none") +
  ggplot2::labs(
    x = "importance weight (log scale, unit weight dashed)",
    y = NULL, title = "How wide the weights are, backend by backend"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-weights-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the weight-distribution",
    "figure is skipped.\n")
```

### The causal test under each backend

Any backend whose weights collapse makes `bd_hsic_test()` raise its
effective-sample-size warning. The runner below captures those messages
instead of letting them print as conditions, so that they can be quoted
where the prose reads them and the vignette still renders without
warnings.

```{r bdhsic-runs}
gate_messages <- character(0)
run_bd <- function(method) {
  withCallingHandlers(
    bd_hsic_test(
      x_treat, y_out, z_conf, density_ratio = method,
      n_permutations = 199L, seed = 1L
    ),
    warning = function(w) {
      gate_messages[[method]] <<- conditionMessage(w)
      invokeRestart("muffleWarning")
    }
  )
}
bd_fits <- lapply(stats::setNames(nm = names(dr_fits)), run_bd)
names(gate_messages)
```

```{r bdhsic-table, echo = FALSE}
bd_tab <- do.call(rbind, lapply(names(bd_fits), function(nm) {
  f1 <- bd_fits[[nm]]
  data.frame(
    backend = nm, statistic = f1$statistic, p_value = f1$p_value,
    ess = f1$ess, abstained = is_kernR_abstention(f1),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  bd_tab, digits = c(0L, 4L, 4L, 1L, 0L), row.names = FALSE,
  col.names = c("Backend", "Statistic", "p-value", "ESS on the test split",
                "Declined"),
  caption = paste0(
    "bd-HSIC on the same design under each available backend, ",
    bd_fits[[1L]]$n_permutations, " cluster permutations."
  )
)
```

## Read

The backend choice changes the answer on this design, which is the most
useful thing this comparison has to report. The weights come out very
differently: the effective sample size runs from
`r round(min(weight_tab$ess), 1)` to `r round(max(weight_tab$ess), 1)`
out of `r n_obs` observations,
`r round(100 * min(weight_tab$ess_fraction))` to
`r round(100 * max(weight_tab$ess_fraction))` per cent, and the largest
single weight from
`r round(min(weight_tab$max_weight), 2)` to
`r round(max(weight_tab$max_weight), 2)`. The figure shows the same fact
as a picture: `r weight_tab$backend[which.max(weight_tab$ess)]` produces
a tight cloud around the unit weight, and
`r weight_tab$backend[which.min(weight_tab$ess)]` a cloud spanning
orders of magnitude.

Those weights carry through to the verdict. The
`r nrow(bd_tab)` backends return p-values of
`r paste(paste0(bd_tab$backend, " ", round(bd_tab$p_value, 3)),
   collapse = ", ")`, so on a design with a causal effect of 0.7 per unit
built into it,
`r sum(bd_tab$p_value <= 0.05)` of `r nrow(bd_tab)` detect it. The
mechanism is not only that the statistic changes -- the four statistics
all sit between `r signif(min(bd_tab$statistic), 3)` and
`r signif(max(bd_tab$statistic), 3)`, within a factor of
`r round(max(bd_tab$statistic) / min(bd_tab$statistic), 1)` of each
other -- but that the *null* changes with it. Without an explicit
`cluster_id`, `bd_hsic_test()` builds its permutation clusters from the
density-ratio weights themselves, so a
backend that produces different weights is scored against a different
null distribution. Two backends can therefore agree on what they measured
and disagree on what it means.

The `proxymix` row is the one to read closely. Its effective sample size
on the test split falls to `r round(bd_tab$ess[bd_tab$backend ==
"proxymix"], 1)`, the gate fires, and the result is stamped as an
abstention:
`r bd_tab$abstained[bd_tab$backend == "proxymix"]` in the `Declined`
column. A two-component mixture is the wrong model for a joint density
with one mode, the fitted ratio is extreme where the two mixtures
disagree in the tails, and the package refuses to stand behind the
p-value that follows rather than reporting
`r round(bd_tab$p_value[bd_tab$backend == "proxymix"], 3)` as a verdict.
The gate says so in its own words:

> `r gate_messages[["proxymix"]]`

That is the behaviour to want from a backend that has been handed a
density it cannot represent. *Does the treatment cause the outcome, or do
they share a cause?* shows the same gate firing for the opposite reason,
on a design with too many covariates rather than the wrong mixture.

## Limits

This is one synthetic design with a smooth, unimodal, two-dimensional
joint density and 200 observations. It cannot rank the four backends in
general, and the disagreement above should be read as a demonstration
that backend choice is consequential, not as a ranking to carry to
another data shape: a design with a multimodal joint density would be
expected to reverse part of it. The classifier backends have a tuning
surface this vignette does not touch -- forest size, boosting depth and
rounds, the noise-sample count -- and only the defaults are compared. The
proxymix backend is run with two mixture components against a design with
one mode per cloud, which is deliberately the wrong specification and is
why its weights are extreme; a fair comparison on a multimodal design
would need its own vignette. Nothing here establishes which p-value is
correct: the design has a real causal effect, so a backend that fails to
detect it has lost power, but this single run cannot separate lost power
from an unlucky permutation draw. Above all, no backend can establish
that the adjustment set is complete, and an incomplete one produces a
confident wrong answer under every backend at once.

## When to reach for which

The classifier backends are the default for good reasons: they tolerate a
misspecified density, scale to a high-dimensional adjustment set, and
their calibration is well understood. `logistic` is the right first
choice, with `ranger` or `xgboost` when the conditional density is
plainly non-linear in the confounders.

Reach for `proxymix` when the joint density is plausibly **multimodal**,
as it is under a multi-regime climate, a paddock by variety design with
distinct production zones, or a cohort with separable subpopulations,
where a two-component mixture represents the structure cleanly and a
classifier smears across the modes; when a **parametric** ratio is wanted
whose components can be inspected or handed to a downstream Bayesian
step through `proxymix::gmm_target_from_posterior()`; or when classifier
calibration is unreliable on the sample at hand, because it is small,
because the joint against marginal split is sharply imbalanced, or
because the features are pathologically scaled.

## What to read next

*Does the treatment cause the outcome, or do they share a cause?* is the
test this backend choice feeds, and shows the effective-sample-size floor
declining a verdict outright. *A treatment that changes the spread and
not the mean* uses the related propensity-model choice for a binary
treatment. *From a coarse grid to a paddock* uses proxymix from the other
direction, as the fitter for the Gaussian-mixture prior that
`aggregate_downscale()` inverts.

## References

- Hoek, J. van der, & Elliott, R. J. (2024). Mixtures of multivariate
  Gaussians. *Stochastic Analysis and Applications*.
  DOI: 10.1080/07362994.2024.2372605.
- Hu, R., Sejdinovic, D., & Evans, R. J. (2024). A kernel test for causal
  association via noise contrastive backdoor adjustment. *Journal of
  Machine Learning Research*, 25(160), 1-56.

## Reproduce

Seed `2026` for the confounded design; `seed = 1L` passed to every
density-ratio fit and every test, so the classifier fits, the mixture
fits and the permutation draws are all fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
