---
title: "HSIC-Based Distributional Sensitivity"
author: "Max Moldovan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HSIC-Based Distributional Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
set.seed(1)
```

## What HSIC-Sensitivity catches that Sobol misses

Variance-based Sobol indices are the workhorse of global sensitivity
analysis: they decompose the output variance into contributions from
each input. They have one structural blind spot -- *distributional*
effects that leave variance unchanged. A parameter that flips the
output skewness, fattens its tails, or shifts a quantile without moving
the variance is invisible to Sobol.

The HSIC-Sensitivity Index (Da Veiga, 2015) replaces the variance with
a kernel-based dependence measure (the Hilbert-Schmidt Independence
Criterion). The normalised first-order index is

\[
S_j^{\mathrm{HSIC}}
  = \frac{\mathrm{HSIC}(X_j, Y)}
         {\sqrt{\mathrm{HSIC}(X_j, X_j) \cdot \mathrm{HSIC}(Y, Y)}}
  \in [0, 1].
\]

It is bounded like a Sobol index, comparable in magnitude across
parameters, and detects mean-preserving distributional effects.

## A sensitivity scan on a stubbed APSIM-like simulator

```{r stub}
library(kernR)

stub_apsim <- function(theta) {
  # theta: n x 4 matrix
  # `mean_effect` shifts the mean (Sobol catches this)
  # `var_effect`  shifts the variance only -- mean-preserving
  # `tail_effect` injects a heavy-tailed component (visible to HSIC)
  # `inert`       has no effect
  n <- nrow(theta)
  yield <- 1.5 * theta[, "mean_effect"] +
           stats::rnorm(n, sd = exp(theta[, "var_effect"])) +
           theta[, "tail_effect"] *
             stats::rt(n, df = 3) * 0.4 +
           stats::rnorm(n, sd = 0.05)
  matrix(yield, ncol = 1L, dimnames = list(NULL, "yield"))
}

bounds <- rbind(
  mean_effect = c(0.0, 1.0),
  var_effect  = c(-0.5, 0.5),
  tail_effect = c(0.0, 1.0),
  inert       = c(0.0, 1.0)
)
design <- lhs_design(n = 200L, bounds = bounds, seed = 11L)
y      <- stub_apsim(design)
```

## The scan

```{r scan}
fit <- hsic_sensitivity(
  theta          = design,
  y              = y,
  n_permutations = 199L,
  seed           = 11L
)
fit
```

The print method ranks parameters by their first-order HSIC-Sensitivity
Index. Expectations on this design:

- `mean_effect` -- large index, small p-value (Sobol would also catch
  this).
- `var_effect` -- non-negligible index despite preserving the
  conditional mean (Sobol misses this).
- `tail_effect` -- detectable distributional contribution.
- `inert` -- index near zero, p-value not significant.

## Visual

```{r plot}
plot(fit)
```

Bars colour-coded by significance at the default `alpha = 0.05` after
Benjamini-Hochberg adjustment.

## Side-by-side with `sensitivity::sobol`

If you have the `sensitivity` package installed, you can compare HSIC
indices to Sobol's first-order indices on the same design:

    library(sensitivity)
    sobol_fit <- sensitivity::sobolEff(
      model = function(X) stub_apsim(as.matrix(X))[, 1L],
      X1    = as.data.frame(design[1:100, ]),
      X2    = as.data.frame(design[101:200, ]),
      nboot = 200
    )
    # Compare sobol_fit$S to fit$index

The headline observation: Sobol will rank `mean_effect` first and
likely flag `tail_effect` and `var_effect` as small/negligible. HSIC
captures the latter two as genuine signal -- the verdict you want when
the decision-relevant question is *whether the parameter changes the
distribution*, not just the variance.

## Tabular form

```{r as-df}
df <- as.data.frame(fit)
df[order(-df$index), ]
```

## Total-order indices

The **first-order** index `S_j` measures the direct contribution of
`X_j` to `Y`. It misses contributions of `X_j` *through interactions*
with other parameters. The **total-order** index, in Da Veiga's
complement formulation,

\[
T_j^{\mathrm{HSIC}}
  = 1 - \frac{\mathrm{HSIC}(X_{\sim j}, Y)}
              {\sqrt{\mathrm{HSIC}(X_{\sim j}, X_{\sim j}) \cdot
                     \mathrm{HSIC}(Y, Y)}}
\]

captures the total -- direct plus all-order interactions -- contribution
of `X_j`. By construction:

- Purely additive model `Y = sum_j f_j(X_j)`: `T_j = S_j`.
- Pure interaction `Y = X_1 X_2`: `S_j` may be near zero (marginal
  signal averages out) but `T_j` is strong.

The difference `T_j - S_j` quantifies the interactive component.

### A pure-interaction example

```{r interaction-stub}
set.seed(42L)
n <- 400L
theta_int <- matrix(stats::runif(n * 2L, min = -1, max = 1), n, 2L,
                    dimnames = list(NULL, c("x1", "x2")))
y_int <- theta_int[, "x1"] * theta_int[, "x2"] +
         stats::rnorm(n, sd = 0.05)

fit_int <- hsic_sensitivity(
  theta          = theta_int,
  y              = y_int,
  total_order    = TRUE,
  p_value        = FALSE,    # total-order p-values not computed (see Notes)
  n_permutations = 199L,
  seed           = 42L
)
fit_int
```

Both `x1` and `x2` should have small first-order indices (the marginal
expectation of `X1 X2` is zero) but large total-order indices. The
`interaction = T - S` column makes the gap explicit.

### Side-by-side plot

```{r interaction-plot, fig.height = 4.5}
plot(fit_int, which = "both")
```

### A near-additive example for contrast

```{r additive-stub}
theta_add <- matrix(stats::runif(n * 2L), n, 2L,
                    dimnames = list(NULL, c("x1", "x2")))
y_add <- 2 * theta_add[, "x1"] + theta_add[, "x2"] +
         stats::rnorm(n, sd = 0.05)

fit_add <- hsic_sensitivity(theta_add, y_add, total_order = TRUE,
                            p_value = FALSE, seed = 1L)
fit_add
```

On this additive model `T_j - S_j` should be close to zero for both
parameters.

### Pair-bootstrap CI for total-order indices

`total_order_ci = TRUE` (since 0.0.0.9013) activates a pair-bootstrap
percentile CI on each total-order index `T_j`:

```{r pairboot-ci}
fit_pf <- hsic_sensitivity(
  theta_add, y_add,
  total_order         = TRUE,
  total_order_ci      = TRUE,
  n_permutations      = 199L,
  n_bootstrap         = 100L,
  ci_level            = 0.95,
  seed                = 7L
)
fit_pf
as.data.frame(fit_pf)[, c("parameter", "output",
                          "index_total_order",
                          "ci_total_order_lower",
                          "ci_total_order_upper")]
```

The CI is uncertainty quantification on the *index magnitude*, not a
hypothesis test. An earlier (0.0.0.9012) `total_order_p_value` field
claimed to test `H_0: T_j = 0` — that was found in critical review
on 2026-05-16 to be mis-calibrated under pure-noise `Y` (all
parameters were assigned `p ≈ 1/(1 + B)` because the bootstrap
samples the empirical joint, not a null-of-no-effect) and was removed
in 0.0.0.9013. A properly null-calibrated total-order significance
test remains future work. Cost scales as `n_bootstrap × p × q` kernel
re-evaluations on resampled designs; default `n_bootstrap = 200`
is a workable starting point for `p, q ≤ 10`, `n ≤ 1000`.

## Notes on total-order indices

- **Cost.** Total-order construction is `O(p n^2)` for kernel matrices
  on each `X_{~j}` plus `O(p q n^2)` for HSIC, similar in scale to
  first-order. For ag-scale designs (`n` in the hundreds-low thousands)
  both are practical. Pick-Freeze bootstrap multiplies this by
  `n_bootstrap` (default 200) — budget accordingly.
- **Nystrom acceleration for total-order** is intentionally not
  exposed in this version. The naive way to fold `nystrom_factor()`
  into `T_j` computation (materialising `F F^T` into a full `n x n`
  kernel) is slower than the exact path at typical kernR scales,
  because materialisation costs `O(n^2 m)`. A proper
  factor-only HSIC primitive (Nystrom-on-Nystrom) is the unblocking
  next step and is on the future-work list.

## Notes on practice

- **Design size.** As with B1 identifiability, `n >= 10p` is a
  reasonable starting point.
- **Permutations.** When the goal is ranking only, set
  `p_value = FALSE` to skip the permutation null entirely -- the
  per-column kernel cache is reused, so the index calculation is
  cheap. Use `p_value = TRUE` (default) when you also want a
  significance verdict on first-order indices.

## Notes on practice

- **Design size.** As with B1 identifiability, `n >= 10p` is a
  reasonable starting point.
- **Permutations.** When the goal is ranking only, set
  `p_value = FALSE` to skip the permutation null entirely -- the
  per-column kernel cache is reused, so the index calculation is
  cheap. Use `p_value = TRUE` (default) when you also want a
  significance verdict.
- **Comparing to Sobol.** HSIC-SI is not numerically equal to a Sobol
  first-order index even on linear additive systems. They are
  *bounded comparable* (both in `[0, 1]`) and typically agree on
  ranking when only mean effects are present. Disagreement is the
  point: when HSIC ranks a Sobol-flat parameter highly, that parameter
  is shifting the output distribution in a non-variance way.

## References

- Da Veiga, S. (2015). Global sensitivity analysis with dependence
  measures. *Journal of Statistical Computation and Simulation*,
  85(7), 1283-1305.
- Gretton, A., Bousquet, O., Smola, A., & Schölkopf, B. (2005).
  Measuring statistical dependence with Hilbert-Schmidt norms. *ALT*,
  63-77.
