---
title: "A treatment that changes the spread and not the mean"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A treatment that changes the spread and not the mean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. Four doubly robust tests at 299 permutations on 300 to 400
     observations, each with a propensity fit and an outcome fit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A risk analyst is looking at a management practice whose average did
nothing. *The mean yield under the new practice is the same as under the
old one, and every mean-based method I have tells me there is no effect.
But the good years are better and the bad years are worse, and it is the
bad years that decide whether the business survives. Is there a test that
calls that an effect?* This vignette answers with the two doubly robust
kernel tests for distributional treatment effects, on synthetic designs
where the treatment shifts the mean, shifts only the variance, and acts
on a subgroup with imperfect overlap.

## What

`dr_date_test()` tests the Distributional Average Treatment Effect
hypothesis $P(Y(1)) = P(Y(0))$ over the whole population: not whether the
two potential-outcome distributions have the same mean, but whether they
are the same distribution. It compares the two kernel mean embeddings, so
a change in variance, skewness, modality or tail weight counts as an
effect exactly as a shift in location does.

`dr_dett_test()` tests the Distributional Effect on the Treated,
$P(Y(1) \mid T = 1) = P(Y(0) \mid T = 1)$, restricting the question to the
units that were actually treated. That restriction buys a weaker
assumption: it needs overlap only on one side, so covariate regions with
no control units do not invalidate it.

Both are doubly robust. Each combines a propensity model
($P(T = 1 \mid X)$, fitted by `estimate_propensity()`) with an outcome
model, and the estimator stays consistent if either of the two is
correctly specified. `assess_overlap()` is the companion diagnostic for
the positivity assumption both rely on. `kernel_causal_test()` reaches
the same tests through a formula, with `method = "dr-date"` or
`"dr-dett"`. Every call returns a `kernel_test_result` carrying the
statistic, the permutation null, the p-value and the effective sample
size of the weights.

## Do

### A mean shift, which any method would find

A synthetic design of 300 observations: two covariates drive treatment
assignment, and treatment adds a constant 1.0 to the outcome.

```{r mean-shift}
set.seed(42L)
n_obs <- 300L
x_cov <- matrix(rnorm(n_obs * 2L), n_obs, 2L)
t_arm <- rbinom(n_obs, 1L, plogis(0.3 * x_cov[, 1L] - 0.2 * x_cov[, 2L]))
y_shift <- t_arm * 1.0 + 0.5 * x_cov[, 1L] + rnorm(n_obs, sd = 0.5)

fit_shift <- dr_date_test(y_shift, t_arm, x_cov,
  n_permutations = 299L, seed = 1L
)
fit_shift
```

### A variance effect and nothing else

The same machinery on a synthetic design in which treatment multiplies
the outcome's standard deviation by 2.5 and leaves the mean exactly where
it was: each arm's noise is centred within the arm, so the treatment
contributes no mean shift by construction rather than only in
expectation. This is the design a mean-based method has no power against.

```{r variance-effect}
set.seed(42L)
n_var <- 400L
x_var <- matrix(rnorm(n_var * 2L), n_var, 2L)
t_var <- rbinom(n_var, 1L, plogis(0.3 * x_var[, 1L]))
noise_treated <- rnorm(n_var, sd = 2.5)
noise_control <- rnorm(n_var, sd = 1)
noise_treated <- noise_treated - mean(noise_treated[t_var == 1L])
noise_control <- noise_control - mean(noise_control[t_var == 0L])
y_var <- (1L - t_var) * noise_control + t_var * noise_treated +
  0.5 * x_var[, 1L]
```

The mean-based reading of this design, for comparison: an ordinary least
squares fit of the outcome on treatment and the two covariates, which is
what a difference-in-means analysis amounts to once the covariates are
adjusted for.

```{r variance-lm}
lm_variance <- lm(y_var ~ t_var + x_var)
lm_effect <- coef(lm_variance)[["t_var"]]
lm_p <- summary(lm_variance)$coefficients["t_var", "Pr(>|t|)"]
sd_treated <- sd(y_var[t_var == 1L])
sd_control <- sd(y_var[t_var == 0L])
round(c(lm_effect = lm_effect, lm_p = lm_p,
        sd_treated = sd_treated, sd_control = sd_control), 3L)
```

```{r variance-effect-test}
fit_variance <- dr_date_test(y_var, t_var, x_var,
  n_permutations = 299L, outcome_model = "zero", seed = 1L
)
fit_variance
```

### The effect on the treated, with imperfect overlap

A synthetic design in which treatment assignment depends more strongly on
the first covariate, so some covariate regions carry almost no controls.
`dr_dett_test()` needs overlap only on the treated side.

```{r dett}
set.seed(42L)
n_dett <- 300L
x_dett <- matrix(rnorm(n_dett * 2L), n_dett, 2L)
t_dett <- rbinom(n_dett, 1L, plogis(0.5 * x_dett[, 1L]))
y_dett <- t_dett * rnorm(n_dett, mean = 0.5, sd = 1.5) +
  (1L - t_dett) * rnorm(n_dett) + x_dett[, 1L]

overlap <- assess_overlap(
  estimate_propensity(t_dett, x_dett)$scores, t_dett
)
overlap

fit_dett <- dr_dett_test(y_dett, t_dett, x_dett,
  n_permutations = 299L, seed = 1L
)
fit_dett
```

### The same question through a formula

```{r formula}
dat_dett <- data.frame(
  y = y_dett, treatment = t_dett,
  x1 = x_dett[, 1L], x2 = x_dett[, 2L]
)
fit_formula <- kernel_causal_test(
  y ~ treatment | x1 + x2,
  data = dat_dett, method = "dr-date",
  n_permutations = 299L, seed = 1L
)
fit_formula
```

### The figure: what the mean cannot show

```{r fig-spread, eval = has_ggplot2, fig.cap = "Covariate-adjusted outcome by arm on the variance-only design (residuals from a fit of the outcome on the two covariates). The two arm means, dashed vertical lines, sit almost on top of each other; the treated arm's distribution is more than twice as wide. A mean-based test compares the two dashed lines; DR-DATE compares the two curves."}
resid_y <- residuals(lm(y_var ~ x_var))
spread_df <- data.frame(
  outcome = resid_y,
  arm = ifelse(t_var == 1L, "treated", "control")
)
mean_df <- data.frame(
  arm = c("control", "treated"),
  outcome = c(mean(resid_y[t_var == 0L]), mean(resid_y[t_var == 1L]))
)
ggplot2::ggplot(spread_df, ggplot2::aes(x = outcome, fill = arm)) +
  ggplot2::geom_density(alpha = 0.55, colour = NA) +
  ggplot2::geom_vline(
    data = mean_df,
    ggplot2::aes(xintercept = outcome, colour = arm),
    linetype = "dashed", linewidth = 0.8
  ) +
  ggplot2::scale_fill_viridis_d(name = "arm", end = 0.85) +
  ggplot2::scale_colour_viridis_d(name = "arm", end = 0.85) +
  ggplot2::labs(
    x = "covariate-adjusted outcome (yield units)", y = "density",
    title = "Equal means, unequal spread"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-spread-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the by-arm density",
    "figure is skipped.\n")
```

```{r verdict-table, echo = FALSE}
verdicts <- data.frame(
  design = c("mean shift of 1.0", "variance only, mean preserved",
             "effect on the treated", "effect on the treated, formula"),
  test = c("dr_date_test()", "dr_date_test()", "dr_dett_test()",
           "kernel_causal_test(dr-date)"),
  statistic = c(fit_shift$statistic, fit_variance$statistic,
                fit_dett$statistic, fit_formula$statistic),
  p_value = c(fit_shift$p_value, fit_variance$p_value,
              fit_dett$p_value, fit_formula$p_value),
  ess = c(fit_shift$ess, fit_variance$ess, fit_dett$ess,
          fit_formula$ess),
  n_used = c(fit_shift$n, fit_variance$n, fit_dett$n, fit_formula$n),
  stringsAsFactors = FALSE
)
verdicts$ess_pct <- 100 * verdicts$ess / verdicts$n_used
any_ess_flag <- any(
  c(fit_shift$ess_warning, fit_variance$ess_warning,
    fit_dett$ess_warning, fit_formula$ess_warning)
)
knitr::kable(
  verdicts[, c("design", "test", "statistic", "p_value", "ess",
               "n_used", "ess_pct")],
  digits = c(0L, 0L, 4L, 4L, 1L, 0L, 1L),
  col.names = c("Design", "Test", "Statistic", "p-value", "ESS", "n",
                "ESS (% of n)"),
  caption = paste0(
    "Doubly robust distributional verdicts on the three synthetic ",
    "designs, each against a ", fit_shift$n_permutations,
    "-draw permutation null."
  )
)
```

## Read

The mean-shift design is the sanity check: DR-DATE returns a statistic of
`r signif(fit_shift$statistic, 3)` with a p-value of
`r round(fit_shift$p_value, 4)`, so the test finds the effect a
difference-in-means would also have found.

The variance-only design is the point of the vignette. The least squares
treatment coefficient is `r round(lm_effect, 3)` yield units with a
p-value of `r round(lm_p, 3)`, so the mean-based reading of this design
is that nothing happened. Meanwhile the treated arm's standard deviation
is `r round(sd_treated, 2)` against `r round(sd_control, 2)` in the
control arm, more than double. DR-DATE returns
`r signif(fit_variance$statistic, 3)` with a p-value of
`r round(fit_variance$p_value, 4)`, the smallest value
`r fit_variance$n_permutations` permutations can express. The figure
shows why: the two dashed mean lines coincide and the two curves do not.
An analyst deciding on downside risk cares about the curves.

On the treated-effect design the propensity scores overlap over
`r round(100 * overlap$overlap_fraction, 1)` per cent of their combined
range, so positivity is imperfect but not broken; the overlap diagnostic
`r if (overlap$overlap_warning) "flags this as poor overlap" else
"does not flag it as poor overlap"`. DR-DETT returns
`r signif(fit_dett$statistic, 3)` with a p-value of
`r round(fit_dett$p_value, 4)`, which
`r if (fit_dett$p_value < 0.05) "clears" else "does not clear"` the
conventional five per cent threshold. The formula front door run on the same
data through DR-DATE gives `r signif(fit_formula$statistic, 3)` and
`r round(fit_formula$p_value, 4)`: the two estimands differ, so the two
numbers are not expected to agree exactly, and the table reports both
rather than choosing between them.

Across the four rows the effective sample size of the weights runs from
`r round(min(verdicts$ess_pct))` to `r round(max(verdicts$ess_pct))` per
cent of the observations each test used. That is well clear of the ten
per cent floor below which both tests set `ess_warning`, and on this run
`r if (any_ess_flag) "at least one row trips that flag" else
"no row trips that flag"`, so none of these verdicts rests on a handful
of dominant weights.

## Limits

All four designs are synthetic, with the propensity model correctly
specified by construction; double robustness is a guarantee about one of
two models being right, not about both being wrong. These tests report
that the two distributions differ, not how or by how much: a rejection
here is a licence to look at the curves, not an effect size, and nothing
in this vignette quantifies the change in the lower tail that motivated
the question. The variance-only design uses `outcome_model = "zero"`,
which leans the estimator entirely on the propensity side, and the
default outcome model would be the better choice whenever the outcome is
predictable from the covariates. Both tests assume positivity; DR-DETT
relaxes it to one side but does not remove it, and the overlap
diagnostic above is a summary of the propensity score, not a proof that
the assumption holds. Permutation p-values cannot fall below
`1 / (n_permutations + 1)`.

## What to read next

*Does the treatment cause the outcome, or do they share a cause?* handles
the continuous-treatment version of the same causal question and shows
the effective-sample-size gate declining a verdict outright. *When plots
sit inside farms* runs these same tests on clustered designs, where the
permutation null used here is not valid. *Did the management change shift
the simulated distribution?* applies DR-DATE to a pair of simulator
ensembles rather than to a field record.

## References

- Fawkes, J., Hu, R., Evans, R. J., & Sejdinovic, D. (2024). Doubly
  robust kernel statistics for testing distributional treatment effects.
  *Transactions on Machine Learning Research*.

## Reproduce

Seed `42` for all three synthetic designs; `seed = 1L` passed to every
test so the propensity fit, the outcome fit and the permutation draws are
fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
