---
title: "Did the management change shift the simulated distribution?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Did the management change shift the simulated distribution?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. One in-process ensemble-smoother calibration of 60 realisations
     over a linear-Gaussian forward model, then three scenario tests at 99
     to 299 permutations. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
library(PESTO)
```

## Why

A systems modeller has one calibrated crop model and two management
rules. *I calibrated the model once against the site record, then ran the
posterior forward twice: once under the current stubble management and
once under retention. The two mean yields differ by less than the spread
of either, so the difference in means tells me nothing. What I want to
know is whether the whole distribution of simulated outcomes moved, and I
want the comparison to refuse to run if the two ensembles are not
actually comparable.* This vignette answers that on a pair of ensemble
manifests, with a synthetic forward model standing in for the simulator.

## What

The canonical structure is one calibration and two forward runs. The
ensemble smoother produces a single parameter posterior; that posterior
is pushed forward under each management scenario; and the question
becomes a balanced two-sample comparison on the outputs.

`PESTO::pesto_ies_callback()` runs the ensemble smoother in process
against any R forward model, and `PESTO::apsim_callback()` is the adapter
that makes a real APSIM run play that part.
`PESTO::pesto_ensemble_manifest()` packages a run as the orchestra's
manifest object, the C2 contract in the federation's contract index: the
parameter ensemble, the output ensemble, the observation targets, the
weights, and the provenance fields that make two runs comparable or
prove that they are not.

`dr_date_scenario()` takes two of those manifests and runs the DR-DATE
statistic of Fawkes, Hu, Evans and Sejdinovic (2024) over their outputs.
When the two manifests share a parameter posterior, the doubly robust
correction collapses to a balanced Maximum Mean Discrepancy on the
outputs. When they do not, because the two scenarios were calibrated
against different data, the doubly robust adjustment absorbs the
covariate shift in parameter space, provided positivity holds. `output`
restricts the test to named observation columns.

## Do

### One calibration

A synthetic linear-Gaussian forward model stands in for APSIM so that
this vignette runs without a simulator installed. Replace `forward` with
`PESTO::apsim_callback()` to drive a real ensemble.

```{r calibration}
set.seed(20260516L)
n_par <- 2L
n_obs <- 4L
n_real <- 60L
sigma <- 0.05

design_matrix <- matrix(rnorm(n_obs * n_par), n_obs, n_par)
y_obs <- as.numeric(design_matrix %*% c(1.0, -0.5)) +
  rnorm(n_obs, sd = sigma)
names(y_obs) <- paste0("o", seq_len(n_obs))

prior <- matrix(
  rnorm(n_real * n_par), n_real, n_par,
  dimnames = list(NULL, c("p1", "p2"))
)
forward <- function(theta) theta %*% t(design_matrix)

fit_ies <- pesto_ies_callback(
  forward_model = forward, prior_ensemble = prior, obs = y_obs,
  obs_sd = sigma, noptmax = 3L, verbose = FALSE
)
```

### Two forward runs from that one posterior

The intervention here is a deterministic shift of 0.6 on every output.
In an agricultural application it would be a management rule that
changes the simulated yield trajectory.

```{r scenarios}
par_post <- as.matrix(
  fit_ies$par_ensemble[, c("p1", "p2"), with = FALSE]
)
out_baseline <- par_post %*% t(design_matrix)
out_intervention <- out_baseline + 0.6
colnames(out_baseline) <- names(y_obs)
colnames(out_intervention) <- names(y_obs)
```

Each scenario is wrapped as a manifest. The two share the same parameter
posterior, which is what makes this the canonical construction.

```{r manifests}
real_names <- fit_ies$par_ensemble$real_name

make_manifest <- function(run_id, outputs) {
  pesto_ensemble_manifest(
    run_id = run_id,
    params = data.frame(real_name = real_names, par_post,
                        check.names = FALSE),
    outputs = data.frame(real_name = real_names, outputs,
                         check.names = FALSE),
    weights = setNames(rep(1 / sigma, n_obs), names(y_obs)),
    obs_target = setNames(y_obs, names(y_obs)),
    data_hash = paste0("sha256:vignette_", run_id),
    pesto_version = as.character(packageVersion("PESTO")),
    timestamp = Sys.time(),
    method = "ies_callback",
    noptmax = 3L,
    lambda_schedule = 1
  )
}

m_baseline <- make_manifest("wagga_baseline_2026", out_baseline)
m_intervention <- make_manifest("wagga_stubble_2026", out_intervention)
m_baseline
```

In production the two manifests come from `PESTO::as_manifest()` on two
separate calibrations, but only when the scenarios genuinely produced
different calibration data. For the question *did the intervention shift
the forward outputs?* the shared-posterior construction above is the
right one.

### The scenario test

```{r scenario-test}
res_shift <- dr_date_scenario(
  baseline = m_baseline, intervention = m_intervention,
  n_permutations = 299L, seed = 1L
)
res_shift
```

### The null case

The same comparison against a replicate of the baseline, with no
intervention effect at all.

```{r null-case}
m_replicate <- make_manifest("wagga_baseline_replicate", out_baseline)
res_null <- dr_date_scenario(
  baseline = m_baseline, intervention = m_replicate,
  n_permutations = 299L, seed = 2L
)
res_null$p_value
```

### Restricting the test to named outputs

For an ensemble with many output columns, `output` focuses the comparison
on the ones the decision turns on.

```{r output-subset}
res_subset <- dr_date_scenario(
  m_baseline, m_intervention, output = c("o1", "o3"),
  n_permutations = 99L, seed = 3L
)
res_subset$outputs_tested
res_subset$p_value
```

```{r scenario-table, echo = FALSE}
scenario_tab <- data.frame(
  comparison = c("baseline against intervention",
                 "baseline against a replicate of itself",
                 "baseline against intervention, two outputs"),
  outputs = c(paste(res_shift$outputs_tested, collapse = ", "),
              paste(res_null$outputs_tested, collapse = ", "),
              paste(res_subset$outputs_tested, collapse = ", ")),
  statistic = c(res_shift$statistic, res_null$statistic,
                res_subset$statistic),
  p_value = c(res_shift$p_value, res_null$p_value, res_subset$p_value),
  ess = c(res_shift$ess, res_null$ess, res_subset$ess),
  stringsAsFactors = FALSE
)
knitr::kable(
  scenario_tab, digits = c(0L, 0L, 4L, 4L, 1L),
  col.names = c("Comparison", "Outputs tested", "Statistic", "p-value",
                "ESS"),
  caption = "The three scenario comparisons in this vignette."
)
```

### The figure: the ensembles the test is comparing

```{r fig-scenarios, eval = has_ggplot2, fig.height = 5, fig.cap = "Forward-simulated output distributions under the two scenarios, one panel per observation column, across the 60 posterior realisations. The intervention shifts every column by the same amount, which is what the scenario test detects as one verdict over all four columns rather than four separate comparisons."}
ens_df <- rbind(
  data.frame(
    value = as.numeric(out_baseline),
    output = rep(colnames(out_baseline), each = nrow(out_baseline)),
    scenario = "baseline", stringsAsFactors = FALSE
  ),
  data.frame(
    value = as.numeric(out_intervention),
    output = rep(colnames(out_intervention),
                 each = nrow(out_intervention)),
    scenario = "intervention", stringsAsFactors = FALSE
  )
)
ggplot2::ggplot(ens_df, ggplot2::aes(x = value, fill = scenario)) +
  ggplot2::geom_density(alpha = 0.55, colour = NA) +
  ggplot2::facet_wrap(~output, scales = "free") +
  ggplot2::scale_fill_viridis_d(name = "scenario", end = 0.85) +
  ggplot2::labs(
    x = "simulated output (model units)", y = "density",
    title = "Two scenarios from one calibrated posterior"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig-scenarios-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the scenario-ensemble",
    "figure is skipped.\n")
```

### What the comparison refuses to do

`dr_date_scenario()` stops rather than proceed when the two manifests
disagree on the parameter column names, because incompatible priors
cannot be compared; on the observation column names, because different
outputs cannot be compared; or on the PESTO major and minor version,
because a manifest written by one version is not guaranteed readable as
the same object by another. A silent comparison of incompatible
scenarios is a worse failure than a noisy refusal, which is why these are
errors and not warnings.

## Read

The shifted intervention is detected. The scenario test returns a
statistic of `r signif(res_shift$statistic, 3)` with a p-value of
`r round(res_shift$p_value, 4)`, the smallest value
`r res_shift$n_permutations` permutations can express, across all
`r length(res_shift$outputs_tested)` output columns of the manifest and
`r res_shift$n_baseline` realisations per scenario. The effective sample
size of the weights is `r round(res_shift$ess, 1)` out of
`r res_shift$n`, and `ess_warning` is `r res_shift$ess_warning`.

Comparing the baseline against a replicate of itself returns a p-value of
`r round(res_null$p_value, 3)`. The two ensembles are the same numbers,
so the observed statistic is one draw from its own permutation null and
the test correctly declines to find an effect; a scenario test that
rejected here would be reporting the permutation scheme rather than the
data.

Restricting to two of the four output columns leaves the verdict intact,
p-value `r round(res_subset$p_value, 4)` on
`r paste(res_subset$outputs_tested, collapse = " and ")` at
`r res_subset$n_permutations` permutations, which is expected here
because the intervention shifted every column equally. On a real
intervention that moved one output and not another, the restricted and
unrestricted tests would part company, and the restricted one is the
question worth asking.

The figure shows what the single verdict is summarising: four pairs of
distributions, each pair separated by the same displacement. The test
does not report four p-values with a multiplicity problem; it reports one
verdict over the joint output, which is why the manifest carries all four
columns rather than being called once per column.

## Limits

The forward model here is a synthetic linear-Gaussian stand-in and the
intervention is a constant displacement, so this vignette demonstrates
the plumbing and the verdict, not the behaviour of a crop model under a
management change. The shared-posterior construction is the easy case:
the doubly robust correction has nothing to correct, and the test reduces
to a balanced two-sample comparison. When the two scenarios come from
*different* calibration data the parameter posteriors differ, the full
doubly robust correction runs, and it assumes positivity: every
parameter region with non-zero density under one scenario must have
non-zero density under the other. Two completely separated posteriors
give the propensity model perfect separation and the test becomes
uninformative, returning a p-value near one; the way out is a shared
prior with overlapping calibration data, or a restriction to outputs
whose causal pathway does not involve the separating parameters. The
proxymix density-ratio backend that `bd_hsic_test()` already offers, and
which *Choosing a density-ratio backend* compares, is not among this
function's propensity options. Finally the null case is a single draw and
demonstrates a verdict, not a type-I error rate.

## What to read next

*A treatment that changes the spread and not the mean* is the same
DR-DATE statistic applied to a field record rather than a pair of
simulator ensembles, and explains the doubly robust construction.
*Is the calibrated model right about more than the mean?* asks the other
question one puts to a calibrated ensemble: not whether two scenarios
differ, but whether either of them is consistent with data held out of
the calibration. *Which parameters are worth calibrating?* covers the
screen that decides which parameters enter the calibration in the first
place.

## References

- Fawkes, J., Hu, R., Evans, R. J., & Sejdinovic, D. (2024). Doubly
  robust kernel statistics for testing distributional treatment effects.
  *Transactions on Machine Learning Research*.

## Reproduce

Seed `20260516` for the synthetic forward model, observations and prior
ensemble; `seed` values `1`, `2` and `3` for the three scenario tests.
The manifests record PESTO
`r as.character(packageVersion("PESTO"))`. Package versions follow.

```{r session-info}
sessionInfo()
```
