---
title: "Causal Association Testing with bd-HSIC"
author: "Max Moldovan and Dino Sejdinovic"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Association Testing with bd-HSIC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
```

## The Problem

Standard independence tests cannot distinguish *causal* association from
*confounded* association. If treatment X and outcome Y share a common
cause Z, they will appear dependent even if X has no causal effect on Y.

The **bd-HSIC** test (Hu, Sejdinovic & Evans, 2024) solves this by testing
the *do-null* hypothesis:

> H_0: p(y | do(x)) = p*(y) for all x

This uses Pearl's do-operator: after intervening on X, is Y still
associated with X?

## How It Works

1. **Density ratio estimation**: Estimate w(x, z) = p*(x) / p(x|z)
   to reweight observational samples to the interventional distribution.
2. **Weighted HSIC**: Compute HSIC between X and Y under the reweighted
   (interventional) distribution.
3. **Cluster-based permutation**: Obtain p-values by permuting Y within
   clusters of similar conditional densities p(x|z).

## Example: Linear Causal Effect

```{r linear-effect}
library(kernR)
set.seed(42)

n <- 300
z <- matrix(rnorm(n * 2), n, 2)
x <- 0.5 * z[, 1] + rnorm(n)           # X depends on confounder Z
y <- 0.8 * x + 0.5 * z[, 2] + rnorm(n) # Y depends causally on X and on Z

result <- bd_hsic_test(x, y, z,
  n_permutations = 200,
  seed = 1
)
result
```

The test detects the causal association between X and Y.

## Example: No Causal Effect (Confounding Only)

```{r no-effect}
set.seed(42)
n <- 300
z <- matrix(rnorm(n * 2), n, 2)
x <- 0.5 * z[, 1] + rnorm(n)
y <- 0.5 * z[, 1] + z[, 2] + rnorm(n) # Y depends on Z, not on X

result_null <- bd_hsic_test(x, y, z,
  n_permutations = 200,
  seed = 1
)
result_null
```

The large p-value correctly indicates no causal effect.

## Example: Non-Linear Causal Effect

A key advantage of bd-HSIC is detecting non-linear effects that
linear methods (PDS, Double ML) completely miss:

```{r nonlinear-effect}
set.seed(42)
n <- 400
z <- matrix(rnorm(n * 2), n, 2)
x <- z[, 1] + rnorm(n)
y <- x^2 + z[, 2] + rnorm(n, sd = 0.5) # Quadratic causal effect

result_nl <- bd_hsic_test(x, y, z,
  n_permutations = 200,
  seed = 1
)
result_nl
```

## Diagnostic: Null Distribution

```{r plot-null, fig.cap = "bd-HSIC permutation null distribution."}
plot(result)
```

## Using the Formula Interface

```{r formula}
dat <- data.frame(y = y, x = x, z1 = z[, 1], z2 = z[, 2])
result_f <- kernel_causal_test(
  y ~ x | z1 + z2,
  data = dat,
  method = "bd-hsic",
  n_permutations = 100,
  seed = 1
)
result_f
```

## When to Use bd-HSIC

| Scenario | Use bd-HSIC? |
|----------|-------------|
| Testing if X causally affects Y (adjusting for Z) | Yes |
| Non-linear or non-monotone causal effects | Yes -- key advantage |
| Continuous, binary, or mixed treatments | Yes |
| Very high-dimensional confounders | Consider using `density_ratio = "ranger"` |
| Extremely strong confounding | Caution: density ratio estimation may fail |

## References

- Hu, R., Sejdinovic, D., & Evans, R. J. (2024). A kernel test for
  causal association via noise contrastive backdoor adjustment. *JMLR*,
  25(160), 1-56.

```{r session-info}
sessionInfo()
```
