---
title: "Posterior-Predictive Check with MMD"
author: "Max Moldovan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior-Predictive Check with MMD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
set.seed(1)
```

## The question

After running an iterative ensemble smoother (PESTO, PEST++, EnKF, ...)
against historical observations, *does the calibrated model produce a
posterior-predictive distribution that is consistent with a held-out year
or paddock?*

The standard answer in PEST++-style workflows is RMSE on the
posterior-predictive mean. That can only catch differences in central
tendency. A distribution can match in mean but fail dramatically in
variance, skewness, or tail probability -- precisely the failures that
matter for climate-risk decisions.

`mmd_ppc()` answers the distributional version of the question: are the
posterior-predictive draws and the held-out observations samples from
the same distribution? It is a model-free, kernel two-sample test
(MMD) plus a Bayesian-flavoured *surprise* diagnostic.

## The stub ensemble

Until PESTO ships its native manifest emitter, we construct a
`pesto_ensemble` object directly. This is the kernR-side schema for the
cross-package contract.

```{r ensemble}
library(kernR)

# True data-generating distribution: bivariate (yield, biomass)
n_obs <- 30L

# Held-out observations from the "real" world
truth_mean <- c(yield = 4.5, biomass = 12.0)
truth_sd   <- c(yield = 0.6, biomass = 1.5)
observed <- cbind(
  yield   = stats::rnorm(n_obs, truth_mean["yield"],   truth_sd["yield"]),
  biomass = stats::rnorm(n_obs, truth_mean["biomass"], truth_sd["biomass"])
)
```

## Case A: well-calibrated posterior

```{r calibrated}
M <- 300L
post_good <- cbind(
  yield   = stats::rnorm(M, truth_mean["yield"],   truth_sd["yield"]),
  biomass = stats::rnorm(M, truth_mean["biomass"], truth_sd["biomass"])
)
ens_good <- pesto_ensemble(
  posterior = post_good,
  observed  = observed,
  metadata  = list(holdout_year = 2018L, ies_iters = 6L)
)
ens_good

fit_good <- mmd_ppc(ens_good, n_permutations = 299L, seed = 1L)
fit_good
```

A small `surprise_bits` (well below `log2(B + 1)`) and a verdict of
*consistent with observations* mean the calibrated model has not been
falsified by the held-out data at the distributional level.

## Case B: miscalibrated posterior (variance too narrow)

A common ensemble-smoother pathology is over-confidence: posterior draws
clustered too tightly around the posterior mean.

```{r miscalibrated-variance}
post_narrow <- cbind(
  yield   = stats::rnorm(M, truth_mean["yield"],   truth_sd["yield"] / 3),
  biomass = stats::rnorm(M, truth_mean["biomass"], truth_sd["biomass"] / 3)
)
ens_narrow <- pesto_ensemble(post_narrow, observed,
                             metadata = list(scenario = "narrow"))
fit_narrow <- mmd_ppc(ens_narrow, n_permutations = 299L, seed = 1L)
fit_narrow
```

The verdict here should be REJECT, with surprise well above the
threshold of ~4.32 bits (the surprise corresponding to `p = 0.05`).

## Case C: miscalibrated posterior (mean-shifted)

```{r miscalibrated-mean}
post_shifted <- cbind(
  yield   = stats::rnorm(M, truth_mean["yield"]   + 0.8, truth_sd["yield"]),
  biomass = stats::rnorm(M, truth_mean["biomass"] - 1.2, truth_sd["biomass"])
)
ens_shifted <- pesto_ensemble(post_shifted, observed)
fit_shifted <- mmd_ppc(ens_shifted, n_permutations = 299L, seed = 1L)
fit_shifted
```

## Comparing surprise across scenarios

```{r summary-table}
res <- list(
  calibrated   = fit_good,
  narrow       = fit_narrow,
  mean_shifted = fit_shifted
)
data.frame(
  scenario      = names(res),
  statistic     = vapply(res, function(z) z$statistic, numeric(1)),
  p_value       = vapply(res, function(z) z$p_value, numeric(1)),
  surprise_bits = vapply(res, function(z) z$surprise_bits, numeric(1)),
  reject        = vapply(res, function(z) z$reject, logical(1)),
  row.names     = NULL
)
```

## Visual diagnostic

The inherited `plot()` method shows the permutation null distribution
with the observed MMD statistic marked:

```{r plot}
plot(fit_narrow)
```

## Notes on practice

- **Ensemble size.** Distributional power is set by both the ensemble
  size `M` and the held-out sample size `n_obs`. For ag-scale problems
  `M` in the low hundreds is usually sufficient; `n_obs >= 20` is a
  practical floor.
- **Permutations.** At `B = 299` the minimum achievable p-value is
  `1 / 300 ~ 0.0033`; correspondingly the maximum surprise is
  `log2(300) ~ 8.23` bits. Increase `B` when working in the small-p
  regime.
- **Interpreting surprise.** Surprise is Shannon information of the
  permutation p-value: 0 bits at `p = 1`, 4.32 bits at `p = 0.05`, and
  capped at `log2(B + 1)`. It is *not* the strict Bayesian-surprise
  KL divergence; use it as an intuitive scalar verdict, not as a
  posterior probability.
- **Multi-dimensional outputs.** `mmd_ppc()` handles `d > 1` natively
  via the kernel choice; the default RBF with median-heuristic
  bandwidth is computed over the pooled posterior + observed sample.

## Cross-package handoff

PESTO 0.3.0 ships a native ensemble emitter — the
`PESTO::pesto_ensemble_manifest` S7 class — which `mmd_ppc()` consumes
via a dedicated method. The legacy lightweight `pesto_ensemble` S3
constructor is unchanged and remains supported.

```{r,eval=requireNamespace("PESTO",quietly=TRUE)}
library(PESTO)

# Build a tiny synthetic manifest (in real workflows: come from
# pesto_ies_callback() + as_manifest()).
nreal <- 80L; nobs <- 3L
post   <- matrix(rnorm(nreal * nobs), nreal, nobs)
colnames(post) <- paste0("o", seq_len(nobs))
m <- pesto_ensemble_manifest(
  run_id          = "vignette_demo",
  params          = data.frame(real_name = paste0("r", seq_len(nreal)),
                                p1 = rnorm(nreal),
                                p2 = rnorm(nreal),
                                check.names = FALSE),
  outputs         = data.frame(real_name = paste0("r", seq_len(nreal)),
                                post, check.names = FALSE),
  weights         = setNames(rep(1, nobs), colnames(post)),
  obs_target      = setNames(rnorm(nobs),  colnames(post)),
  data_hash       = "sha256:vignette_demo",
  pesto_version   = as.character(packageVersion("PESTO")),
  timestamp       = Sys.time(),
  method          = "ies_callback",
  noptmax         = 1L,
  lambda_schedule = 1
)

# Out-of-sample PPC: supply held-out observations explicitly. The
# manifest's `obs_target` slot is a single nobs-dim point (the data
# the posterior was fit to) and is not a valid two-sample comparator
# on its own.
held_out <- matrix(rnorm(20L * nobs), 20L, nobs)
colnames(held_out) <- colnames(post)
mmd_ppc(m, observed = held_out, n_permutations = 199L, seed = 1L)
```

The `outputs =` argument focuses the check on specific observation
columns (parallel to `dr_date_scenario()`'s convention).

## References

- Gelman, A., Meng, X.-L., & Stern, H. (1996). Posterior predictive
  assessment of model fitness via realized discrepancies.
  *Statistica Sinica*, 6(4), 733-760.
- Gretton, A., Borgwardt, K. M., Rasch, M. J., Schölkopf, B., &
  Smola, A. (2012). A kernel two-sample test. *JMLR*, 13, 723-773.
