---
title: "Pre-IES Identifiability Screening with HSIC"
author: "Max Moldovan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pre-IES Identifiability Screening with HSIC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
set.seed(1)
```

## Why screen before calibrating?

Iterative Ensemble-Smoother (IES) calibration of a mechanistic
agricultural model such as APSIM is expensive: each ensemble member is
a full simulator run. Spending realisations on parameters that have no
detectable effect on the outputs of interest is wasted budget.

A pre-IES screen answers a sharper question than variance-based Sobol
analysis: *does this parameter influence the distribution of any output
at all?* HSIC -- the Hilbert-Schmidt Independence Criterion -- detects
both linear and non-linear, mean-shift and tail-shift effects, including
those Sobol misses when variance is preserved.

The workflow is three lines:

1. Build a Latin-hypercube design across plausible parameter ranges.
2. Run the simulator on the design.
3. Pass the design + outputs to `hsic_identifiability()`.

## A stubbed APSIM archetype

Real APSIM workflows are heavy, so this vignette uses a small simulator
stub with four parameters -- two genuinely identifiable, one weakly
nonlinear, and one inert -- and two outputs (yield and biomass).

```{r stub-simulator}
library(kernR)

stub_apsim <- function(theta) {
  # theta: n x 4 matrix with columns slope, curvature, weak, inert
  yield   <- 1.5 * theta[, "slope"] +
             0.3 * stats::rnorm(nrow(theta), sd = 0.1)
  biomass <- theta[, "curvature"]^2 +
             0.4 * theta[, "weak"] +
             stats::rnorm(nrow(theta), sd = 0.1)
  cbind(yield = yield, biomass = biomass)
}
```

## Latin-hypercube design

```{r design}
bounds <- rbind(
  slope     = c(0.0, 2.0),   # active, linear in yield
  curvature = c(-1.0, 1.0),  # active, quadratic in biomass
  weak     = c(0.0, 1.0),    # weak linear effect on biomass
  inert    = c(0.0, 1.0)     # no effect
)
design <- lhs_design(n = 80L, bounds = bounds, seed = 11L)
head(design)
```

## Simulate and screen

```{r screen}
outputs <- stub_apsim(design)
fit <- hsic_identifiability(
  theta          = design,
  y              = outputs,
  alpha          = 0.05,
  p_adjust       = "BH",
  n_permutations = 299L,
  seed           = 11L
)
fit
```

The print method ranks parameters by their maximum HSIC across outputs.
`slope` and `curvature` should be flagged identifiable with small
p-values; `weak` may or may not survive depending on noise; `inert`
should fall below the threshold.

## Visual diagnostic

```{r plot}
plot(fit)
```

## Handing off to PESTO

The identifiable subset is precisely the set you would forward to
`pesto_ies()` (PESTO's IES wrapper) as the parameter prior. In pseudo-R:

    identifiable_params <- names(fit$identifiable)[fit$identifiable]
    pesto::pesto_ies(..., params = identifiable_params, ...)

For the cross-package workflow specification (kernR -> PESTO -> APSIM
ensemble loop), see the architecture document
`uq_ag_stack_design_v0.md`.

## Tabular form

For downstream reporting (or to feed into a manuscript table), call
`as.data.frame()`:

```{r as-df}
df <- as.data.frame(fit)
head(df)
```

## Notes on practice

- **Design size.** `n` should comfortably exceed the number of
  parameters; we recommend `n >= 10p` as a starting point for
  ag-systems-scale designs.
- **Permutations.** 199-499 permutations is usually sufficient; the
  minimum achievable p-value at `n_permutations` is
  `1 / (n_permutations + 1)`, so `n_permutations >= 199` is the floor
  for testing at `alpha = 0.05`.
- **Multiple testing.** With many parameters and several outputs,
  unadjusted p-values will produce spurious "identifiable" flags. The
  default Benjamini-Hochberg adjustment controls the false discovery
  rate across the `p x q` grid.
- **Cost.** Kernel matrices are reused across the grid, so cost scales
  as `O((p + q) n^2)` for kernel construction plus
  `O(p q B n^2)` for the permutation null, where `B = n_permutations`.

## References

- Hu, Z., Sejdinovic, D., & Evans, R. J. (2024).
  *A kernel-based statistical test for causal inference with backdoor
  adjustment.* Journal of Machine Learning Research, 25.
- Gretton, A., Fukumizu, K., Teo, C. H., Song, L., Schölkopf, B., &
  Smola, A. J. (2008). A kernel statistical test of independence.
  *NeurIPS*, 20.
- Da Veiga, S. (2015). Global sensitivity analysis with dependence
  measures. *Journal of Statistical Computation and Simulation*,
  85(7), 1283-1305.
