---
title: "Does the treatment cause the outcome, or do they share a cause?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Does the treatment cause the outcome, or do they share a cause?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. Four bd-HSIC tests at 199 to 299 cluster permutations on 300 to
     400 observations, plus one density-ratio fit per test. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

An agronomist has an observational record rather than a randomised
trial. *Paddocks that got the higher residue-retention intensity also
happen to be the wetter, higher-nitrogen paddocks. Yield and management
move together in my data, but so would they if management did nothing at
all and the soil were doing all the work. Can I ask whether management
moves yield, having subtracted what the soil explains, without first
assuming the effect is a straight line?* This vignette answers that with
the backdoor-adjusted HSIC test, on three synthetic designs where the
right answer is known by construction: a linear causal effect, no causal
effect at all, and a curved causal effect. It closes with the case in
which the test declines to answer.

## What

`bd_hsic_test()` tests the do-null hypothesis of Hu, Sejdinovic and Evans
(2024),

$$H_0 : p(y \mid do(x)) = p^{*}(y) \quad \text{for all } x,$$

which asks whether the outcome is still associated with the treatment
*after* intervening on the treatment, rather than merely in the record as
observed. It gets there in three steps. First it estimates the density
ratio $w(x, z) = p^{*}(x) / p(x \mid z)$, which reweights the
observational sample so that it looks like a sample in which treatment
was assigned independently of the confounders; `estimate_density_ratio()`
and `fit_density_ratio()` expose that step on its own, with four
interchangeable backends. Second it computes HSIC between treatment and
outcome under those weights, so any dependence the kernel can represent
counts, curved or otherwise. Third it builds the null by permuting the
outcome *within clusters of similar conditional density*
$p(x \mid z)$, so the permutation respects the confounding structure
instead of destroying it.

`kernel_causal_test()` is the formula front door to the same machinery:
`y ~ x | z1 + z2` names outcome, treatment and adjustment set in one
expression. Every call returns a `kernel_test_result` carrying the
statistic, the permutation null, the p-value, and the effective sample
size (ESS) of the density-ratio weights -- the diagnostic that decides
whether the verdict is worth anything, which is the subject of the last
section.

## Do

### A linear causal effect

A synthetic design of 300 observations. A two-column confounder drives
both treatment and outcome, and the treatment additionally has a genuine
linear effect on the outcome of 0.8 per unit.

```{r linear-effect}
set.seed(42L)
n_obs <- 300L
z_conf <- matrix(rnorm(n_obs * 2L), n_obs, 2L)
x_lin <- 0.5 * z_conf[, 1L] + rnorm(n_obs)
y_lin <- 0.8 * x_lin + 0.5 * z_conf[, 2L] + rnorm(n_obs)

fit_linear <- bd_hsic_test(x_lin, y_lin, z_conf,
  n_permutations = 299L, seed = 1L
)
fit_linear
```

### No causal effect, confounding only

The same confounder and the same treatment, but the outcome is now built
from the confounder alone. Treatment and outcome are still correlated in
the record; nothing flows from one to the other.

```{r no-effect}
set.seed(42L)
z_null <- matrix(rnorm(n_obs * 2L), n_obs, 2L)
x_null <- 0.5 * z_null[, 1L] + rnorm(n_obs)
y_null <- 0.5 * z_null[, 1L] + z_null[, 2L] + rnorm(n_obs)
marginal_r <- cor(x_null, y_null)
round(marginal_r, 3L)

fit_confounded <- bd_hsic_test(x_null, y_null, z_null,
  n_permutations = 299L, seed = 1L
)
fit_confounded
```

### A curved causal effect

The advantage of a kernel statistic over a partialling-out method is that
it does not have to be told the shape of the effect. Here the treatment
enters the outcome as its square, so the average slope is close to zero.

```{r nonlinear-effect}
set.seed(42L)
n_quad <- 400L
z_quad <- matrix(rnorm(n_quad * 2L), n_quad, 2L)
x_quad <- z_quad[, 1L] + rnorm(n_quad)
y_quad <- x_quad^2 + z_quad[, 2L] + rnorm(n_quad, sd = 0.5)

fit_quadratic <- bd_hsic_test(x_quad, y_quad, z_quad,
  n_permutations = 299L, seed = 1L
)
fit_quadratic
```

### The formula interface

```{r formula}
dat_quad <- data.frame(
  y = y_quad, x = x_quad, z1 = z_quad[, 1L], z2 = z_quad[, 2L]
)
fit_formula <- kernel_causal_test(
  y ~ x | z1 + z2,
  data = dat_quad, method = "bd-hsic",
  n_permutations = 299L, seed = 1L
)
fit_formula
```

### The figure: two nulls, two verdicts

```{r fig-nulls, eval = has_ggplot2, fig.cap = "Cluster-permutation null distributions of the weighted HSIC statistic for the linear-effect design and the confounding-only design, with each observed statistic marked by a vertical line in the matching colour. The two nulls are nearly the same shape; only the observed statistics separate the two designs."}
null_df <- rbind(
  data.frame(
    statistic = fit_linear$null_distribution, design = "linear effect"
  ),
  data.frame(
    statistic = fit_confounded$null_distribution,
    design = "confounding only"
  )
)
obs_df <- data.frame(
  design = c("linear effect", "confounding only"),
  statistic = c(fit_linear$statistic, fit_confounded$statistic)
)
ggplot2::ggplot(null_df, ggplot2::aes(x = statistic, fill = design)) +
  ggplot2::geom_histogram(bins = 40L, alpha = 0.65,
                          position = "identity", colour = NA) +
  ggplot2::geom_vline(
    data = obs_df,
    ggplot2::aes(xintercept = statistic, colour = design),
    linewidth = 1
  ) +
  ggplot2::scale_fill_viridis_d(name = "design", end = 0.85) +
  ggplot2::scale_colour_viridis_d(name = "design", end = 0.85) +
  ggplot2::labs(
    x = "weighted HSIC statistic (unitless)",
    y = "cluster permutations",
    title = "Where the observed statistic falls in its own null"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-nulls-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the permutation-null",
    "figure is skipped.\n")
```

```{r verdict-table, echo = FALSE}
verdicts <- data.frame(
  design = c("linear causal effect", "confounding only",
             "quadratic causal effect", "quadratic, formula interface"),
  statistic = c(fit_linear$statistic, fit_confounded$statistic,
                fit_quadratic$statistic, fit_formula$statistic),
  p_value = c(fit_linear$p_value, fit_confounded$p_value,
              fit_quadratic$p_value, fit_formula$p_value),
  ess = c(fit_linear$ess, fit_confounded$ess,
          fit_quadratic$ess, fit_formula$ess),
  n_test = c(fit_linear$n, fit_confounded$n,
             fit_quadratic$n, fit_formula$n),
  stringsAsFactors = FALSE
)
knitr::kable(
  verdicts, digits = c(0L, 4L, 4L, 1L, 0L),
  col.names = c("Design", "Statistic", "p-value", "ESS", "Test n"),
  caption = paste0(
    "bd-HSIC verdicts on the three synthetic designs, each against a ",
    fit_linear$n_permutations, "-draw cluster-permutation null."
  )
)
```

### When the test declines to answer

The density-ratio weights are the whole adjustment, so a verdict is worth
exactly as much as they are. `bd_hsic_test()` measures that with the
effective sample size of the weights and refuses to stand behind a
p-value computed from a handful of dominant observations. The design
below is the ordinary way that happens in a field record: too many
covariates for too few plots. Fifty plots carry a 24-column soil and
climate block, and the management variable is a function of the whole
block, so the reweighting has almost no independent information to work
with.

```{r ess-floor}
set.seed(9L)
n_plots <- 50L
n_cov <- 24L
z_thin <- matrix(rnorm(n_plots * n_cov), n_plots, n_cov)
x_thin <- rowSums(z_thin) / sqrt(n_cov) + rnorm(n_plots)
y_thin <- 0.5 * x_thin + rnorm(n_plots)

gate_message <- NULL
fit_thin <- withCallingHandlers(
  bd_hsic_test(x_thin, y_thin, z_thin, n_permutations = 199L, seed = 1L),
  warning = function(w) {
    gate_message <<- conditionMessage(w)
    invokeRestart("muffleWarning")
  }
)
is_kernR_abstention(fit_thin)
class(fit_thin)
```

The gate states its own case; the message is captured above rather than
printed as a condition so that this vignette renders without warnings.

```{r ess-floor-message, echo = FALSE, results = "asis"}
cat("> ", gate_message, "\n", sep = "")
```

## Read

On the linear design the weighted HSIC statistic is
`r signif(fit_linear$statistic, 3)` against a null whose draws reach
`r signif(max(fit_linear$null_distribution), 3)`, giving a p-value of
`r round(fit_linear$p_value, 4)`: the effect survives adjustment, which
it should, because it was built in. On the confounding-only design the
raw correlation between treatment and outcome is
`r round(marginal_r, 3)` -- an association any unadjusted test would
report -- yet the adjusted statistic is
`r signif(fit_confounded$statistic, 3)` with a p-value of
`r round(fit_confounded$p_value, 3)`. The reweighting has removed the
association, which is the whole point: what looked like an effect was the
soil.

The curved design is where a kernel statistic separates from a linear
adjustment. The treatment enters the outcome as its square, so the
average slope over the sample is close to zero and a partialling-out
method has nothing to find; bd-HSIC returns
`r signif(fit_quadratic$statistic, 3)` with a p-value of
`r round(fit_quadratic$p_value, 4)`. The formula interface reaches the
same verdict on the same data, statistic
`r signif(fit_formula$statistic, 3)`, p-value
`r round(fit_formula$p_value, 4)`.

The figure carries the part the table cannot. The two permutation nulls
are almost the same distribution: the null does not know which design
produced it, because within-cluster permutation preserves the
confounding structure in both cases. What differs is where the observed
statistic lands, and that is what the p-values report.

The last design is the one an analyst most needs to recognise. The
effective sample size of the density-ratio weights collapses to
`r round(fit_thin$ess, 1)` out of `r fit_thin$n` test observations, a
fraction of `r round(fit_thin$ess_fraction, 3)`, below the default floor
of `r fit_thin$min_ess_fraction`. The test still returns a p-value --
`r round(fit_thin$p_value, 3)` on this run -- and that number should not
be used. `is_kernR_abstention()` is
`r is_kernR_abstention(fit_thin)` and the result carries the class
`kernR_abstention` ahead of `kernel_test_result`, which is how an
orchestra consumer downstream recognises that kernR declined to stand
behind its own verdict without having to know kernR's field names. The
remedy is in the message: more plots, a different density-ratio backend,
or a smaller adjustment set.

## Limits

The three designs here are synthetic, with the confounding structure
known and the adjustment set complete by construction. No test in this
package can tell you that your adjustment set is complete: bd-HSIC
assumes the backdoor criterion holds for the covariates you supplied, and
an omitted confounder produces a confident wrong answer with a healthy
ESS. The density-ratio step is a fitted model too, so a badly specified
ratio biases the weighted statistic even when the ESS looks acceptable;
the ESS floor catches collapse, not misspecification. The cluster
permutation assumes exchangeability within its propensity-similarity
clusters, which is a weaker assumption than independence but not a free
one, and it is the wrong assumption when the design has its own
hierarchy. Finally the p-values above are permutation p-values and cannot
fall below `1 / (n_permutations + 1)`.

## What to read next

*Choosing a density-ratio backend* compares the four backends behind the
adjustment step on one confounded design, and is the place to go when the
ESS diagnostic above is unsatisfactory. *Multi-site trials and the
permutation null* replaces the propensity-similarity clustering used here
with the design's own sites, which is the right choice whenever the
design tells you who is exchangeable with whom. *A treatment that changes
the spread and not the mean* asks the distributional version of the same
causal question for a binary treatment.

## References

- Hu, R., Sejdinovic, D., & Evans, R. J. (2024). A kernel test for causal
  association via noise contrastive backdoor adjustment. *Journal of
  Machine Learning Research*, 25(160), 1-56.

## Reproduce

Seed `42` for the three causal designs and `9` for the collapsed-ESS
design; `seed = 1L` passed to every test so the density-ratio fit and the
permutation draws are fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
