---
title: "Getting Started with kernR"
author: "Max Moldovan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with kernR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
```

## What is kernR?

**kernR** provides kernel-based statistical tests for causal inference and
distributional comparison. It implements:
- **HSIC test**: independence testing via the Hilbert-Schmidt Independence Criterion
- **MMD test**: two-sample testing via Maximum Mean Discrepancy
- **bd-HSIC test**: causal association testing with backdoor adjustment
- **DR-DATE / DR-DETT**: doubly robust distributional treatment effect tests

This vignette covers the basics: kernels, MMD, and HSIC.

## Kernel Basics

A *kernel* is a function that measures similarity between observations. kernR
supports RBF (Gaussian), Matern, linear, and polynomial kernels.

```{r kernel-spec}
library(kernR)

# Default: RBF kernel with automatic bandwidth (median heuristic)
k <- kernel_spec()
k

# Fixed bandwidth
k_fixed <- kernel_spec("rbf", bandwidth = 1.5)
k_fixed

# Linear kernel
k_lin <- kernel_spec("linear")
k_lin
```

## Computing Kernel Matrices

```{r kernel-matrix}
set.seed(42)
x <- matrix(rnorm(200), 100, 2)

# Compute the 100 x 100 kernel (Gram) matrix
K <- kernel_matrix(x)
dim(K)

# Visualise a corner
K[1:5, 1:5]
```

## Two-Sample Testing with MMD

The MMD test asks: *do two samples come from the same distribution?*

```{r mmd-same}
set.seed(123)

# Two samples from the same distribution
x <- matrix(rnorm(200), 100, 2)
y <- matrix(rnorm(200), 100, 2)

result <- mmd_test(x, y, seed = 1)
result
```

The p-value is large — no evidence of different distributions. Now with a
mean shift:
```{r mmd-different}
y_shifted <- matrix(rnorm(200, mean = 0.5), 100, 2)
result <- mmd_test(x, y_shifted, seed = 1)
result
```

The small p-value correctly detects the distributional difference.

## Independence Testing with HSIC

HSIC tests whether two variables are independent — including non-linear
dependencies that correlation would miss.

```{r hsic-nonlinear}
set.seed(456)
n <- 300
x <- rnorm(n)

# Non-linear dependence: Y = X^2 + noise
# Note: cor(x, y) is approximately 0 (no linear correlation)
y <- x^2 + rnorm(n, sd = 0.3)
cat("Pearson correlation:", round(cor(x, y), 3), "\n")

# HSIC detects the non-linear dependence
result <- hsic_test(x, y, seed = 1)
result
```

HSIC successfully detects the quadratic relationship even though the
Pearson correlation is near zero.

## Visualising Results

Every test result can be plotted to see where the observed statistic
falls relative to the permutation null distribution:

```{r plot-result, fig.cap = "HSIC permutation null distribution with observed statistic (dashed red line)."}
plot(result)
```

## Next Steps
- `vignette("kernR-bdhsic")` — Causal association testing with bd-HSIC
- `vignette("kernR-drtest")` — Distributional treatment effect tests
- `vignette("kernR-hierarchical")` — Tests for hierarchical/nested data

```{r session-info}
sessionInfo()
```
