---
title: "When plots sit inside farms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When plots sit inside farms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~11 s under rmarkdown::render(); macOS aarch64, R 4.5.2,
     one core. Five hierarchical DR-DATE tests at 99 within-cluster
     permutations on 600 observations in 20 clusters; the permutation loop
     dominates. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A trial agronomist has 600 plots, but not 600 independent plots. *The
fertiliser trial ran on twenty farms, thirty plots each. The farms differ
in ways I did not measure and cannot control for, and every test I know
assumes my plots are independent, which they plainly are not. If I hand
the whole file to a two-sample test, am I measuring the fertiliser or am
I measuring the farms?* This vignette answers that on a synthetic
twenty-farm trial: it shows what the clustered test does with the data,
how much of the signal is within farms and how much is between them, and
what the same machinery says when the treatment does nothing at all.

## What

`hierarchical_test()` runs any of the package's causal tests on data with
a known grouping, and does two things a flat test does not. It
*decomposes* the statistic into a within-cluster part -- the test run
inside each cluster and averaged -- and a between-cluster part, computed
on the cluster-level summaries; and it builds the permutation null by
shuffling *within* clusters only, so cluster-level variation stays in the
null instead of being counted as signal.

The two parts are combined according to `weight_method`. `"equal"` gives
the within and between components the same weight; `"icc"` weights them
by the intraclass correlation estimated from the data, moving weight to
the between component when the clusters genuinely differ; `"within_only"`
discards the between component, which is the conservative choice when
cluster-level comparisons are not credible. `method` selects the test
that runs inside the decomposition, here `"dr-date"`, the doubly robust
distributional treatment-effect test described in *A treatment that
changes the spread and not the mean*.

The result is a `kernel_test_result` with an extra `hierarchical`
element carrying `within_stats` (one statistic per cluster),
`between_stat`, `n_clusters` and the `weight_method` used.

## Do

### A synthetic twenty-farm trial

Twenty farms, thirty plots each. Each farm carries its own random effect
on yield with standard deviation 2, which is twice the residual noise;
soil covariates drive treatment assignment, so the design is
observational rather than randomised; and the treatment has a real
average effect of 0.8 yield units.

```{r trial-design}
set.seed(42L)
n_farms <- 20L
n_plots <- 30L
n_obs <- n_farms * n_plots
farm_id <- rep(seq_len(n_farms), each = n_plots)
farm_effect <- rnorm(n_farms, sd = 2)[farm_id]
soil <- matrix(rnorm(n_obs * 2L), n_obs, 2L)
treatment <- rbinom(n_obs, 1L, plogis(0.3 * soil[, 1L]))
yield <- 0.8 * treatment + farm_effect + 0.5 * soil[, 1L] +
  rnorm(n_obs)
```

```{r trial-test}
fit_icc <- hierarchical_test(
  y = yield, treatment = treatment, covariates = soil,
  cluster_id = farm_id, method = "dr-date",
  n_permutations = 99L, weight_method = "icc", seed = 1L
)
fit_icc
```

### The decomposition

```{r decomposition}
within_mean <- mean(fit_icc$hierarchical$within_stats, na.rm = TRUE)
between <- fit_icc$hierarchical$between_stat
round(c(within_mean = within_mean, between = between,
        combined = fit_icc$statistic), 4L)
fit_icc$hierarchical$n_clusters
```

### The three weighting conventions on the same data

```{r weight-methods}
fit_equal <- hierarchical_test(
  y = yield, treatment = treatment, covariates = soil,
  cluster_id = farm_id, method = "dr-date",
  n_permutations = 99L, weight_method = "equal", seed = 1L
)
fit_within <- hierarchical_test(
  y = yield, treatment = treatment, covariates = soil,
  cluster_id = farm_id, method = "dr-date",
  n_permutations = 99L, weight_method = "within_only", seed = 1L
)
```

```{r weight-table, echo = FALSE}
weight_tab <- data.frame(
  weight_method = c("equal", "icc", "within_only"),
  behaviour = c(
    "within and between weighted alike",
    "weighted by the estimated intraclass correlation",
    "between component discarded"
  ),
  statistic = c(fit_equal$statistic, fit_icc$statistic,
                fit_within$statistic),
  p_value = c(fit_equal$p_value, fit_icc$p_value, fit_within$p_value),
  stringsAsFactors = FALSE
)
knitr::kable(
  weight_tab, digits = c(0L, 0L, 4L, 4L),
  col.names = c("weight_method", "Behaviour", "Statistic", "p-value"),
  caption = paste0(
    "The same trial under the three weighting conventions, each against ",
    "a ", fit_icc$n_permutations, "-draw within-cluster permutation null."
  )
)
```

### The same trial with no treatment effect

The identical farms, plots, soil and treatment assignment, with the
treatment term removed from the yield.

```{r null-trial}
set.seed(42L)
yield_null <- farm_effect + 0.5 * soil[, 1L] + rnorm(n_obs)
fit_null <- hierarchical_test(
  y = yield_null, treatment = treatment, covariates = soil,
  cluster_id = farm_id, method = "dr-date",
  n_permutations = 99L, weight_method = "icc", seed = 1L
)
fit_null$p_value
```

### The figure: where the signal sits

```{r fig-decomposition, eval = has_ggplot2, fig.cap = "Per-farm within-cluster DR-DATE statistic for the trial with a real treatment effect (points), against the single between-cluster statistic (dashed line) and the mean of the within-cluster statistics (solid line). The spread across farms is what the combined statistic averages over, and what a flat test would silently mix with the between-farm contrast."}
within_df <- data.frame(
  farm = seq_along(fit_icc$hierarchical$within_stats),
  statistic = fit_icc$hierarchical$within_stats
)
ggplot2::ggplot(within_df, ggplot2::aes(x = farm, y = statistic)) +
  ggplot2::geom_point(colour = "#440154", size = 2.5) +
  ggplot2::geom_hline(
    yintercept = within_mean, colour = "#21918c", linewidth = 0.9
  ) +
  ggplot2::geom_hline(
    yintercept = between, colour = "#FDE725", linewidth = 0.9,
    linetype = "dashed"
  ) +
  ggplot2::annotate(
    "text", x = 1, y = within_mean, label = "within-cluster mean",
    hjust = 0, vjust = -0.7, colour = "#21918c"
  ) +
  ggplot2::annotate(
    "text", x = 1, y = between, label = "between-cluster statistic",
    hjust = 0, vjust = -0.7, colour = "#B8A400"
  ) +
  ggplot2::labs(
    x = "farm", y = "DR-DATE statistic (unitless)",
    title = "Within-farm and between-farm contributions"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-decomposition-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the decomposition figure",
    "is skipped.\n")
```

## Read

On the trial with a real effect, the intraclass-correlation weighting
returns a combined statistic of `r signif(fit_icc$statistic, 3)` with a
p-value of `r round(fit_icc$p_value, 3)` against a
`r fit_icc$n_permutations`-draw within-cluster null. The decomposition
says where that comes from: the mean within-farm statistic is
`r signif(within_mean, 3)` and the between-farm statistic is
`r signif(between, 3)`, across `r fit_icc$hierarchical$n_clusters`
clusters. The figure shows the within-farm statistics are not
interchangeable -- they range from
`r signif(min(within_df$statistic, na.rm = TRUE), 3)` to
`r signif(max(within_df$statistic, na.rm = TRUE), 3)` -- so the combined
number is an average over farms that differ in how strongly the treatment
registers, not a single well-determined quantity.

The three weighting conventions give statistics of
`r signif(fit_equal$statistic, 3)` (equal),
`r signif(fit_icc$statistic, 3)` (intraclass correlation) and
`r signif(fit_within$statistic, 3)` (within only). They are not the same
number and are not meant to be: each is a different summary of the same
decomposition, and the choice belongs to the analyst before the data are
seen, not after the p-values are read.

With the treatment term removed the same test returns a p-value of
`r round(fit_null$p_value, 3)`. The farm effects are still there, still
large relative to the residual noise, and the within-cluster permutation
scheme keeps them out of the verdict, which is the reason for using it.

## Limits

This is a synthetic trial, with balanced clusters of exactly thirty plots
and a single level of nesting. Real designs are unbalanced and often
nested more deeply, and `hierarchical_test()` takes one `cluster_id`
vector, so a plots-in-paddocks-in-farms design has to be collapsed to the
level the analyst is willing to call exchangeable. The decomposition
reports the two components on the raw statistic scale and does not
weight the within-cluster statistics by cluster size, so an unbalanced
design should be summarised with care. A single null run cannot
demonstrate a type-I error rate; the null trial above shows one draw, not
a calibration study, and a claim about false-positive control would need
a repeated-sampling experiment this vignette does not run. The
intraclass correlation used by `weight_method = "icc"` is estimated from
the same data as the statistic, so its weight carries the noise of that
estimate.

## What to read next

*Multi-site trials and the permutation null* runs the clustered
permutation scheme through `bd_hsic_test()` instead, for a continuous
treatment, and reports the per-cluster breakdown as a diagnostic for
site-specific effect modification. *A treatment that changes the spread
and not the mean* is the flat version of the test that runs inside the
decomposition here. *Getting started with kernR* introduces the kernel
and permutation machinery all of these share.

## Reproduce

Seed `42` for the synthetic trial and for the null variant; `seed = 1L`
passed to every test so the propensity fit and the within-cluster
permutation draws are fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
