---
title: "Distributional Treatment Effect Tests (DR-DATE / DR-DETT)"
author: "Max Moldovan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional Treatment Effect Tests (DR-DATE / DR-DETT)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
```

## Beyond Mean Effects

Standard causal inference methods (Double ML, TMLE) test whether treatment
shifts the *mean* outcome. But many real treatment effects are
**distributional** -- they change variance, shape, or modality without
necessarily changing the mean.

**DR-DATE** and **DR-DETT** (Fawkes, Hu, Evans & Sejdinovic, 2024) test
for *any* distributional difference between Y(1) and Y(0), using doubly
robust kernel embeddings.

## Key Concepts

- **DR-DATE** (Distributional Average Treatment Effect): Tests whether
  P(Y(1)) = P(Y(0)) over the entire population.
- **DR-DETT** (Distributional Effect on the Treated): Tests whether
  P(Y(1)|T=1) = P(Y(0)|T=1), focusing on the treated subgroup. Requires
  only one-sided overlap.
- **Double robustness**: Consistent if either the propensity model or
  the outcome model is correctly specified.

## Example: Mean Shift (Detectable by All Methods)

```{r mean-shift}
library(kernR)
set.seed(42)

n <- 300
x <- matrix(rnorm(n * 2), n, 2)
logit_p <- 0.3 * x[, 1] - 0.2 * x[, 2]
t <- rbinom(n, 1, plogis(logit_p))
y <- t * 1.0 + 0.5 * x[, 1] + rnorm(n, sd = 0.5) # Mean shift of 1.0

result <- dr_date_test(y, t, x,
  n_permutations = 200,
  seed = 1
)
result
```

## Example: Variance Effect Only (Invisible to Mean-Based Tests)

This is where DR-DATE shines. The treatment changes the *variance* of
the outcome but not the mean -- DML and TMLE would have zero power here.

```{r variance-effect}
set.seed(42)
n <- 400
x <- matrix(rnorm(n * 2), n, 2)
t <- rbinom(n, 1, plogis(0.3 * x[, 1]))

# Treatment doubles the variance but does NOT shift the mean
y <- (1 - t) * rnorm(n, sd = 1) + t * rnorm(n, sd = 2.5) + 0.5 * x[, 1]

cat("Mean difference:", mean(y[t == 1]) - mean(y[t == 0]), "\n")
cat("SD treated:", sd(y[t == 1]), "  SD control:", sd(y[t == 0]), "\n")

result_var <- dr_date_test(y, t, x,
  n_permutations = 200,
  outcome_model = "zero",
  seed = 1
)
result_var
```

## DR-DETT: Effect on the Treated

When overlap is imperfect (some covariate regions have nearly all
treated or all control units), DR-DETT is more robust because it
requires only one-sided overlap.

```{r dett}
set.seed(42)
n <- 300
x <- matrix(rnorm(n * 2), n, 2)
t <- rbinom(n, 1, plogis(0.5 * x[, 1]))
y <- t * rnorm(n, mean = 0.5, sd = 1.5) + (1 - t) * rnorm(n) + x[, 1]

result_dett <- dr_dett_test(y, t, x,
  n_permutations = 200,
  seed = 1
)
result_dett
```

## Comparing the Tests

```{r comparison}
cat("DR-DATE p-value:", result_var$p_value, "\n")
cat("DR-DETT p-value:", result_dett$p_value, "\n")
```

## Using the Formula Interface

```{r formula}
dat <- data.frame(y = y, treatment = t, x1 = x[, 1], x2 = x[, 2])
result_f <- kernel_causal_test(
  y ~ treatment | x1 + x2,
  data = dat,
  method = "dr-date",
  n_permutations = 100,
  seed = 1
)
result_f
```

## When to Use Which Test

| Test | Detects | Overlap Requirement | Best For |
|------|---------|---------------------|----------|
| **DR-DATE** | Any distributional difference | Both sides | Population-level effects |
| **DR-DETT** | Distributional effect on treated | One-sided only | Imperfect overlap; policy questions about treated |
| **DML/TMLE** | Mean shifts only | Both sides | When only mean effects matter |

## References

- Fawkes, J., Hu, R., Evans, R. J., & Sejdinovic, D. (2024). Doubly
  robust kernel statistics for testing distributional treatment effects.
  *Transactions on Machine Learning Research*.

```{r session-info}
sessionInfo()
```
