---
title: "DR-DATE for APSIM Scenario Counterfactuals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DR-DATE for APSIM Scenario Counterfactuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
set.seed(20260516)
```

## The ag-scenario question

Given an APSIM model calibrated on observational data, we want to ask:
*does an intervention (e.g. stubble retention) shift the distribution
of simulated outputs?* The canonical setup is **one PESTO posterior,
forward-simulated under two scenarios** — baseline management and the
intervention. With identical parameter posteriors, the question reduces
to a balanced two-sample distributional comparison on outputs.

`dr_date_scenario()` runs the DR-DATE statistic of Fawkes-Hu-Evans-
Sejdinovic (2024) over a pair of PESTO 0.3.0 `pesto_ensemble_manifest`
objects. With identical parameter columns, the doubly-robust
correction collapses to a balanced MMD on the outputs; with different
parameter posteriors, the DR adjustment absorbs the covariate shift
(provided positivity holds).

```{r}
library(kernR)
library(PESTO)
```

## One calibration, two scenarios

We use a synthetic linear-Gaussian forward model in place of APSIM so
the vignette runs without an APSIM install. Replace `forward` with
`PESTO::apsim_callback(...)` to drive a real APSIM ensemble.

```{r}
npar  <- 2L
nobs  <- 4L
nreal <- 60L
sigma <- 0.05

G <- matrix(rnorm(nobs * npar), nobs, npar)
y_obs <- as.numeric(G %*% c(1.0, -0.5)) + rnorm(nobs, sd = sigma)
names(y_obs) <- paste0("o", seq_len(nobs))

prior <- matrix(rnorm(nreal * npar), nreal, npar,
                dimnames = list(NULL, c("p1", "p2")))
forward <- function(theta) theta %*% t(G)
```

Calibrate **once** with PESTO's in-process IES driver:

```{r}
fit <- pesto_ies_callback(
  forward_model  = forward,
  prior_ensemble = prior,
  obs            = y_obs,
  obs_sd         = sigma,
  noptmax        = 3L,
  verbose        = FALSE
)
```

Now forward-simulate that posterior under two scenarios. The
intervention here is a deterministic shift on the outputs — in a real
ag application this would be e.g. a stubble-retention management rule
that changes APSIM's predicted yield trajectory.

```{r}
par_post   <- as.matrix(fit$par_ensemble[, c("p1", "p2"), with = FALSE])
out_base   <- par_post %*% t(G)
out_intv   <- out_base + 0.6                       # the intervention
colnames(out_base) <- colnames(out_intv) <- names(y_obs)
```

Wrap each scenario as a manifest. **The two manifests share the same
parameter posterior** — that's the canonical structure.

```{r}
real_names <- fit$par_ensemble$real_name

m_baseline <- pesto_ensemble_manifest(
  run_id          = "wagga_baseline_2026",
  params          = data.frame(real_name = real_names, par_post,
                               check.names = FALSE),
  outputs         = data.frame(real_name = real_names, out_base,
                               check.names = FALSE),
  weights         = setNames(rep(1 / sigma, nobs), names(y_obs)),
  obs_target      = setNames(y_obs, names(y_obs)),
  data_hash       = "sha256:vignette_baseline",
  pesto_version   = as.character(packageVersion("PESTO")),
  timestamp       = Sys.time(),
  method          = "ies_callback",
  noptmax         = 3L,
  lambda_schedule = 1
)

m_intervention <- pesto_ensemble_manifest(
  run_id          = "wagga_stubble_2026",
  params          = data.frame(real_name = real_names, par_post,
                               check.names = FALSE),     # SAME params
  outputs         = data.frame(real_name = real_names, out_intv,
                               check.names = FALSE),
  weights         = setNames(rep(1 / sigma, nobs), names(y_obs)),
  obs_target      = setNames(y_obs, names(y_obs)),
  data_hash       = "sha256:vignette_intervention",
  pesto_version   = as.character(packageVersion("PESTO")),
  timestamp       = Sys.time(),
  method          = "ies_callback",
  noptmax         = 3L,
  lambda_schedule = 1
)
m_baseline
```

In production you would build these manifests via `as_manifest()` on
two separate IES runs *only when the scenarios genuinely produced
different calibration data*. For the "did the intervention shift the
forward outputs?" question, the same-posterior construction above is
the right one.

## Run the DR-DATE scenario test

```{r}
res <- dr_date_scenario(
  baseline       = m_baseline,
  intervention   = m_intervention,
  n_permutations = 299L,
  seed           = 1L
)
print(res)
```

With a clear distributional shift, the test should reject and return
a low `p_value`. The `Verdict` line gives the directly actionable
read.

## Null case — same outputs

If the intervention has no effect (shift = 0), the two output
distributions are identical and the test should fail to reject:

```{r}
out_null <- out_base                       # no intervention effect
m_null <- pesto_ensemble_manifest(
  run_id          = "wagga_baseline_replicate",
  params          = data.frame(real_name = real_names, par_post,
                               check.names = FALSE),
  outputs         = data.frame(real_name = real_names, out_null,
                               check.names = FALSE),
  weights         = setNames(rep(1 / sigma, nobs), names(y_obs)),
  obs_target      = setNames(y_obs, names(y_obs)),
  data_hash       = "sha256:vignette_null",
  pesto_version   = as.character(packageVersion("PESTO")),
  timestamp       = Sys.time(),
  method          = "ies_callback",
  noptmax         = 3L,
  lambda_schedule = 1
)
res_null <- dr_date_scenario(
  baseline       = m_baseline,
  intervention   = m_null,
  n_permutations = 299L,
  seed           = 2L
)
res_null$p_value
```

## What gets validated

`dr_date_scenario()` hard-stops if the two manifests don't agree on:

- parameter column names (you can't compare incompatible priors),
- observation column names (you can't compare different outputs),
- PESTO major.minor version (forward / backward incompatibility safety).

This is by design — silent comparison of incompatible scenarios is a
worse failure mode than a noisy abort.

## Output sub-selection

For ensembles with many output columns, focus the test on the outputs
of scientific interest with `output =`:

```{r}
dr_date_scenario(
  m_baseline, m_intervention,
  output         = c("o1", "o3"),     # subset
  n_permutations = 99L, seed = 3L
)$outputs_tested
```

## When parameter posteriors genuinely differ

If the two scenarios came from *different calibration data* (so the
two PESTO runs produced different posteriors), then `params` will
differ between the two manifests. DR-DATE will then run its full
doubly-robust correction, with one important caveat: it assumes
**positivity** — every parameter region with non-zero baseline
density also has non-zero intervention density (and vice versa). If
the two posteriors are completely separated in parameter space, the
test will see "perfect separation" in the propensity-model fit and
become uninformative (`p_value ~ 1`). For that regime, use
`PESTO::pesto_ies_callback()` with a shared prior + overlapping
calibration data, or restrict to outputs whose causal pathway is
independent of the separating parameters.

## Where the cross-package plumbing lives

- Forward-model run: `PESTO::pesto_ies_callback()` (in-process R
  callback) or `PESTO::apsim_callback()` (apsimx adapter).
- Run capture: `PESTO::as_manifest()` → `pesto_ensemble_manifest`.
- Persistence: `PESTO::write_manifest()` /
  `PESTO::read_manifest()` / `PESTO::verify_manifest()`.
- Distributional verdict: this function.
- (Future) proxymix density-ratio backend plug-in: tracked under §C1
  of the roadmap; will become a fourth propensity-model option once
  it lands.

## Reproducibility

```{r}
sessionInfo()
```
