---
title: "Is the calibrated model right about more than the mean?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Is the calibrated model right about more than the mean?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. Four posterior-predictive checks at 199 to 299 permutations on
     ensembles of 300 draws against 30 held-out observations. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A modeller has just finished a calibration and wants to know whether to
believe it. *The ensemble smoother converged, the posterior-predictive
mean sits within a few per cent of the held-out season, and the root mean
squared error is the best I have had. But the decision I am supporting is
about the bad years, so what I need to know is whether the predictive
distribution is right, not whether its centre is. Is the held-out year a
plausible draw from what the calibrated model predicts?* This vignette
answers that on three synthetic ensembles: one calibrated correctly, one
over-confident, and one displaced.

## What

`mmd_ppc()` is a posterior-predictive check built on a two-sample kernel
test. It compares the posterior-predictive draws against a held-out
sample with the Maximum Mean Discrepancy and scores the observed
discrepancy against a permutation null, so a difference in spread,
shape, modality or tail weight counts exactly as a difference in
location does. That is the whole point: the standard check in an
ensemble-smoother workflow is the root mean squared error of the
posterior-predictive mean, and a distribution can match in mean while
failing badly everywhere else.

Alongside the p-value it reports `surprise_bits`, the Shannon
information of the permutation p-value: 0 bits at a p-value of 1, and
capped at the base-two logarithm of one more than the permutation count.
It is an intuitive scalar summary of the same verdict, not a posterior
probability and not the Bayesian-surprise divergence.

`pesto_ensemble()` is the lightweight constructor: posterior draws,
observations, and optional metadata. `mmd_ppc()` also consumes a
`PESTO::pesto_ensemble_manifest()`, the C2 manifest contract PESTO's own
emitter produces, through a dedicated method, and `outputs` restricts the
check to named observation columns on that path.

## Do

### A held-out sample

Thirty synthetic observations of a bivariate outcome, standing in for a
year or a paddock kept out of the calibration.

```{r observations}
set.seed(1L)
n_held <- 30L
truth_mean <- c(yield = 4.5, biomass = 12.0)
truth_sd <- c(yield = 0.6, biomass = 1.5)
observed <- cbind(
  yield = rnorm(n_held, truth_mean[["yield"]], truth_sd[["yield"]]),
  biomass = rnorm(n_held, truth_mean[["biomass"]],
                  truth_sd[["biomass"]])
)
```

### A calibrated ensemble

```{r calibrated}
n_draws <- 300L
post_good <- cbind(
  yield = rnorm(n_draws, truth_mean[["yield"]], truth_sd[["yield"]]),
  biomass = rnorm(n_draws, truth_mean[["biomass"]],
                  truth_sd[["biomass"]])
)
ens_good <- pesto_ensemble(
  posterior = post_good, observed = observed,
  metadata = list(holdout_year = 2018L, ies_iters = 6L)
)
ens_good

fit_good <- mmd_ppc(ens_good, n_permutations = 299L, seed = 1L)
fit_good
```

### An over-confident ensemble

The most common ensemble-smoother pathology: posterior draws clustered
too tightly around the posterior mean. Here the spread is a third of what
it should be, and the centre is exactly right.

```{r narrow}
post_narrow <- cbind(
  yield = rnorm(n_draws, truth_mean[["yield"]],
                truth_sd[["yield"]] / 3),
  biomass = rnorm(n_draws, truth_mean[["biomass"]],
                  truth_sd[["biomass"]] / 3)
)
ens_narrow <- pesto_ensemble(post_narrow, observed,
                             metadata = list(scenario = "narrow"))
fit_narrow <- mmd_ppc(ens_narrow, n_permutations = 299L, seed = 1L)
fit_narrow
```

### A displaced ensemble

```{r shifted}
post_shifted <- cbind(
  yield = rnorm(n_draws, truth_mean[["yield"]] + 0.8,
                truth_sd[["yield"]]),
  biomass = rnorm(n_draws, truth_mean[["biomass"]] - 1.2,
                  truth_sd[["biomass"]])
)
ens_shifted <- pesto_ensemble(post_shifted, observed)
fit_shifted <- mmd_ppc(ens_shifted, n_permutations = 299L, seed = 1L)
fit_shifted
```

```{r summary-table, echo = FALSE}
fits <- list(calibrated = fit_good, over_confident = fit_narrow,
             displaced = fit_shifted)
ppc_tab <- data.frame(
  ensemble = c("calibrated", "over-confident", "displaced"),
  rmse_of_mean = vapply(
    list(post_good, post_narrow, post_shifted),
    function(p1) {
      sqrt(mean((colMeans(p1) - colMeans(observed))^2))
    }, numeric(1L)
  ),
  statistic = vapply(fits, function(z) z$statistic, numeric(1L)),
  p_value = vapply(fits, function(z) z$p_value, numeric(1L)),
  surprise_bits = vapply(fits, function(z) z$surprise_bits, numeric(1L)),
  reject = vapply(fits, function(z) z$reject, logical(1L)),
  row.names = NULL, stringsAsFactors = FALSE
)
surprise_at_alpha <- -log2(fit_good$alpha)
surprise_cap <- log2(fit_good$n_permutations + 1)
knitr::kable(
  ppc_tab, digits = c(0L, 3L, 4L, 4L, 2L, 0L),
  col.names = c("Ensemble", "RMSE of the mean", "MMD statistic",
                "p-value", "Surprise (bits)",
                paste0("Reject at ", fit_good$alpha)),
  caption = paste0(
    "The three ensembles against the same ", n_held,
    " held-out observations. Surprise is capped at ",
    round(surprise_cap, 2), " bits by the ", fit_good$n_permutations,
    "-permutation budget; ", round(surprise_at_alpha, 2),
    " bits corresponds to a p-value of ", fit_good$alpha, "."
  )
)
```

### The package's own diagnostic

```{r plot-null, fig.cap = "Permutation null distribution of the MMD statistic for the over-confident ensemble, with the observed statistic marked against the held-out sample."}
plot(fit_narrow)
```

### The figure: what each ensemble got wrong

```{r fig-predictive, eval = has_ggplot2, fig.height = 5, fig.cap = "Posterior-predictive distributions of the three ensembles against the held-out observations, shown as tick marks along the axis, one panel per output. The calibrated curve covers the observations; the over-confident curve is centred correctly and far too narrow to have produced them; the displaced curve has the right width in the wrong place. The root mean squared error of the mean cannot separate the first two."}
pred_df <- do.call(rbind, lapply(
  list(calibrated = post_good, `over-confident` = post_narrow,
       displaced = post_shifted),
  function(p1) {
    data.frame(
      value = as.numeric(p1),
      output = rep(colnames(p1), each = nrow(p1)),
      stringsAsFactors = FALSE
    )
  }
))
pred_df$ensemble <- rep(
  c("calibrated", "over-confident", "displaced"),
  each = length(post_good)
)
obs_df <- data.frame(
  value = as.numeric(observed),
  output = rep(colnames(observed), each = nrow(observed)),
  stringsAsFactors = FALSE
)
ggplot2::ggplot(pred_df, ggplot2::aes(x = value, colour = ensemble)) +
  ggplot2::geom_density(linewidth = 0.9) +
  ggplot2::geom_rug(
    data = obs_df, ggplot2::aes(x = value), inherit.aes = FALSE,
    colour = "grey30", alpha = 0.8
  ) +
  ggplot2::facet_wrap(~output, scales = "free") +
  ggplot2::scale_colour_viridis_d(name = "ensemble", end = 0.85) +
  ggplot2::labs(
    x = "output value", y = "posterior-predictive density",
    title = "Three ensembles against the same held-out observations"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig-predictive-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the",
    "posterior-predictive figure is skipped.\n")
```

### The same check on a PESTO manifest

PESTO's own emitter produces a `pesto_ensemble_manifest`, and `mmd_ppc()`
consumes it through a dedicated method. Held-out observations are
supplied explicitly, because the manifest's `obs_target` slot is the
single point the posterior was fitted to and is not a valid two-sample
comparator on its own.

```{r manifest-path}
n_manifest <- 80L
n_cols <- 3L
post_manifest <- matrix(rnorm(n_manifest * n_cols), n_manifest, n_cols)
colnames(post_manifest) <- paste0("o", seq_len(n_cols))
manifest <- PESTO::pesto_ensemble_manifest(
  run_id = "vignette_demo",
  params = data.frame(
    real_name = paste0("r", seq_len(n_manifest)),
    p1 = rnorm(n_manifest), p2 = rnorm(n_manifest),
    check.names = FALSE
  ),
  outputs = data.frame(
    real_name = paste0("r", seq_len(n_manifest)), post_manifest,
    check.names = FALSE
  ),
  weights = setNames(rep(1, n_cols), colnames(post_manifest)),
  obs_target = setNames(rnorm(n_cols), colnames(post_manifest)),
  data_hash = "sha256:vignette_demo",
  pesto_version = as.character(packageVersion("PESTO")),
  timestamp = Sys.time(),
  method = "ies_callback",
  noptmax = 1L,
  lambda_schedule = 1
)

held_out <- matrix(rnorm(20L * n_cols), 20L, n_cols)
colnames(held_out) <- colnames(post_manifest)
fit_manifest <- mmd_ppc(manifest, observed = held_out,
                        n_permutations = 199L, seed = 1L)
fit_manifest$p_value
```

## Read

The calibrated ensemble is not falsified. Its MMD statistic is
`r signif(fit_good$statistic, 3)` with a p-value of
`r round(fit_good$p_value, 3)` and
`r round(fit_good$surprise_bits, 2)` bits of surprise, well below the
`r round(surprise_at_alpha, 2)` bits that correspond to a p-value of
`r fit_good$alpha`: the held-out observations are an unremarkable draw
from what the model predicts.

The over-confident ensemble is where the check earns its place. Its
posterior mean is correct -- the root mean squared error of the mean is
`r round(ppc_tab$rmse_of_mean[ppc_tab$ensemble == "over-confident"], 3)`
against `r round(ppc_tab$rmse_of_mean[
  ppc_tab$ensemble == "calibrated"], 3)` for the calibrated ensemble, so
the standard diagnostic ranks the two as
`r ifelse(abs(diff(ppc_tab$rmse_of_mean[1:2])) < 0.2,
  "nearly equivalent", "clearly different")` -- and `mmd_ppc()` returns a
statistic of
`r signif(fit_narrow$statistic, 3)` with a p-value of
`r round(fit_narrow$p_value, 4)` and
`r round(fit_narrow$surprise_bits, 2)` bits of surprise, which is
`r if (fit_narrow$reject) "a rejection" else "not a rejection"`. The
figure shows why: the narrow curve is a spike where the observations are
a spread.

The displaced ensemble is rejected too, statistic
`r signif(fit_shifted$statistic, 3)`, p-value
`r round(fit_shifted$p_value, 4)`,
`r round(fit_shifted$surprise_bits, 2)` bits, and this one the root mean
squared error of the mean would also have caught, at
`r round(ppc_tab$rmse_of_mean[ppc_tab$ensemble == "displaced"], 3)`. The
value of the distributional check is not that it finds displacement; it
is that it finds the over-confidence that displacement-based diagnostics
cannot.

The manifest path reaches the same machinery from PESTO's own object. On
the synthetic manifest above, where the posterior draws and the held-out
sample are both standard normal and no failure was built in, the check
returns a p-value of `r round(fit_manifest$p_value, 3)`.

## Limits

All three ensembles are synthetic, and the failures were built in so that
the verdicts could be read against a known answer. Distributional power
is set by both the ensemble size and the held-out sample size; with
`r n_held` held-out observations, only a fairly gross miscalibration is
detectable, and a check that fails to reject is not evidence of
calibration. The surprise measure is a monotone transformation of the
p-value and carries no information the p-value does not: it inherits the
permutation cap of `r round(surprise_cap, 2)` bits at
`r fit_good$n_permutations` permutations, so it saturates instead of
growing once the p-value hits its floor. This is a check against
*held-out* data; run against the data the model was calibrated on, it
tests goodness of fit rather than predictive adequacy and will be
optimistic. Finally the check is silent about *why* an ensemble fails: it
reports that the predictive distribution and the observations differ, and
the figure, not the statistic, is what says whether the failure is in
location, spread or shape.

## Notes on practice

**Ensemble and sample size.** An ensemble in the low hundreds is usually
enough at agricultural scale; twenty held-out observations is a practical
floor below which the test has little to work with.

**Permutations.** At `r fit_good$n_permutations` permutations the
smallest p-value is
`r signif(1 / (fit_good$n_permutations + 1), 3)` and the largest surprise
is `r round(surprise_cap, 2)` bits. Raise the budget when working near
the floor.

**Multi-dimensional outputs.** `mmd_ppc()` handles more than one output
natively through the kernel; the default RBF bandwidth is set by the
median heuristic over the pooled posterior and observed sample, so the
output columns should be on comparable scales or standardised first.

## What to read next

*Are these draws calibrated, and do the engines agree?* asks the
one-sample version of the same question, comparing draws against a target
density rather than against held-out data, and adds the concordance test
across several sources. *Did the management change shift the simulated
distribution?* compares two calibrated ensembles against each other
instead of against observations. *Getting started with kernR* introduces
the MMD two-sample test this check is built on.

## References

- Gelman, A., Meng, X.-L., & Stern, H. (1996). Posterior predictive
  assessment of model fitness via realized discrepancies. *Statistica
  Sinica*, 6(4), 733-760.
- Gretton, A., Borgwardt, K. M., Rasch, M. J., Scholkopf, B., & Smola, A.
  (2012). A kernel two-sample test. *Journal of Machine Learning
  Research*, 13, 723-773.

## Reproduce

Seed `1` for the held-out sample and all three ensembles;
`seed = 1L` passed to every check so the permutation draws are fixed.
Package versions follow.

```{r session-info}
sessionInfo()
```
