---
title: "Getting started with kernR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with kernR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~3 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. The permutation nulls here are small (199 to 999 draws on a few
     hundred points), so the time is dominated by knitr overhead. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

An analyst has two samples and two variables, and the question that
matters is not about means. *My control and treated yields have the same
average, so the t-test says nothing happened -- but the two histograms do
not look alike. And this soil covariate has a Pearson correlation of
roughly zero with yield, which I do not believe for a moment. What is the
smallest thing I can write that tests the distributions themselves, and
tests dependence without assuming it is linear?* This vignette answers
both halves on two synthetic toys, and shows what the answer looks like
when it comes back.

## What

Four objects and verbs carry this vignette. `kernel_spec()` declares the
kernel -- the similarity function that turns raw observations into a Gram
matrix -- with a family (`"rbf"`, `"matern"`, `"linear"`, `"polynomial"`)
and a bandwidth, which is resolved from the data by the median heuristic
unless a number is given. `kernel_matrix()` evaluates that specification
on a data matrix and returns the Gram matrix itself, which is the only
object the tests below ever see of the raw data.

`mmd_test()` is the two-sample test: it compares the kernel mean
embeddings of two samples through the Maximum Mean Discrepancy, so it
sees any difference the kernel can represent, not just a shift in
location. `hsic_test()` is the independence test: it computes the
Hilbert-Schmidt Independence Criterion between two variables, which is
zero if and only if they are independent under a characteristic kernel,
so it detects curved dependence that a correlation coefficient reports as
nothing. Both return a `kernel_test_result` carrying the observed
statistic, the permutation null it was scored against, and the p-value
that follows, and both take a `seed` so the permutation draw is
reproducible.

## Do

### The kernel specification

```{r kernel-spec}
k_default <- kernel_spec()
k_default

k_fixed <- kernel_spec("rbf", bandwidth = 1.5)
k_fixed

k_linear <- kernel_spec("linear")
k_linear
```

### The Gram matrix

A synthetic toy: 100 draws from a standard bivariate normal, used only to
show the shape of the object the tests consume.

```{r kernel-matrix}
set.seed(42L)
x_toy <- matrix(rnorm(200L), 100L, 2L)
gram <- kernel_matrix(x_toy)
dim(gram)
round(gram[1:4, 1:4], 3L)
```

### Two samples, same distribution

Two synthetic samples of 100 points each, both standard bivariate normal.

```{r mmd-same}
set.seed(123L)
x_a <- matrix(rnorm(200L), 100L, 2L)
y_a <- matrix(rnorm(200L), 100L, 2L)
mmd_null <- mmd_test(x_a, y_a, n_permutations = 999L, seed = 1L)
mmd_null
```

### Two samples, one shifted

The same comparison with the second sample's mean moved by 0.5 in both
coordinates.

```{r mmd-different}
y_shift <- matrix(rnorm(200L, mean = 0.5), 100L, 2L)
mmd_shift <- mmd_test(x_a, y_shift, n_permutations = 999L, seed = 1L)
mmd_shift
```

### Dependence a correlation cannot see

A synthetic toy in which the outcome is the square of the covariate plus
noise: the dependence is exact, and the linear correlation is close to
zero because the relationship is symmetric about the origin.

```{r hsic-nonlinear}
set.seed(456L)
n_hsic <- 300L
x_q <- rnorm(n_hsic)
y_q <- x_q^2 + rnorm(n_hsic, sd = 0.3)
pearson_r <- cor(x_q, y_q)
round(pearson_r, 3L)

hsic_quad <- hsic_test(x_q, y_q, n_permutations = 999L, seed = 1L)
hsic_quad
```

### The figure: where the observed statistic sits

```{r fig-null, eval = has_ggplot2, fig.cap = "Permutation null distribution of the HSIC statistic on the quadratic toy (999 label permutations), with the observed statistic marked. The observed value sits far outside the range the null ever reaches, which is what a p-value at the permutation floor means."}
null_df <- data.frame(statistic = hsic_quad$null_distribution)
ggplot2::ggplot(null_df, ggplot2::aes(x = statistic)) +
  ggplot2::geom_histogram(bins = 40L, fill = "#440154", colour = NA) +
  ggplot2::geom_vline(
    xintercept = hsic_quad$statistic, colour = "#FDE725", linewidth = 1
  ) +
  ggplot2::annotate(
    "text", x = hsic_quad$statistic, y = Inf, hjust = 1.1, vjust = 2,
    label = "observed", colour = "#FDE725"
  ) +
  ggplot2::labs(
    x = "HSIC statistic (permutation draws, unitless)",
    y = "permutation draws",
    title = "HSIC permutation null against the observed statistic"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-null-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the permutation-null",
    "figure is skipped.\n")
```

### The three verdicts side by side

```{r verdict-table, echo = FALSE}
verdicts <- data.frame(
  question = c(
    "Two samples, same distribution",
    "Two samples, mean shifted by 0.5",
    "Quadratic dependence, correlation near zero"
  ),
  test = c("mmd_test()", "mmd_test()", "hsic_test()"),
  statistic = c(mmd_null$statistic, mmd_shift$statistic,
                hsic_quad$statistic),
  p_value = c(mmd_null$p_value, mmd_shift$p_value, hsic_quad$p_value),
  stringsAsFactors = FALSE
)
knitr::kable(
  verdicts, digits = 4L,
  col.names = c("Question", "Test", "Statistic", "p-value"),
  caption = paste0(
    "The three verdicts in this vignette, each scored against a ",
    mmd_null$n_permutations, "-draw permutation null."
  )
)
```

## Read

The two samples drawn from the same distribution give an MMD statistic of
`r signif(mmd_null$statistic, 3)` and a p-value of
`r round(mmd_null$p_value, 3)`: the observed discrepancy is an ordinary
member of its own permutation null, so there is no evidence the two
samples differ. Shifting the second sample's mean by 0.5 raises the
statistic to `r signif(mmd_shift$statistic, 3)` and drops the p-value to
`r round(mmd_shift$p_value, 4)`, the smallest value
`r mmd_shift$n_permutations` permutations can express
(`1 / (B + 1)` = `r round(1 / (mmd_shift$n_permutations + 1), 4)`), so the
test has separated the two samples as decisively as this permutation
budget allows.

The quadratic toy is where the kernel earns its keep. The Pearson
correlation between covariate and outcome is
`r round(pearson_r, 3)`, near enough to zero that a linear screen would
discard the covariate. HSIC returns
`r signif(hsic_quad$statistic, 3)` against a null whose largest of
`r hsic_quad$n_permutations` draws is only
`r signif(max(hsic_quad$null_distribution), 3)`, giving a p-value of
`r round(hsic_quad$p_value, 4)`: the dependence is real, strong, and
invisible to correlation. The figure shows the separation directly -- the
whole null sits in a narrow band near zero and the observed statistic is
nowhere near it.

## Limits

Both toys here are synthetic and generously sized for the effects they
carry, so they show that the tests work rather than how they behave at
the margin. A permutation p-value cannot go below
`1 / (n_permutations + 1)`, so a small p-value in this vignette reports
the budget as much as the evidence; raise `n_permutations` before reading
a small p-value as a magnitude. Neither test here adjusts for anything:
`mmd_test()` and `hsic_test()` answer questions about association and
distributional difference, and an association is not a causal effect
whenever a common cause is in play. The tests that do adjust are the
subject of the next two vignettes. Finally, both tests inherit the
kernel's blind spots: a linear kernel cannot see the quadratic dependence
above, and the median-heuristic bandwidth is a default, not a tuned
choice.

## What to read next

*Does the treatment cause the outcome, or do they share a cause?*
adds backdoor adjustment, so the association measured here becomes a
causal one. *A treatment that changes the spread and not the mean* runs
the doubly robust distributional tests for treatment effects that a
mean-based method cannot see. *When plots sit inside farms* handles the
clustered designs in which the permutation null used above is no longer
valid.

## Reproduce

Seeds `42` (Gram matrix toy), `123` (two-sample toys) and `456`
(quadratic toy) for data generation, and `seed = 1L` passed to every test
so the permutation draws are fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
