---
title: "Multi-site trials and the permutation null"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-site trials and the permutation null}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. Four bd-HSIC tests at 299 permutations on 240 observations in 8
     sites, each with its own density-ratio fit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A trial statistician is looking at a panel rather than a field. *The
residue-retention trial ran at eight sites and the sites differ in ways
the covariates do not capture. When the causal test shuffles the outcome
to build its null, it shuffles across sites, so the null absorbs the
between-site variation as though it were signal I could have produced by
chance. That inflates the null, or deflates it, and either way the
p-value is answering a question about a design I do not have. Can I make
the null respect the design?* This vignette answers with the
`cluster_id` argument of the backdoor-adjusted test, on a synthetic
eight-site panel with a real effect and on the same panel with the effect
removed.

## What

`bd_hsic_test()` builds its permutation null by shuffling the outcome
within clusters of observations that are exchangeable. Which
observations those are is the whole question, and the function offers
three answers.

The default, `permutation = "auto"` with no `cluster_id`, clusters on
similarity in the estimated conditional density of treatment given the
confounders, following Hu, Sejdinovic and Evans (2024); it is the right
choice when the design has no natural grouping, and the vignette *Does
the treatment cause the outcome, or do they share a cause?* uses it
throughout. `permutation = "naive"` shuffles across the whole sample,
which assumes every observation is exchangeable with every other.
Supplying `cluster_id` activates within-cluster permutation on the
design's own groups, so site-level effects are held fixed in the null
rather than being shuffled into it.

Supplying `cluster_id` also turns on a stratified breakdown:
`per_cluster_statistic` carries the weighted bd-HSIC computed inside each
cluster, which is where a single site driving the whole verdict becomes
visible.

## Do

### A synthetic eight-site panel

Eight sites, thirty plots each. Each site carries a random effect on
yield with standard deviation 1, twice the residual noise. Rainfall and
soil nitrogen are the measured confounders; the management variable
depends on rainfall and on the site effect; and yield carries a genuine
causal effect of 0.6 per unit of management.

```{r design}
set.seed(101L)
n_sites <- 8L
n_per_site <- 30L
n_obs <- n_sites * n_per_site
site <- factor(paste0("site", rep(seq_len(n_sites), each = n_per_site)))
site_effect <- rnorm(n_sites, sd = 1.0)[as.integer(site)]

z_conf <- cbind(
  rainfall = rnorm(n_obs),
  soil_n = rnorm(n_obs)
)
x_manage <- 0.4 * z_conf[, "rainfall"] + 0.3 * site_effect + rnorm(n_obs)
y_yield <- 0.6 * x_manage + 0.5 * z_conf[, "soil_n"] + site_effect +
  rnorm(n_obs, sd = 0.5)
```

### The same data, two permutation schemes

```{r compare-schemes}
fit_naive <- bd_hsic_test(
  x_manage, y_yield, z_conf, permutation = "naive",
  n_permutations = 299L, seed = 1L
)
fit_site <- bd_hsic_test(
  x_manage, y_yield, z_conf, cluster_id = site,
  n_permutations = 299L, seed = 1L
)
c(fit_naive$permutation_scheme, fit_site$permutation_scheme)
```

### The same panel with the effect removed

Identical sites, plots, confounders and management assignment; the
management term is dropped from the yield, so the only structure left is
the site effect and the confounders.

```{r null-case}
y_null <- 0.5 * z_conf[, "soil_n"] + site_effect + rnorm(n_obs, sd = 0.5)
fit_naive_null <- bd_hsic_test(
  x_manage, y_null, z_conf, permutation = "naive",
  n_permutations = 299L, seed = 1L
)
fit_site_null <- bd_hsic_test(
  x_manage, y_null, z_conf, cluster_id = site,
  n_permutations = 299L, seed = 1L
)
```

```{r scheme-table, echo = FALSE}
scheme_tab <- data.frame(
  panel = c("real effect", "real effect", "no effect", "no effect"),
  scheme = c(fit_naive$permutation_scheme, fit_site$permutation_scheme,
             fit_naive_null$permutation_scheme,
             fit_site_null$permutation_scheme),
  statistic = c(fit_naive$statistic, fit_site$statistic,
                fit_naive_null$statistic, fit_site_null$statistic),
  null_mean = c(mean(fit_naive$null_distribution),
                mean(fit_site$null_distribution),
                mean(fit_naive_null$null_distribution),
                mean(fit_site_null$null_distribution)),
  null_sd = c(sd(fit_naive$null_distribution),
              sd(fit_site$null_distribution),
              sd(fit_naive_null$null_distribution),
              sd(fit_site_null$null_distribution)),
  p_value = c(fit_naive$p_value, fit_site$p_value,
              fit_naive_null$p_value, fit_site_null$p_value),
  stringsAsFactors = FALSE
)
knitr::kable(
  scheme_tab, digits = c(0L, 0L, 5L, 5L, 5L, 4L), row.names = FALSE,
  col.names = c("Panel", "Permutation scheme", "Statistic",
                "Null mean", "Null SD", "p-value"),
  caption = paste0(
    "The two permutation schemes on the same eight-site panel, with and ",
    "without a causal effect, each against ", fit_site$n_permutations,
    " permutations."
  )
)
```

### The stratified breakdown

```{r per-cluster}
round(fit_site$per_cluster_statistic, 4L)
```

### The figure: how the two nulls differ

```{r fig-nulls, eval = has_ggplot2, fig.height = 5, fig.cap = "Permutation null distributions under the two schemes, on the panel with a real effect and on the panel without one, with the observed statistic marked in each. Shuffling within sites holds the site effects fixed, which changes the shape of the null the observed statistic is scored against; the naive scheme shuffles those effects into the null."}
null_df <- rbind(
  data.frame(statistic = fit_naive$null_distribution,
             scheme = "naive", panel = "real effect"),
  data.frame(statistic = fit_site$null_distribution,
             scheme = "within site", panel = "real effect"),
  data.frame(statistic = fit_naive_null$null_distribution,
             scheme = "naive", panel = "no effect"),
  data.frame(statistic = fit_site_null$null_distribution,
             scheme = "within site", panel = "no effect")
)
obs_df <- data.frame(
  scheme = c("naive", "within site", "naive", "within site"),
  panel = c("real effect", "real effect", "no effect", "no effect"),
  statistic = c(fit_naive$statistic, fit_site$statistic,
                fit_naive_null$statistic, fit_site_null$statistic)
)
ggplot2::ggplot(null_df, ggplot2::aes(x = statistic, fill = scheme)) +
  ggplot2::geom_histogram(bins = 40L, alpha = 0.65,
                          position = "identity", colour = NA) +
  ggplot2::geom_vline(
    data = obs_df,
    ggplot2::aes(xintercept = statistic, colour = scheme), linewidth = 1
  ) +
  ggplot2::facet_wrap(~panel, scales = "free", ncol = 1L) +
  ggplot2::scale_fill_viridis_d(name = "scheme", end = 0.85) +
  ggplot2::scale_colour_viridis_d(name = "scheme", end = 0.85) +
  ggplot2::labs(
    x = "weighted bd-HSIC statistic (unitless)",
    y = "permutation draws",
    title = "Two permutation schemes, two panels"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig-nulls-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the permutation-null",
    "figure is skipped.\n")
```

```{r fig-clusters, eval = has_ggplot2, fig.cap = "Weighted bd-HSIC computed inside each site on the panel with a real effect, with the mean across sites marked. A site whose stratified statistic dwarfs the rest is a candidate for follow-up, either as genuine site-specific effect modification or as a data-quality problem."}
cluster_df <- data.frame(
  site = names(fit_site$per_cluster_statistic),
  statistic = as.numeric(fit_site$per_cluster_statistic),
  stringsAsFactors = FALSE
)
ggplot2::ggplot(
  cluster_df, ggplot2::aes(x = site, y = statistic)
) +
  ggplot2::geom_col(fill = "#3b528b") +
  ggplot2::geom_hline(
    yintercept = mean(cluster_df$statistic, na.rm = TRUE),
    colour = "#FDE725", linewidth = 0.9, linetype = "dashed"
  ) +
  ggplot2::labs(
    x = "site", y = "weighted bd-HSIC within site (unitless)",
    title = "Stratified contributions, site by site"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45,
                                                     hjust = 1))
```

```{r fig-clusters-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the per-site figure is",
    "skipped.\n")
```

## Read

On the panel with a real effect both schemes reject: the naive scheme
gives a p-value of `r round(fit_naive$p_value, 4)` and the within-site
scheme `r round(fit_site$p_value, 4)`, against the same observed
statistic of `r signif(fit_site$statistic, 3)`, because the statistic
does not depend on the permutation scheme. What differs is the null it is
scored against, and the figure shows that directly: the naive null has a
standard deviation of
`r formatC(scheme_tab$null_sd[scheme_tab$panel == "real effect" &
  scheme_tab$scheme == "naive"], format = "f", digits = 5L)` against
`r formatC(scheme_tab$null_sd[scheme_tab$panel == "real effect" &
  scheme_tab$scheme == "within_cluster"], format = "f", digits = 5L)`
for the within-site null, and a mean of
`r formatC(scheme_tab$null_mean[scheme_tab$panel == "real effect" &
  scheme_tab$scheme == "naive"], format = "f", digits = 5L)` against
`r formatC(scheme_tab$null_mean[scheme_tab$panel == "real effect" &
  scheme_tab$scheme == "within_cluster"], format = "f", digits = 5L)`:
holding the site effects fixed shifts the whole null upward and widens
it.

On the panel with the effect removed the two schemes give p-values of
`r round(fit_naive_null$p_value, 3)` (naive) and
`r round(fit_site_null$p_value, 3)` (within site) against an observed
statistic of `r signif(fit_site_null$statistic, 3)`. Both
`r if (max(fit_naive_null$p_value, fit_site_null$p_value) > 0.05 &&
  min(fit_naive_null$p_value, fit_site_null$p_value) > 0.05)
  "decline to reject, which is the right answer on a panel built with no
  effect" else
  "should be read against the table above rather than summarised here"`.
One draw is not a type-I error rate: what this pair of runs shows is the
mechanism, that the two schemes construct different nulls from identical
data, and not a measured false-positive rate for either. Establishing
that would need the whole panel regenerated many times over, which is a
simulation study and not a vignette.

The stratified breakdown is the diagnostic worth keeping. Across the
`r sum(!is.na(fit_site$per_cluster_statistic))` sites with enough test
observations to score, the within-site statistic runs from
`r signif(min(fit_site$per_cluster_statistic, na.rm = TRUE), 3)` to
`r signif(max(fit_site$per_cluster_statistic, na.rm = TRUE), 3)`, a
factor of
`r round(max(fit_site$per_cluster_statistic, na.rm = TRUE) /
   max(min(fit_site$per_cluster_statistic, na.rm = TRUE), 1e-8))`
between the quietest site and the loudest. The second figure shows the
spread. A panel in which one site towers over the rest is telling you
something the pooled verdict does not: either the effect is real only
there, or that site's data need looking at.

## Limits

This is a synthetic panel with balanced sites and one level of nesting,
and every result above is a single run. The within-site scheme needs at
least two test observations per cluster, and the test split takes only
part of the sample, so small clusters are reported as `NA` in
`per_cluster_statistic` and contribute the identity permutation to the
null; a design of paddocks with three to five plots each has little
within-cluster permutation power and is usually better pooled to the
site. The per-cluster statistics are reported on the raw scale and are
not weighted by cluster size, so an unbalanced panel needs a size-weighted
summary computed outside the function before the sites are compared. The
choice of clustering level is an assumption about exchangeability, not a
fact the data can supply: within-site permutation assumes plots inside a
site are exchangeable, which a systematic within-site layout would break.
Finally, everything the vignette *Does the treatment cause the outcome,
or do they share a cause?* says about the density-ratio step applies
unchanged here, including the effective-sample-size floor.

## Choosing the scheme

Where the observations are genuinely independent and the design has no
grouping, leave `permutation` at `"auto"` and supply no `cluster_id`: the
propensity-similarity clustering is the intended default. Where the
design has sites, seasons or paddocks that the covariates do not fully
explain, supply `cluster_id` at the level you are willing to call
exchangeable, and take the within-cluster scheme it activates.
`permutation = "naive"` is for the case where within-cluster
independence is plausible on other grounds and maximum power matters
more, which is rare in a field trial.

## Notes on practice

**Cluster size.** Within-cluster permutation needs at least two test
observations per cluster, where the test split is one minus
`split_ratio` of the sample. Clusters below that appear as `NA` in the
stratified breakdown.

**Many small clusters.** Permutation power falls with cluster size.
Pooling to a coarser level, site rather than paddock, is usually the
better trade.

**Imbalanced designs.** Combine the per-cluster statistics with the
cluster sizes yourself if a size-weighted summary is what the report
needs.

## What to read next

*Does the treatment cause the outcome, or do they share a cause?*
introduces the backdoor-adjusted test and its density-ratio step, and is
the place to start if bd-HSIC is new. *When plots sit inside farms*
applies the same clustering idea to the doubly robust distributional
tests, with a within-cluster and between-cluster decomposition this
vignette does not compute. *Choosing a density-ratio backend* covers the
adjustment step whose weights also determine the default clustering.

## References

- Hu, R., Sejdinovic, D., & Evans, R. J. (2024). A kernel test for causal
  association via noise contrastive backdoor adjustment. *Journal of
  Machine Learning Research*, 25(160), 1-56.

## Reproduce

Seed `101` for the synthetic panel and for the null variant;
`seed = 1L` passed to every test so the density-ratio fit and the
permutation draws are fixed. Package versions follow.

```{r session-info}
sessionInfo()
```
