---
title: "Which parameters change the shape of the output distribution?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Which parameters change the shape of the output distribution?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~21 s under rmarkdown::render(); macOS aarch64, R 4.5.2,
     one core. Five sensitivity scans on designs of 400 to 600 rows, two of
     them with total-order indices and one with a 100-draw pair bootstrap;
     the 600-row first-order scan and the conditional-permutation null
     dominate. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A risk modeller is ranking the parameters of a crop model and does not
trust the ranking. *The variance budget says one parameter dominates and
the rest are negligible. But two of those negligible parameters are the
ones that decide whether a bad season is merely poor or a write-off: one
fattens the lower tail and one skews the whole distribution, neither of
them changes the variance much. Is there a sensitivity index that ranks a
parameter by what it does to the distribution rather than by how much
variance it explains?* This vignette answers with the HSIC-based
sensitivity index, on a synthetic simulator whose parameters were built
to separate the two questions, and then adds total-order indices for the
parameters that act only in combination.

## What

A variance-based Sobol index decomposes the output variance into
contributions from each input, which makes it blind by construction to
any effect that leaves the variance where it was. The HSIC-Sensitivity
Index of Da Veiga (2015) replaces the variance with a kernel dependence
measure. The normalised first-order index is

\[
S_j^{\mathrm{HSIC}}
  = \frac{\mathrm{HSIC}(X_j, Y)}
         {\sqrt{\mathrm{HSIC}(X_j, X_j) \cdot \mathrm{HSIC}(Y, Y)}}
  \in [0, 1],
\]

bounded like a Sobol index, comparable across parameters, and sensitive
to any change in the conditional distribution of the output, including
the mean-preserving ones.

`hsic_sensitivity()` computes it for every parameter against every
output, with a permutation p-value per pair and a Benjamini-Hochberg
adjustment across the grid. `total_order = TRUE` adds the total-order
index in Da Veiga's complement form,

\[
T_j^{\mathrm{HSIC}}
  = 1 - \frac{\mathrm{HSIC}(X_{\sim j}, Y)}
              {\sqrt{\mathrm{HSIC}(X_{\sim j}, X_{\sim j}) \cdot
                     \mathrm{HSIC}(Y, Y)}},
\]

which measures the contribution of $X_j$ directly plus through every
interaction it takes part in. `total_order_ci = TRUE` attaches a
pair-bootstrap percentile interval to each total-order index, and
`total_order_test = "cond_perm"` attaches a conditional-permutation
significance test to it. `lhs_design()` builds the design; `plot()` and
`as.data.frame()` render the result.

## Do

### A simulator built to separate the two questions

A synthetic stub with four parameters, each carrying one kind of effect.
`mean_effect` shifts the mean, which a variance budget also sees.
`var_effect` changes the spread and nothing else. `tail_effect` injects a
heavy-tailed component, changing the shape without moving the centre.
`inert` does nothing.

```{r stub}
stub_apsim <- function(theta) {
  n_row <- nrow(theta)
  yield <- 1.5 * theta[, "mean_effect"] +
    rnorm(n_row, sd = exp(theta[, "var_effect"])) +
    theta[, "tail_effect"] * rt(n_row, df = 3) * 0.4 +
    rnorm(n_row, sd = 0.05)
  matrix(yield, ncol = 1L, dimnames = list(NULL, "yield"))
}

set.seed(1L)
bounds <- rbind(
  mean_effect = c(0.0, 1.0),
  var_effect  = c(-0.5, 0.5),
  tail_effect = c(0.0, 1.0),
  inert       = c(0.0, 1.0)
)
design <- lhs_design(n = 600L, bounds = bounds, seed = 11L)
yield <- stub_apsim(design)
```

### The first-order scan

```{r scan}
fit <- hsic_sensitivity(
  theta = design, y = yield, n_permutations = 199L, seed = 11L
)
fit
```

```{r scan-table, echo = FALSE}
alpha_level <- 0.05
scan_df <- as.data.frame(fit)
scan_df <- scan_df[order(-scan_df$index), ]
knitr::kable(
  scan_df, digits = c(0L, 0L, 4L, 5L, 4L, 4L), row.names = FALSE,
  col.names = c("Parameter", "Output", "Index", "HSIC statistic",
                "p-value", "Adjusted p-value"),
  caption = paste0(
    "First-order HSIC-Sensitivity Indices on the ", nrow(design),
    "-row design, ranked, with p-values adjusted by ", fit$p_adjust, "."
  )
)
```

```{r plot-scan, fig.cap = "The package's own diagnostic: first-order HSIC-Sensitivity Index per parameter, bars coloured by significance after Benjamini-Hochberg adjustment at alpha = 0.05."}
plot(fit)
```

### The figure: the distributions behind the index

```{r fig-shapes, eval = has_ggplot2, fig.height = 5, fig.cap = "The yield distribution when each parameter is in the lower half of its range against the upper half, one panel per parameter. Under mean_effect the curve slides sideways, which a variance budget also detects. Under var_effect and tail_effect the curve changes width and weight while staying where it is, which is the effect a variance budget is blind to. Under inert the two curves lie on top of each other."}
halves <- do.call(rbind, lapply(colnames(design), function(p1) {
  data.frame(
    parameter = p1,
    half = ifelse(design[, p1] > median(design[, p1]),
                  "upper half of range", "lower half of range"),
    yield = yield[, "yield"],
    stringsAsFactors = FALSE
  )
}))
ggplot2::ggplot(halves, ggplot2::aes(x = yield, fill = half)) +
  ggplot2::geom_density(alpha = 0.55, colour = NA) +
  ggplot2::facet_wrap(~parameter) +
  ggplot2::scale_fill_viridis_d(name = NULL, end = 0.8) +
  ggplot2::labs(
    x = "yield (arbitrary units)", y = "density",
    title = "What each parameter does to the output distribution"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig-shapes-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the conditional",
    "distribution figure is skipped.\n")
```

### Comparing against Sobol

The `sensitivity` package is not a dependency of kernR, so the comparison
below is shown as code rather than run here. The expected contrast is
that a variance-based index ranks `mean_effect` first and returns small
first-order indices for `var_effect` and `tail_effect`, because neither
moves the output variance far.

    half <- nrow(design) / 2
    sobol_fit <- sensitivity::sobolEff(
      model = function(X) stub_apsim(as.matrix(X))[, 1L],
      X1    = as.data.frame(design[seq_len(half), ]),
      X2    = as.data.frame(design[seq_len(half) + half, ]),
      nboot = 200
    )
    # compare sobol_fit$S against fit$index

The disagreement is the point. When the HSIC index ranks a
variance-flat parameter highly, that parameter is changing the output
distribution in a way the variance budget cannot express.

### Total-order indices, and a pure interaction

The first-order index measures what a parameter does on its own. It
misses a parameter that acts only through an interaction, and the
total-order index is built for exactly that case. The synthetic design
below is the extreme version: the output is the product of two
parameters, so each has a marginal expectation of zero and neither
matters alone.

```{r interaction-stub}
set.seed(42L)
n_int <- 400L
theta_int <- matrix(
  runif(n_int * 2L, min = -1, max = 1), n_int, 2L,
  dimnames = list(NULL, c("x1", "x2"))
)
y_int <- theta_int[, "x1"] * theta_int[, "x2"] + rnorm(n_int, sd = 0.05)

fit_int <- hsic_sensitivity(
  theta = theta_int, y = y_int, total_order = TRUE,
  p_value = FALSE, n_permutations = 199L, seed = 42L
)
fit_int
```

```{r interaction-plot, fig.height = 4.5, fig.cap = "The package's own diagnostic: first-order against total-order HSIC-Sensitivity Index for x1 and x2 on the pure-interaction model, side by side. The gap between the two bars is the interaction screen in graphical form."}
plot(fit_int, which = "both")
```

### A near-additive design for contrast

```{r additive-stub}
theta_add <- matrix(
  runif(n_int * 2L), n_int, 2L, dimnames = list(NULL, c("x1", "x2"))
)
y_add <- 2 * theta_add[, "x1"] + theta_add[, "x2"] +
  rnorm(n_int, sd = 0.05)

fit_add <- hsic_sensitivity(
  theta_add, y_add, total_order = TRUE, p_value = FALSE, seed = 1L
)
fit_add
```

```{r gap-table, echo = FALSE}
int_df <- as.data.frame(fit_int)
add_df <- as.data.frame(fit_add)
gap_tab <- data.frame(
  design = c(rep("interaction, Y = X1 X2", nrow(int_df)),
             rep("additive, Y = 2 X1 + X2", nrow(add_df))),
  parameter = c(int_df$parameter, add_df$parameter),
  first_order = c(int_df$index, add_df$index),
  total_order = c(int_df$index_total_order, add_df$index_total_order),
  stringsAsFactors = FALSE
)
gap_tab$gap <- gap_tab$total_order - gap_tab$first_order
knitr::kable(
  gap_tab, digits = c(0L, 0L, 3L, 3L, 3L), row.names = FALSE,
  col.names = c("Design", "Parameter", "First-order S", "Total-order T",
                "T minus S"),
  caption = paste0(
    "First-order and total-order indices on the pure-interaction and ",
    "near-additive designs, both with ", n_int, " design rows."
  )
)
```

### An interval on the total-order index

`total_order_ci = TRUE` attaches a pair-bootstrap percentile interval to
each total-order index. It quantifies uncertainty in the *magnitude* of
the index; it is not a hypothesis test.

```{r pairboot-ci}
fit_ci <- hsic_sensitivity(
  theta_add, y_add, total_order = TRUE, total_order_ci = TRUE,
  n_permutations = 199L, n_bootstrap = 100L, ci_level = 0.95, seed = 7L
)
ci_df <- as.data.frame(fit_ci)[, c(
  "parameter", "output", "index_total_order",
  "ci_total_order_lower", "ci_total_order_upper"
)]
```

```{r pairboot-table, echo = FALSE}
knitr::kable(
  ci_df, digits = 3L, row.names = FALSE,
  col.names = c("Parameter", "Output", "Total-order T",
                "Lower", "Upper"),
  caption = paste0(
    "Pair-bootstrap percentile interval on each total-order index, ",
    100 * fit_ci$ci_level, " per cent level, ", fit_ci$n_bootstrap,
    " resamples."
  )
)
```

An earlier release attached a `total_order_p_value` to these bootstrap
draws, claiming to test $H_0 : T_j = 0$. Critical review on 2026-05-16
found it mis-calibrated under a pure-noise output, because the bootstrap
resamples the empirical joint distribution rather than a
null-of-no-effect, and every parameter was assigned a p-value of
approximately $1 / (1 + B)$ regardless. The field was removed.

### The properly calibrated significance test

The replacement tests $H_0$: $X_j$ is independent of $Y$ given
$X_{\sim j}$, by clustering the design on similarity in $X_{\sim j}$,
permuting the output within each cluster, and recomputing $T_j$ on every
permuted design.

```{r cond-perm-test}
fit_cp <- hsic_sensitivity(
  theta_add, y_add, total_order = TRUE,
  total_order_test = "cond_perm", n_permutations = 199L, seed = 21L
)
cp_df <- as.data.frame(fit_cp)[, c(
  "parameter", "output", "index_total_order", "p_value_total_order"
)]
```

```{r cond-perm-table, echo = FALSE}
knitr::kable(
  cp_df, digits = c(0L, 0L, 3L, 4L), row.names = FALSE,
  col.names = c("Parameter", "Output", "Total-order T",
                "Conditional-permutation p-value"),
  caption = paste0(
    "Conditional-permutation test of the total-order index on the ",
    "near-additive design, ", fit_cp$n_permutations, " permutations."
  )
)
```

## Read

On the `r nrow(design)`-row design the index ranks `mean_effect` first at
`r signif(fit$index[["mean_effect", "yield"]], 3)`, with an adjusted
p-value of
`r signif(scan_df$p_value_adjusted[
  scan_df$parameter == "mean_effect"], 3)`; a variance budget would find
that one too. The parameter that carries the argument is `var_effect`,
which returns `r signif(fit$index[["var_effect", "yield"]], 3)` with an
adjusted p-value of
`r signif(scan_df$p_value_adjusted[
  scan_df$parameter == "var_effect"], 3)`, despite being built to leave
the conditional mean exactly where it was. `inert` returns
`r signif(fit$index[["inert", "yield"]], 3)` with an adjusted p-value of
`r signif(scan_df$p_value_adjusted[scan_df$parameter == "inert"], 3)`,
which is the floor the design was built to produce.

`tail_effect` is the instructive case. Its index is
`r signif(fit$index[["tail_effect", "yield"]], 3)`, above `inert` but
well below `var_effect`, and its adjusted p-value is
`r signif(scan_df$p_value_adjusted[
  scan_df$parameter == "tail_effect"], 3)`, which
`r if (scan_df$p_value_adjusted[scan_df$parameter == "tail_effect"] <=
  alpha_level) "clears" else "does not clear"` the conventional
threshold of `r alpha_level`. The effect is real by construction, and at
this design size the screen does not have the power to certify it: a
heavy tail
scaled by a parameter in the unit interval is a weaker signal than a
change of spread, and a larger design or a longer permutation run is what
would settle it. Reporting that is the point of the p-value column.

The figure shows what the index is responding to: the `var_effect` and
`tail_effect` panels have two curves of different width and tail weight
sitting in the same place, which is the signature a variance budget
cannot express, while the `inert` panel has two curves lying on top of
each other.

On the pure-interaction design the first-order indices are
`r signif(int_df$index[int_df$parameter == "x1"], 3)` and
`r signif(int_df$index[int_df$parameter == "x2"], 3)` -- close to
nothing, because the marginal effect of either parameter averages out
over the other -- while the total-order indices are
`r signif(int_df$index_total_order[int_df$parameter == "x1"], 3)` and
`r signif(int_df$index_total_order[int_df$parameter == "x2"], 3)`. The
gap between them is
`r signif(mean(int_df$index_total_order - int_df$index), 2)` on average,
against `r signif(mean(add_df$index_total_order - add_df$index), 2)` on
the near-additive design. That contrast is the interaction screen: a
large gap is evidence of interaction, a small one is not evidence of its
absence, and the difference has no null distribution of its own.

The pair-bootstrap interval on the near-additive design puts the
total-order index for `x1` between
`r signif(ci_df$ci_total_order_lower[ci_df$parameter == "x1"], 3)` and
`r signif(ci_df$ci_total_order_upper[ci_df$parameter == "x1"], 3)`, an
interval on the magnitude and not a test. The conditional-permutation
test supplies the verdict the interval cannot: on this additive design
the two total-order p-values are
`r signif(cp_df$p_value_total_order[cp_df$parameter == "x1"], 3)` and
`r signif(cp_df$p_value_total_order[cp_df$parameter == "x2"], 3)`, so
neither total-order index survives its own null. That is the expected
answer where there is no interaction to find, and it agrees with the
small `T` minus `S` gap in the table above.

## Limits

The simulator here is a stub with independent, uniformly distributed
parameters, chosen so that each effect is isolated and the right answer
is known; correlated parameters break the interpretation of both indices,
because $X_{\sim j}$ then carries information about $X_j$. The
HSIC-Sensitivity Index is not numerically equal to a Sobol index even on
a linear additive system: the two are bounded comparably, both lying in
the unit interval, and typically agree on ranking when only mean effects
are present, so the disagreement rather than the agreement is what
carries information. The quantity `T` minus `S` is an uncalibrated
contrast, not a variance-budget identity, and is not zero on an additive
model, because the two indices are built from different normalisers; the
conditional-permutation test is the calibrated statement about the same
question. The Sobol comparison in this vignette is stated as code and a
claim about the direction of the result; it is not run here, and a reader
who needs the numbers should run it. Finally, every p-value here is a
permutation p-value bounded below by `1 / (n_permutations + 1)`.

## Notes on practice

**Design size.** As for the identifiability screen, a design with at
least ten rows per parameter is a workable starting point.

**Permutations.** When the goal is ranking only, `p_value = FALSE` skips
the permutation null entirely; the per-column kernel cache is reused, so
the indices themselves are cheap. Use the default `p_value = TRUE` when a
significance verdict on the first-order indices is also wanted.

**Cost.** Total-order construction costs order $p n^2$ for the kernel
matrices on each $X_{\sim j}$ plus order $p q n^2$ for the HSIC
evaluations, similar in scale to first-order. The pair bootstrap
multiplies that by `n_bootstrap`, which defaults to 200, and the
conditional-permutation test multiplies it by the permutation count, so
both should be budgeted for.

**Nystrom acceleration for total-order** is deliberately not exposed.
Folding `nystrom_factor()` into the total-order computation by
materialising the low-rank product into a full kernel matrix costs order
$n^2 m$ and is slower than the exact path at the scales kernR runs at; a
factor-only HSIC primitive is the step that would unblock it.

## What to read next

*Which parameters are worth calibrating?* is the pass-or-fail version of
this scan, used before a calibration to decide which parameters enter the
prior at all. *Scaling HSIC to large ensembles* covers the low-rank
approximations that make the first-order scan affordable on designs too
large for an exact Gram matrix. *Getting started with kernR* introduces
the HSIC statistic these indices normalise.

## References

- Da Veiga, S. (2015). Global sensitivity analysis with dependence
  measures. *Journal of Statistical Computation and Simulation*, 85(7),
  1283-1305.
- Gretton, A., Bousquet, O., Smola, A., & Scholkopf, B. (2005).
  Measuring statistical dependence with Hilbert-Schmidt norms.
  *Algorithmic Learning Theory*, 63-77.

## Reproduce

Seed `1` for the simulator noise, `seed = 11L` for the Latin-hypercube
design and the first-order permutation null, `42` for the
pure-interaction design, and `seed` values `1`, `7` and `21` for the
additive design, the pair bootstrap and the conditional-permutation test.
Package versions follow.

```{r session-info}
sessionInfo()
```
