---
title: "From a coarse grid to a paddock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a coarse grid to a paddock}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~3 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. One conditional-mean-embedding fit with leave-one-out selection
     of the ridge penalty, five distribution regressions on 60 bags, and two
     aggregate inversions, one of them by importance sampling. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

An agronomist has the wrong resolution. *The climate product I can get is
a grid cell far larger than any paddock I manage, and the decision I have
to make is about a paddock. Sometimes I have historical paddock records
to learn the step down from; sometimes what I have per paddock is a bag
of soil cores rather than one number; and sometimes I have no paddock
data at all, only the coarse observation and the knowledge that it is an
average of the paddocks inside it. Are those three the same problem?*
This vignette says they are three problems, gives the kernR primitive for
each, and shows on synthetic data what each one needs and what it
returns.

## What

`kernel_downscale()` handles the case with paired training data. Each
training row is one coarse-resolution vector and one fine-resolution
vector, and the fit is a conditional mean embedding: kernel ridge
regression in conditional-distribution form, with the ridge penalty
chosen by leave-one-out cross-validation and the kernel bandwidths by the
median heuristic. `fit_cme()` exposes the same operator at a lower level
for callers who want to build on the conditional embedding directly.

`dist_regression()` handles the case where the coarse covariate is itself
a *bag of points* rather than a vector: a paddock's soil cores, an
ensemble of climate realisations, a set of spectral samples. Each bag is
summarised by its empirical kernel mean embedding, and an outer kernel on
those embeddings carries the regression. `outer = "linear"` is the inner
product of embeddings; `outer = "rbf"` is a Gaussian on embedding-space
distance, and is the one to reach for when the map from embedding to
target is itself non-linear. Bags may differ in size.

`aggregate_downscale()` handles the case with no paired data at all. What
is available is an aggregate observation `y = T(x) + e` for a *known*
operator `T` -- a spatial or temporal average, a satellite footprint
convolution, a mass-balance constraint -- plus a parametric prior on the
fine-scale latent `x`, which might come from `proxymix::fit_proxymix()`
fitted to historical fine-scale data. For a linear operator the posterior
is a per-component Kalman update with reweighted mixture weights, exact
in closed form; for a non-linear operator it is importance sampling
within each prior component, with the per-component effective sample size
reported as the reliability diagnostic.
`posterior_sample_aggregate()` draws from the resulting mixture
posterior.

## Do

### Paired data: coarse climate to paddock yield

A synthetic training set of 80 seasons. The coarse covariates are a mean
monthly temperature and a cumulative rainfall; the fine-resolution
outputs are a paddock yield, quadratic in temperature and linear in
rainfall, and a biomass, linear in both.

```{r downscale-setup}
set.seed(1L)
n_train <- 80L
coarse <- matrix(
  c(rnorm(n_train, mean = 18, sd = 2),
    rnorm(n_train, mean = 450, sd = 80)),
  ncol = 2L, dimnames = list(NULL, c("temp", "rainfall"))
)
truth <- function(z) {
  cbind(
    yield = 0.02 * z[, "rainfall"] - 0.1 * (z[, "temp"] - 18)^2 +
      rnorm(nrow(z), sd = 0.5),
    biomass = 0.03 * z[, "rainfall"] + 0.05 * z[, "temp"] +
      rnorm(nrow(z), sd = 0.7)
  )
}
fine <- truth(coarse)
```

```{r downscale-fit}
new_coarse <- matrix(
  c(rnorm(20L, 18, 2), rnorm(20L, 450, 80)),
  ncol = 2L, dimnames = list(NULL, c("temp", "rainfall"))
)
fit_raw <- kernel_downscale(coarse, fine, new_coarse)
fit_raw
head(round(fit_raw$prediction, 3L))
```

The two coarse columns are on very different scales: temperature has a
standard deviation of about 2 and rainfall about 80. The median heuristic
resolves one bandwidth for the joint input, so on the raw scale the
pairwise distances are almost entirely rainfall and the temperature
response is nearly invisible to the kernel. Standardising both columns
before the fit is the fix, and the comparison below is worth making
explicitly because it is the most common way a conditional mean embedding
underperforms in practice.

```{r downscale-scaled}
coarse_centre <- colMeans(coarse)
coarse_scale <- apply(coarse, 2L, sd)
coarse_std <- scale(coarse, center = coarse_centre, scale = coarse_scale)
new_std <- scale(new_coarse, center = coarse_centre, scale = coarse_scale)
fit_cme_paddock <- kernel_downscale(coarse_std, fine, new_std)
fit_cme_paddock
```

```{r downscale-error, echo = FALSE}
new_truth <- truth(new_coarse)
rmse_of <- function(pred) {
  vapply(seq_len(ncol(pred)), function(j1) {
    sqrt(mean((pred[, j1] - new_truth[, j1])^2))
  }, numeric(1L))
}
cme_err <- data.frame(
  output = rep(colnames(fine), 2L),
  fit = rep(c("raw scale", "standardised"), each = ncol(fine)),
  rmse = c(rmse_of(fit_raw$prediction),
           rmse_of(fit_cme_paddock$prediction)),
  sd_truth = rep(apply(new_truth, 2L, sd), 2L),
  noise_sd = rep(c(0.5, 0.7), 2L),
  stringsAsFactors = FALSE
)
cme_err$rmse_over_sd <- cme_err$rmse / cme_err$sd_truth
knitr::kable(
  cme_err[, c("output", "fit", "rmse", "sd_truth", "noise_sd",
              "rmse_over_sd")],
  digits = 3L, row.names = FALSE,
  col.names = c("Output", "Fit", "RMSE", "SD of held-out draw",
                "Noise SD in the generator", "RMSE / SD"),
  caption = paste0(
    "Prediction error at ", nrow(new_coarse),
    " held-out coarse points, against a fresh draw from the same ",
    "generating process, on the raw and standardised input scales."
  )
)
```

### The figure: the downscaled surface against the field it came from

```{r fig-surface, eval = has_ggplot2, fig.height = 5, fig.cap = "Downscaled prediction against the coarse rainfall input at the 20 held-out points (line and points), with a fresh draw from the generating process at the same points for comparison (crosses). The conditional mean embedding recovers the rainfall response in both outputs; the vertical scatter of the crosses around the line is the noise the fit is not trying to reproduce."}
surface_df <- rbind(
  data.frame(
    rainfall = new_coarse[, "rainfall"],
    value = fit_cme_paddock$prediction[, "yield"],
    output = "yield", source = "downscaled prediction"
  ),
  data.frame(
    rainfall = new_coarse[, "rainfall"], value = new_truth[, "yield"],
    output = "yield", source = "fresh draw from the field"
  ),
  data.frame(
    rainfall = new_coarse[, "rainfall"],
    value = fit_cme_paddock$prediction[, "biomass"],
    output = "biomass", source = "downscaled prediction"
  ),
  data.frame(
    rainfall = new_coarse[, "rainfall"], value = new_truth[, "biomass"],
    output = "biomass", source = "fresh draw from the field"
  )
)
ggplot2::ggplot(
  surface_df,
  ggplot2::aes(x = rainfall, y = value, colour = source, shape = source)
) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::facet_wrap(~output, scales = "free_y") +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.8) +
  ggplot2::scale_shape_manual(name = NULL, values = c(4L, 16L)) +
  ggplot2::labs(
    x = "coarse cumulative rainfall (mm)",
    y = "fine-resolution output",
    title = "Downscaled prediction against the field"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig-surface-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the downscaled-surface",
    "figure is skipped.\n")
```

### Bags of points: distribution regression

The simplest illustrative target: recover the mean of a distribution
from a finite sample of it. Sixty synthetic bags, forty points each.

```{r dr-mean}
set.seed(2L)
n_bags <- 60L
mu_train <- runif(n_bags, -3, 3)
bags_train <- lapply(mu_train, function(m1) {
  matrix(rnorm(40L, mean = m1, sd = 1), ncol = 1L)
})
fit_dr <- dist_regression(bags_train, y = mu_train, outer = "linear")
fit_dr
```

```{r dr-predict}
n_new <- 10L
mu_new <- runif(n_new, -3, 3)
bags_new <- lapply(mu_new, function(m1) {
  matrix(rnorm(40L, mean = m1, sd = 1), ncol = 1L)
})
pred_dr <- predict(fit_dr, bags_new)
```

```{r dr-table, echo = FALSE}
dr_tab <- data.frame(
  truth = mu_new, prediction = pred_dr,
  abs_error = abs(pred_dr - mu_new)
)
knitr::kable(
  dr_tab, digits = 3L, row.names = FALSE,
  col.names = c("True bag mean", "Prediction", "Absolute error"),
  caption = paste0(
    "Distribution regression on ", n_new, " fresh bags of 40 points."
  )
)
```

Targets may be vectors as well as scalars, one column per bag-level
quantity.

```{r dr-multivariate}
y_mat <- cbind(mean = mu_train, mean_sq = mu_train^2)
fit_mv <- dist_regression(bags_train, y = y_mat, outer = "linear")
round(predict(fit_mv, bags_new[1:3]), 3L)
```

### Choosing the outer kernel

The outer kernel governs the regression on embedding space, not what the
embedding can represent. The synthetic bags below are all centred at zero
and differ only in their spread, so the target is a property of the bag
that no bag mean carries, and the two outer kernels can be compared on
it directly.

```{r dr-variance}
set.seed(3L)
sd_train <- runif(n_bags, 0.5, 2.5)
bags_var <- lapply(sd_train, function(s1) {
  matrix(rnorm(60L, mean = 0, sd = s1), ncol = 1L)
})
fit_linear <- dist_regression(bags_var, y = sd_train, outer = "linear")
fit_rbf <- dist_regression(bags_var, y = sd_train, outer = "rbf")

sd_new <- runif(30L, 0.5, 2.5)
bags_sd_new <- lapply(sd_new, function(s1) {
  matrix(rnorm(60L, mean = 0, sd = s1), ncol = 1L)
})
rmse_outer <- c(
  linear = sqrt(mean((predict(fit_linear, bags_sd_new) - sd_new)^2)),
  rbf = sqrt(mean((predict(fit_rbf, bags_sd_new) - sd_new)^2)),
  spread_of_target = sd(sd_new)
)
round(rmse_outer, 3L)
```

Bag sizes may differ across the design, which is what a set of paddocks
with different numbers of soil cores looks like; the double-sum embedding
averages over the inner Gram matrix, so nothing needs padding.

```{r dr-variable-size}
set.seed(4L)
mu_v <- runif(20L, -2, 2)
bags_v <- lapply(mu_v, function(m1) {
  matrix(rnorm(sample(20:80, 1L), mean = m1), ncol = 1L)
})
fit_v <- dist_regression(bags_v, y = mu_v, outer = "linear")
range(vapply(fit_v$bags_train, nrow, integer(1L)))
```

### No paired data: inverting a known aggregator

Two adjacent paddocks are observed only through their spatial average.
The prior on the pair is a synthetic two-component Gaussian mixture, one
component near the origin and one near `(2, 2)`, and the observation
points at the second.

```{r aggregate-linear}
agg_matrix <- matrix(c(0.5, 0.5), nrow = 1L)
latent_prior <- list(
  means = list(c(0, 0), c(2, 2)),
  covariances = list(diag(2L), diag(2L)),
  weights = c(0.5, 0.5)
)
fit_agg_lin <- aggregate_downscale(
  y = 1.8, aggregator = agg_matrix, latent_prior = latent_prior,
  sigma_y = 0.15
)
fit_agg_lin
```

A non-linear aggregator, standing in for a sensor whose response
saturates, is handled by importance sampling within each prior
component.

```{r aggregate-nonlinear}
agg_fn <- function(x) matrix(sin(rowSums(x)), ncol = 1L)
fit_agg_nl <- aggregate_downscale(
  y = 0.5, aggregator = agg_fn, latent_prior = latent_prior,
  sigma_y = 0.1, n_samples_per_component = 400L, seed = 1L
)
round(fit_agg_nl$ess_per_component, 1L)
round(fit_agg_nl$posterior_mean, 3L)
fit_agg_nl$ess_warning
```

```{r aggregate-sample}
samples_agg <- posterior_sample_aggregate(fit_agg_nl, n = 500L,
                                          seed = 2L)
round(head(samples_agg, 3L), 3L)
```

### The three primitives side by side

```{r comparison-table, echo = FALSE}
comparison <- data.frame(
  aspect = c("Training data", "Coarse-to-fine map", "Latent prior",
             "Closed form", "Multi-output", "Uncertainty returned"),
  kernel_downscale = c(
    "paired coarse and fine", "learned by regression",
    "implicit, empirical", "yes", "yes", "point plus ridge interval"
  ),
  dist_regression = c(
    "bags plus bag-level targets", "learned by regression",
    "implicit, empirical", "yes", "yes", "point plus ridge interval"
  ),
  aggregate_downscale = c(
    "none paired, aggregate only", "known operator",
    "explicit, parametric mixture",
    "yes for a linear operator, importance sampling otherwise",
    "yes", "full mixture posterior"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  comparison,
  col.names = c("Aspect", "kernel_downscale()", "dist_regression()",
                "aggregate_downscale()"),
  caption = "What each primitive needs and what it gives back."
)
```

## Read

The input scale matters more than anything else in this fit. On the raw
scale the conditional mean embedding predicts yield at the
`r nrow(new_coarse)` held-out coarse points with a root mean squared
error of
`r round(cme_err$rmse[cme_err$output == "yield" &
  cme_err$fit == "raw scale"], 3)`, which is
`r round(100 * cme_err$rmse_over_sd[cme_err$output == "yield" &
  cme_err$fit == "raw scale"])` per cent of the spread of the held-out
values and far above the irreducible floor of 0.5 that the generator's
own noise sets. Standardising the two coarse columns first brings it to
`r round(cme_err$rmse[cme_err$output == "yield" &
  cme_err$fit == "standardised"], 3)`, essentially the noise floor: the
fit now recovers the yield response as well as anyone could. Biomass
moves the other way by a smaller amount, from
`r round(cme_err$rmse[cme_err$output == "biomass" &
  cme_err$fit == "raw scale"], 3)` to
`r round(cme_err$rmse[cme_err$output == "biomass" &
  cme_err$fit == "standardised"], 3)` against its own noise floor of 0.7,
because biomass depends on rainfall much more strongly than on
temperature and the raw-scale bandwidth was already tuned to rainfall.
The single shared bandwidth is what forces the trade: it is the reason to
standardise, and the reason a per-column bandwidth would be the better
tool if one output needed a different resolution from another. The ridge
penalty chosen by leave-one-out on the standardised fit was
`r formatC(fit_cme_paddock$lambda, format = "e", digits = 2L)`. The
figure shows the standardised fit as a surface rather than a number: the
predictions trace the rainfall response cleanly in both outputs, and the
crosses scatter around them by roughly the noise that was put in.

Distribution regression recovers the bag means it was asked for: across
the `r n_new` fresh bags the mean absolute error is
`r round(mean(dr_tab$abs_error), 3)` on bag means spanning
`r round(min(mu_new), 2)` to `r round(max(mu_new), 2)`, and the largest
single error is `r round(max(dr_tab$abs_error), 3)`.

The spread-only bags settle what the outer kernel does and does not do.
Both outer kernels recover the bag standard deviation well: the linear
one gives a root mean squared error of
`r round(rmse_outer[["linear"]], 3)` and the Gaussian one
`r round(rmse_outer[["rbf"]], 3)`, against a target spread of
`r round(rmse_outer[["spread_of_target"]], 3)`, so on this design the two
are `r if (abs(rmse_outer[["linear"]] - rmse_outer[["rbf"]]) <
  0.1 * rmse_outer[["linear"]]) "indistinguishable" else
  "clearly separated"`. That is worth stating plainly, because the
natural expectation is the opposite. A linear *outer* kernel is not a
linear model of the bag: the inner RBF kernel embeds each bag as a
function, and a bag with a wider spread has a visibly different embedding
from a narrow one at the same mean, so the higher moment is already
carried before the outer kernel sees anything. The outer kernel chooses
how the regression behaves on embedding space, and the Gaussian one earns
its keep when that relationship is itself non-linear, not merely when the
target is a higher moment.

The linear aggregate inversion moves the posterior weight where the
observation points. The prior gave the two components weights of
`r paste(latent_prior$weights, collapse = " and ")`; after observing an
average of 1.8 with a noise standard deviation of 0.15, the posterior
weights are
`r paste(round(fit_agg_lin$posterior_weights, 3), collapse = " and ")`,
and the posterior mean is
`r paste(round(fit_agg_lin$posterior_mean, 2), collapse = ", ")`, close
to the `(2, 2)` component the observation is consistent with. The
non-linear inversion is where the reliability diagnostic earns its place:
its per-component effective sample sizes are
`r paste(round(fit_agg_nl$ess_per_component, 1), collapse = " and ")` out
of `r fit_agg_nl$n_samples_per_component` importance draws each, and
`ess_warning` is `r fit_agg_nl$ess_warning`. A collapsed effective sample
size there means the posterior is being carried by a handful of draws and
should be read as exploratory rather than as a result.

## Limits

Every dataset here is synthetic, with the generating process known and
the coarse-to-fine relationship smooth; a real downscaling problem has
covariate shift between the training seasons and the season to be
predicted, which none of these primitives detects. `kernel_downscale()`
returns a conditional *mean*, not a conditional distribution, so it gives
a point prediction with a ridge interval and cannot answer a question
about the fine-scale tail; the closed-form solve costs order `n^3` at
training time, which bounds the training set well before the prediction
step becomes the constraint. `dist_regression()` assumes the bags are
samples from the distributions of interest, so a bag that is
systematically sampled -- the cores taken from the accessible corner of
the paddock -- biases the embedding with no diagnostic to say so.
`aggregate_downscale()` requires the aggregator to be *known*, and its
answer is only as good as the latent prior it inverts: a prior that
excludes the truth returns a confident posterior in the wrong place. The
per-component effective sample size on the non-linear path is the one
reliability gate in this vignette, and it detects importance-sampling
collapse, not prior misspecification.

## Notes on practice

**Sizing a conditional mean embedding.** Both training and prediction
inputs must have the same coarse-resolution shape, one vector per row.
The closed-form ridge solve costs order `n^3` at training time and
predicts in order `n_new` times `n_train`, so the training set is the
binding constraint; standardise the input columns before fitting unless
they are already on a common scale.

**Sizing a distribution regression.** Tens to hundreds of points per bag
and tens to low thousands of bags is the range this construction is built
for. The inner Gram matrices cost order `M^2` times the square of the
mean bag size, where `M` is the number of bags, so bag size is the more
expensive of the two dimensions.

**Sizing an aggregate inversion.** The linear path costs order `N` times
the cube of the latent dimension plus the cube of the observation
dimension, for `N` prior components. The non-linear path costs order `N`
times the draws per component times the sum of the two dimensions, and
its `n_samples_per_component` should be raised until the per-component
effective sample sizes are a workable fraction of it.

**Reproducibility.** The non-linear aggregate path and
`posterior_sample_aggregate()` are randomised; pass `seed` to both.

## What to read next

*Which parameters are worth calibrating?* and *Did the management change
shift the simulated distribution?* cover the calibration step that
usually produces the ensemble a latent prior is fitted to. *Choosing a
density-ratio backend* describes the proxymix binding, whose
`fit_proxymix()` is the canonical fitter for the Gaussian-mixture prior
`aggregate_downscale()` inverts. *Scaling HSIC to large ensembles*
covers the low-rank factorisations that would be the route to a
conditional mean embedding on a training set too large for an exact
solve.

## References

- Muandet, K., Fukumizu, K., Sriperumbudur, B., & Scholkopf, B. (2017).
  Kernel mean embedding of distributions: a review and beyond.
  *Foundations and Trends in Machine Learning*, 10(1-2), 1-141.
- Park, J., Muandet, K., Fukumizu, K., & Sejdinovic, D. (2013). Kernel
  embeddings of conditional distributions. *IEEE Signal Processing
  Magazine*, 30(4), 98-111.
- Szabo, Z., Sriperumbudur, B. K., Poczos, B., & Gretton, A. (2016).
  Learning theory for distribution regression. *Journal of Machine
  Learning Research*, 17(152), 1-40.

## Reproduce

Seeds `1` (paired climate toy), `2` (bags for the mean target), `3`
(spread-only bags), `4` (variable-size bags), and `seed = 1L` and
`seed = 2L` for the importance sampler and the posterior draw. Package
versions follow.

```{r session-info}
sessionInfo()
```
