---
title: "Density-Ratio Backends and the proxymix Binding"
author: "Max Moldovan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density-Ratio Backends and the proxymix Binding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
have_proxymix <- requireNamespace("proxymix", quietly = TRUE)
```

## The cross-package contract

`kernR::estimate_density_ratio()` is the entry point for every kernR
estimator that needs to reweight observational samples to an
interventional distribution — most notably `bd_hsic_test()` for the
backdoor-HSIC causal test. The function offers four density-ratio
backends behind a single signature:

| Backend       | Family                          | When to use                                         |
|---------------|---------------------------------|-----------------------------------------------------|
| `logistic`    | Noise-contrastive classifier    | Default; robust on smooth, unimodal densities       |
| `ranger`      | Random-forest classifier        | Flexible non-linear NCE; needs `ranger`             |
| `xgboost`     | Gradient-boosted classifier     | Strong on tabular interactions; needs `xgboost`     |
| `proxymix`    | Gaussian-mixture density ratio  | Multimodal/skewed densities; parametric alternative |

The `proxymix` backend is the cross-package wedge between kernR (the
distributional verdict layer of the UQ ag stack) and proxymix (the
Gaussian-mixture proxy / KL-density-ratio bridge, Hoek & Elliott 2024).
It fits one GMM to the joint sample cloud `(x, z)`, a second GMM to the
product-of-marginals cloud `(x_perm, z)`, then evaluates the analytic
ratio of the two mixture densities at each observation. No classifier
calibration step — the ratio is closed-form in the fitted parameters.

```{r dep-note, echo=FALSE, results='asis'}
if (!have_proxymix) {
  cat(
    "> **Note.** The `proxymix` package is not installed in the current ",
    "library; chunks that compare the GMM backend below are skipped. ",
    "Install via `remotes::install_github(\"max578/proxymix\")` or from ",
    "the local source tarball to reproduce the full comparison.\n\n",
    sep = ""
  )
}
```

## Four backends, one problem

A toy confounded design: `z` is a 2-D Gaussian confounder; `x` is a
linear-Gaussian function of `z`; `y` carries a real causal effect from
`x` plus a confounded path through `z`.

```{r data}
suppressPackageStartupMessages(library(kernR))
set.seed(2026L)
n <- 200L
z <- matrix(rnorm(n * 2L), n, 2L)
x <- z[, 1L] + rnorm(n, sd = 0.5)
y <- 0.7 * x + z[, 2L] + rnorm(n, sd = 0.4)
```

Each backend produces a `density_ratio_fit` object exposing ESS and a
weight vector. We tabulate them side-by-side.

```{r logistic-ranger-xgb}
dr_logistic <- estimate_density_ratio(x, z, method = "logistic", seed = 1L)
dr_ranger <- if (requireNamespace("ranger", quietly = TRUE)) {
  estimate_density_ratio(x, z, method = "ranger", seed = 1L)
} else NULL
dr_xgb <- if (requireNamespace("xgboost", quietly = TRUE)) {
  estimate_density_ratio(x, z, method = "xgboost", seed = 1L)
} else NULL
```

```{r proxymix-fit, eval=have_proxymix}
dr_proxymix <- estimate_density_ratio(
  x, z,
  method = "proxymix",
  proxymix_components = 2L,
  seed = 1L
)
```

```{r summary-table, echo=FALSE}
rows <- list(
  list(name = "logistic", fit = dr_logistic),
  if (!is.null(dr_ranger)) list(name = "ranger",   fit = dr_ranger),
  if (!is.null(dr_xgb))    list(name = "xgboost",  fit = dr_xgb),
  if (have_proxymix)       list(name = "proxymix", fit = dr_proxymix)
)
rows <- Filter(Negate(is.null), rows)
summary_df <- do.call(rbind, lapply(rows, function(r) {
  w <- r$fit$weights
  data.frame(
    backend     = r$name,
    ess         = round(r$fit$ess, 1),
    min_weight  = signif(min(w), 3),
    max_weight  = signif(max(w), 3),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(summary_df, caption = "Density-ratio diagnostics across backends.")
```

## bd-HSIC under each backend

```{r bdhsic-classifiers}
res_logistic <- bd_hsic_test(
  x, y, z, density_ratio = "logistic",
  n_permutations = 199L, seed = 1L
)
```

```{r bdhsic-proxymix, eval=have_proxymix}
res_proxymix <- bd_hsic_test(
  x, y, z, density_ratio = "proxymix",
  n_permutations = 199L, seed = 1L
)
```

```{r bdhsic-table, echo=FALSE}
rows2 <- list(
  data.frame(backend = "logistic", statistic = signif(res_logistic$statistic, 3),
             p_value = res_logistic$p_value, ess = round(res_logistic$ess, 1),
             stringsAsFactors = FALSE)
)
if (have_proxymix) {
  rows2 <- c(rows2, list(
    data.frame(backend = "proxymix", statistic = signif(res_proxymix$statistic, 3),
               p_value = res_proxymix$p_value, ess = round(res_proxymix$ess, 1),
               stringsAsFactors = FALSE)
  ))
}
knitr::kable(do.call(rbind, rows2),
             caption = "bd-HSIC test under each density-ratio backend.")
```

## When to reach for proxymix

The classifier-based backends (`logistic`, `ranger`, `xgboost`) are the
default for a reason: they tolerate misspecified densities, scale to
high-dimensional `z`, and have well-understood calibration. Reach for
`method = "proxymix"` when:

* the joint density is plausibly **multimodal** (multi-regime climate,
  paddock × variety designs with distinct production zones, animal
  cohorts with separable subpopulations) — a 2-component GMM
  represents this cleanly where a classifier would smear across the
  modes;
* you need a **parametric** density-ratio whose components you can
  inspect, hand off to a downstream Bayesian step (via
  `proxymix::gmm_target_from_posterior()`), or use as the seed of a
  KLD-EM refinement on a target you can evaluate but not sample from;
* classifier calibration is unreliable on the cohort at hand
  (small `n`, sharp class imbalance after the joint-vs-marginal split,
  pathological feature scaling).

The `proxymix` package is GRDC-firewalled (MIT, no GRDC IP flows in)
and ships its full Gaussian-mixture proxy API independently of kernR.
kernR consumes it as a soft dependency via `requireNamespace()` — the
binding is one-way and rebuildable from the local
`proxymix_*.tar.gz` source.

## References

* Hoek, J. van der & Elliott, R. J. (2024). *Mixtures of multivariate
  Gaussians.* Stochastic Analysis and Applications.
  DOI: 10.1080/07362994.2024.2372605.
* Hu, R., Sejdinovic, D. & Evans, R. J. (2024). A kernel test for
  causal association via noise contrastive backdoor adjustment.
  *Journal of Machine Learning Research*, 25(160), 1–56.

```{r session-info}
sessionInfo()
```
