---
title: "Scaling HSIC to large ensembles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling HSIC to large ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~10 s under rmarkdown::render(); macOS aarch64, R 4.5.2,
     one core. One reconstruction check at n = 80 and a four-point timing
     grid from n = 250 to n = 2000; the exact test at n = 2000 dominates,
     which is the vignette's point. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

An analyst has outgrown the exact test. *My ensemble used to have three
hundred members and the independence test was instant. It now has
several thousand, the Gram matrix alone is hundreds of megabytes, and
every permutation iterates over it. I do not want a different test with
different assumptions; I want the same test to finish. What does the
approximation cost me, and does it change the verdict?* This vignette
answers with the two low-rank approximations kernR ships, a
reconstruction check against the exact kernel, and a timing grid from
250 to 2000 observations.

## What

An exact kernel method stores an `n` by `n` Gram matrix and iterates over
it once per permutation, so both memory and time grow with the square of
the sample size. Two standard low-rank factorisations replace that matrix
with a thin factor `F` such that `K` is approximately `F F'`, dropping
the cost to order `n m` for a rank `m` much smaller than `n`.

`nystrom_factor()` implements the Nystrom method of Williams and Seeger
(2001): sample $m$ landmark observations, compute their Gram matrix $W$
($m$ by $m$) and the cross-Gram $C$ ($n$ by $m$), and approximate

\[ K \approx C W^{-1} C^{\top} = F F^{\top}, \qquad
   F = C \, \mathrm{chol}(W)^{-1}. \]

It works for any kernel and is data-dependent, because the landmarks are
drawn from the data.

`rff_features()` implements Random Fourier Features (Rahimi and Recht,
2007), available for shift-invariant kernels, which in kernR means the
RBF kernel. Draw frequencies $\omega$ from a normal with covariance
$\sigma^{-2} I$ and offsets $b$ uniform on $[0, 2\pi]$; the feature map

\[ \phi(x) = \sqrt{2 / D} \, \cos(\omega^{\top} x + b) \]

gives $K \approx \Phi \Phi^{\top}$ for $D$ features. The projection is
data-independent, so the same random features apply to any point cloud at
that bandwidth.

`hsic_test_nystrom()` wires both into a drop-in accelerated independence
test: `method` selects the factorisation and `m` sets the rank or the
feature count. It returns the same `kernel_test_result` as `hsic_test()`.

## Do

### Does the factorisation reproduce the kernel?

A synthetic cloud of 80 points in two dimensions, small enough that the
exact Gram matrix can be compared against both approximations directly.

```{r correctness}
set.seed(1L)
n_check <- 80L
x_check <- matrix(rnorm(n_check * 2L), n_check, 2L)
kern <- kernel_spec("rbf", bandwidth = 1.0)
k_full <- kernel_matrix(x_check, kernel = kern)

fac_ny <- nystrom_factor(x_check, kernel = kern, m = n_check - 1L,
                         seed = 1L)
k_ny <- tcrossprod(fac_ny$F)

fac_rf <- rff_features(x_check, kernel = kern, D = 1500L, seed = 1L)
k_rf <- tcrossprod(fac_rf$F)

rel_err <- function(k_hat) {
  sqrt(sum((k_full - k_hat)^2)) / sqrt(sum(k_full^2))
}
recon <- data.frame(
  method = c("nystrom", "rff"),
  rank = c(ncol(fac_ny$F), ncol(fac_rf$F)),
  rel_err = c(rel_err(k_ny), rel_err(k_rf)),
  stringsAsFactors = FALSE
)
```

```{r correctness-table, echo = FALSE}
knitr::kable(
  recon, digits = c(0L, 0L, 6L),
  col.names = c("Method", "Rank or features",
                "Relative Frobenius error"),
  caption = paste0(
    "Reconstruction of the exact ", n_check, " by ", n_check,
    " Gram matrix by each factorisation."
  )
)
```

### What it costs and what it saves

The same independence test at four sample sizes, run exactly and through
both approximations. The data are a synthetic pair with a genuine linear
dependence, so all three tests should reject.

```{r benchmark}
benchmark_hsic <- function(n, m, n_perm = 49L, seed = 1L) {
  set.seed(seed)
  x_bench <- rnorm(n)
  y_bench <- x_bench + rnorm(n, sd = 0.5)
  t_exact <- system.time(
    res_exact <- hsic_test(x_bench, y_bench, n_permutations = n_perm,
                           seed = seed)
  )[["elapsed"]]
  t_nystrom <- system.time(
    res_nystrom <- hsic_test_nystrom(x_bench, y_bench, m = m,
                                     n_permutations = n_perm, seed = seed)
  )[["elapsed"]]
  t_rff <- system.time(
    res_rff <- hsic_test_nystrom(x_bench, y_bench, method = "rff", m = m,
                                 n_permutations = n_perm, seed = seed)
  )[["elapsed"]]
  data.frame(
    n = n, m = m,
    method = c("exact", "nystrom", "rff"),
    elapsed_s = c(t_exact, t_nystrom, t_rff),
    p_value = c(res_exact$p_value, res_nystrom$p_value, res_rff$p_value),
    stringsAsFactors = FALSE
  )
}

bench <- rbind(
  benchmark_hsic(n = 250L, m = 40L),
  benchmark_hsic(n = 500L, m = 60L),
  benchmark_hsic(n = 1000L, m = 80L),
  benchmark_hsic(n = 2000L, m = 100L)
)
```

```{r benchmark-slopes, echo = FALSE}
log_slope <- function(method_name) {
  sub_tab <- bench[bench$method == method_name, ]
  unname(coef(lm(log(elapsed_s) ~ log(n), data = sub_tab))[2L])
}
slope_exact <- log_slope("exact")
slope_nystrom <- log_slope("nystrom")
slope_rff <- log_slope("rff")
```

```{r benchmark-table, echo = FALSE}
knitr::kable(
  bench, digits = c(0L, 0L, 0L, 3L, 4L),
  col.names = c("n", "Rank m", "Method", "Elapsed (s)", "p-value"),
  caption = paste0(
    "Elapsed time and verdict for the exact test and the two ",
    "approximations, at ", length(unique(bench$n)),
    " sample sizes, 49 permutations each."
  )
)
```

### The figure: where the two curves separate

```{r fig-scaling, eval = has_ggplot2, fig.cap = "Elapsed seconds against sample size for the exact HSIC test and the two low-rank approximations, on logarithmic axes. The exact curve rises with the square of the sample size; the approximate curves rise close to linearly, so the gap widens with every doubling."}
ggplot2::ggplot(
  bench, ggplot2::aes(x = n, y = elapsed_s, colour = method)
) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_viridis_d(name = "method", end = 0.85) +
  ggplot2::labs(
    x = "observations n (log scale)",
    y = "elapsed time in seconds (log scale)",
    title = "Cost of one HSIC test as the sample grows"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-scaling-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the scaling figure is",
    "skipped.\n")
```

## Read

Both factorisations reconstruct the exact Gram matrix closely. Nystrom at
rank `r recon$rank[recon$method == "nystrom"]`, one below the sample
size, has a relative Frobenius error of
`r formatC(recon$rel_err[recon$method == "nystrom"], format = "e",
   digits = 2L)`; Random Fourier
Features at `r recon$rank[recon$method == "rff"]` features has
`r signif(recon$rel_err[recon$method == "rff"], 3)`. The two get there
differently: Nystrom converges as its rank approaches the sample size,
Random Fourier Features as the feature count grows, and neither is exact.

The timing grid is the argument. At `r min(bench$n)` observations the
exact test takes
`r signif(bench$elapsed_s[bench$n == min(bench$n) &
  bench$method == "exact"], 2)` seconds and the difference hardly
matters. At `r max(bench$n)` observations it takes
`r signif(bench$elapsed_s[bench$n == max(bench$n) &
  bench$method == "exact"], 3)` seconds against
`r signif(bench$elapsed_s[bench$n == max(bench$n) &
  bench$method == "nystrom"], 2)` for Nystrom and
`r signif(bench$elapsed_s[bench$n == max(bench$n) &
  bench$method == "rff"], 2)` for Random Fourier Features, a speed-up of
about
`r round(bench$elapsed_s[bench$n == max(bench$n) &
  bench$method == "exact"] / bench$elapsed_s[bench$n == max(bench$n) &
  bench$method == "nystrom"])`
times. The figure shows the mechanism rather than the numbers. Fitting a
straight line to each curve on logarithmic axes gives a slope of
`r round(slope_exact, 2)` for the exact test against
`r round(slope_nystrom, 2)` for Nystrom and `r round(slope_rff, 2)` for
Random Fourier Features: the exact test tracks the square law its Gram
matrix implies, and the two approximations grow substantially more
slowly, so the gap is not a constant factor but one that widens with
every doubling.

The verdicts agree.
`r sum(bench$method == "exact")` sample sizes, three methods each, and
every p-value is
`r signif(unique(bench$p_value), 3)`, which is the smallest value 49
permutations can express. That the three methods agree at the
permutation floor shows the approximation did not lose the signal; it is
not a demonstration that they would agree on a borderline case, which
would need a design near the threshold and many more permutations.

## Limits

The reconstruction check is run at 80 observations with Nystrom at
near-full rank, which is the easiest case there is; at the ranks the
timing grid uses, 40 to 100 against samples of 250 to 2000, the
reconstruction error is larger and is not measured here. The timing grid
is a single run per cell on one machine and one core, so the numbers are
indicative and not a benchmark: the ratios are the durable part, the
absolute seconds are not. The two approximations also change the
estimator. The exact `hsic_test()` uses the unbiased HSIC estimator of
Gretton and colleagues (2008); the low-rank path uses the biased form
$n^{-2} \, \mathrm{tr}(H K_x H K_y)$, because that is what factors
cleanly through a low-rank approximation. The bias is of order `1 / n`, which is
negligible in exactly the large-sample regime where these approximations
are worth using, and is not negligible at the small sample sizes where
the exact test is affordable anyway. Finally, `hsic_test_nystrom()`
recomputes its factorisation on every call, so repeated tests against the
same covariate cannot yet share one cached factor even though
`nystrom_factor()` and `rff_features()` expose it.

## Notes on practice

**Choosing the method.** Below about 500 observations the exact test is
fast enough and there is no reason to approximate. Between 500 and 5000,
Nystrom works for any kernel; a rank in the range 50 to 200 is the usual
choice. Above 5000 with an RBF kernel, Random Fourier Features is the
cheapest option because its projection is data-independent; 200 to 1000
features is usually adequate for a verdict.

**Choosing the rank.** A rank of about twice the square root of the
sample size is a defensible starting point for Nystrom. When precision
matters, measure the reconstruction error on a small subsample as the
first section of this vignette does.

**Choosing the feature count.** The variance of a Random Fourier Feature
approximation falls as one over the feature count and its bias is zero in
expectation, so more features are always more accurate and cost more
memory.

**Reproducibility.** Both factorisations are randomised. Pass `seed` for
deterministic output; without it, two runs of the same test give
different numbers.

## What to read next

*Getting started with kernR* introduces the exact HSIC test these
approximations stand in for. *Which parameters change the shape of the
output distribution?* explains why the total-order sensitivity indices do
not yet use these factorisations. *Are these draws calibrated, and do the
engines agree?* covers `ksd_test_nystrom()` and
`concordance_test_nystrom()`, the same idea applied to the
goodness-of-fit and concordance tests.

## References

- Gretton, A., Fukumizu, K., Teo, C. H., Song, L., Scholkopf, B., &
  Smola, A. J. (2008). A kernel statistical test of independence.
  *Advances in Neural Information Processing Systems*, 20.
- Gretton, A., Bousquet, O., Smola, A., & Scholkopf, B. (2005).
  Measuring statistical dependence with Hilbert-Schmidt norms.
  *Algorithmic Learning Theory*, 63-77.
- Rahimi, A., & Recht, B. (2007). Random features for large-scale kernel
  machines. *Advances in Neural Information Processing Systems*, 20.
- Williams, C. K. I., & Seeger, M. (2001). Using the Nystrom method to
  speed up kernel machines. *Advances in Neural Information Processing
  Systems*, 13.

## Reproduce

Seed `1` for the reconstruction check and for every cell of the timing
grid, passed to the factorisations and to the permutation null alike.
Timings measured on macOS aarch64, R 4.5.2, one core. Package versions
follow.

```{r session-info}
sessionInfo()
```
