---
title: "Which parameters are worth calibrating?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Which parameters are worth calibrating?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~1 s under rmarkdown::render(); macOS aarch64, R 4.5.2, one
     core. One 80-point Latin-hypercube design, four parameters against two
     outputs, screened at 299 permutations. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(kernR)
```

## Why

A modeller is about to spend a fortnight of compute on calibration.
*Every realisation of the ensemble smoother is a full simulator run, and
I have four parameters I could put in the prior. If one of them does not
move the outputs I care about, every realisation spent exploring it is
wasted, and worse, it widens the posterior for no reason. Before I start
the calibration, which parameters actually reach the outputs?* This
vignette answers that with a pre-calibration screen: a Latin-hypercube
design over the parameter ranges, one cheap pass through the simulator,
and a kernel independence test between every parameter and every output.

## What

The screen is three verbs. `lhs_design()` builds a Latin-hypercube design
over a matrix of lower and upper bounds, one row per simulator run, so
the parameter space is covered evenly with far fewer runs than a grid.
The simulator is then run once per design row, producing a matrix of
outputs.

`hsic_identifiability()` takes the design and the outputs and tests every
parameter against every output with the Hilbert-Schmidt Independence
Criterion. HSIC is the right statistic here because it asks a broader
question than a variance-based Sobol index: *does this parameter
influence the distribution of the output at all?* A parameter that
changes an output's spread or its tail while leaving its variance intact
is invisible to a variance budget and visible to HSIC. The test returns
one statistic and one permutation p-value per parameter-output pair,
adjusts the p-values across the whole grid (Benjamini-Hochberg by
default), and reports each parameter as identifiable or not at `alpha`.
`as.data.frame()` flattens the grid for reporting; `plot()` draws the
package's own ranked diagnostic.

The identifiable subset is what a calibration prior should contain, and
`PESTO::pesto_ies()` is where it goes next.

## Do

### A stand-in simulator

Running APSIM inside a vignette is not practical, so this is a synthetic
stub with four parameters whose roles are known by construction: `slope`
drives yield linearly, `curvature` drives biomass quadratically, `weak`
has a small linear effect on biomass, and `inert` does nothing at all.

```{r stub-simulator}
stub_apsim <- function(theta) {
  yield <- 1.5 * theta[, "slope"] +
    0.3 * rnorm(nrow(theta), sd = 0.1)
  biomass <- theta[, "curvature"]^2 +
    0.4 * theta[, "weak"] +
    rnorm(nrow(theta), sd = 0.1)
  cbind(yield = yield, biomass = biomass)
}
```

### The design

```{r design}
set.seed(1L)
bounds <- rbind(
  slope     = c(0.0, 2.0),
  curvature = c(-1.0, 1.0),
  weak      = c(0.0, 1.0),
  inert     = c(0.0, 1.0)
)
design <- lhs_design(n = 80L, bounds = bounds, seed = 11L)
head(round(design, 3L))
```

### One pass through the simulator, then the screen

```{r screen}
outputs <- stub_apsim(design)
fit <- hsic_identifiability(
  theta = design, y = outputs, alpha = 0.05, p_adjust = "BH",
  n_permutations = 299L, seed = 11L
)
fit
```

```{r screen-table, echo = FALSE}
screen_df <- as.data.frame(fit)
knitr::kable(
  screen_df, digits = c(0L, 0L, 4L, 4L, 4L),
  col.names = c("Parameter", "Output", "HSIC statistic", "p-value",
                "Adjusted p-value"),
  caption = paste0(
    "Every parameter against every output: ", nrow(screen_df),
    " tests, p-values adjusted across the whole grid by ",
    fit$p_adjust, "."
  )
)
```

### The package's ranked diagnostic

```{r plot-ranked, fig.cap = "The package's own diagnostic: maximum HSIC statistic per parameter across the two outputs, with the identifiability threshold marked."}
plot(fit)
```

### The figure: the dependence the index is measuring

```{r fig-grid, eval = has_ggplot2, fig.height = 5.5, fig.cap = "Each design column against each simulator output, one panel per parameter-output pair. The screen's verdict is a number per panel; the panels show why. Yield rises linearly with slope, biomass traces a parabola in curvature and a shallow slope in weak, and every panel involving inert is a formless cloud."}
grid_df <- do.call(rbind, lapply(colnames(design), function(p1) {
  do.call(rbind, lapply(colnames(outputs), function(o1) {
    data.frame(
      parameter = p1, output = o1,
      value = design[, p1], response = outputs[, o1],
      stringsAsFactors = FALSE
    )
  }))
}))
ggplot2::ggplot(grid_df, ggplot2::aes(x = value, y = response)) +
  ggplot2::geom_point(colour = "#3b528b", alpha = 0.7, size = 1.3) +
  ggplot2::facet_grid(output ~ parameter, scales = "free") +
  ggplot2::labs(
    x = "parameter value (design units)",
    y = "simulator output (arbitrary units)",
    title = "What the screen is looking at"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-grid-skip, eval = !has_ggplot2, echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the parameter-output",
    "grid figure is skipped.\n")
```

### Handing the survivors to the calibration

```{r handoff}
identifiable_params <- names(fit$identifiable)[fit$identifiable]
identifiable_params
```

Those names are the parameter set to pass to `PESTO::pesto_ies()` as the
calibration prior; the parameters left out are the ones that would have
consumed realisations without moving the outputs. The manifest contract
that carries a PESTO ensemble on to the rest of the orchestra is
documented in PESTO itself, and the vignette *Did the management change
shift the simulated distribution?* shows a pair of those manifests being
compared.

## Read

The screen recovers the construction. `slope` and `curvature` are ranked
`r match("slope", names(sort(fit$max_statistic, decreasing = TRUE)))`
and
`r match("curvature", names(sort(fit$max_statistic, decreasing = TRUE)))`
by maximum HSIC, with maximum
HSIC statistics of `r signif(fit$max_statistic[["slope"]], 3)` and
`r signif(fit$max_statistic[["curvature"]], 3)`, and their smallest
adjusted p-values are
`r signif(min(screen_df$p_value_adjusted[
  screen_df$parameter == "slope"]), 3)` and
`r signif(min(screen_df$p_value_adjusted[
  screen_df$parameter == "curvature"]), 3)` against a permutation floor
of `r round(1 / (fit$n_permutations + 1), 4)`. Each drives one output
almost deterministically, and the corresponding panels of the figure are
a straight line and a parabola. `weak` returns a maximum statistic of
`r signif(fit$max_statistic[["weak"]], 3)` with a minimum adjusted
p-value of `r signif(min(screen_df$p_value_adjusted[
  screen_df$parameter == "weak"]), 3)`, and the screen calls it
`r if (fit$identifiable[["weak"]]) "identifiable" else "not identifiable"`
at `alpha` = `r fit$alpha`: its effect on biomass is real but small
against the quadratic term, which is what the shallow trend in its panel
shows. `inert` returns
`r signif(fit$max_statistic[["inert"]], 3)` with a minimum adjusted
p-value of `r signif(min(screen_df$p_value_adjusted[
  screen_df$parameter == "inert"]), 3)` and is not identifiable, as
built.

`r length(identifiable_params)` of the `r length(fit$identifiable)`
parameters survive the screen. Sending only those into the calibration
is the saving: on a simulator where each realisation is a full model run,
the parameters left out would have consumed the same budget as the
parameters that matter and returned a wider posterior for it.

## Limits

The simulator here is a stub, chosen so that the right answer is known
and the vignette runs in seconds; a real crop model has correlated
parameters, interactions and outputs on very different scales, and the
screen inherits all of that. This is a *first-order* screen: it tests
each parameter against each output marginally, so a parameter that acts
only through an interaction can pass unnoticed, which is exactly the case
the total-order indices in *Which parameters change the shape of the
output distribution?* are built for. A parameter declared not
identifiable is not shown to be inert; it is shown to have no detectable
effect at this design size, on these outputs, at this noise level, and a
larger design can promote it. The reverse holds too: with many
parameters and many outputs, unadjusted p-values would manufacture
identifiable parameters, which is why the grid is adjusted by default.
The screen also says nothing about the *sign* or the shape of an effect,
only that one is present.

## Notes on practice

**Design size.** The number of design rows should comfortably exceed the
number of parameters; `n` at least ten times the parameter count is a
workable starting point at agricultural-systems scale.

**Permutations.** The smallest p-value a permutation null can express is
`1 / (n_permutations + 1)`, so testing at `alpha` = 0.05 needs at least
199 permutations before adjustment; 199 to 499 covers most screens.

**Multiple testing.** With `p` parameters and `q` outputs the screen runs
`p` times `q` tests, and the default Benjamini-Hochberg adjustment
controls the false discovery rate across that whole grid.

**Cost.** Kernel matrices are reused across the grid, so construction
costs order `(p + q) n^2` and the permutation null costs order
`p q B n^2` for `B` permutations.

## What to read next

*Which parameters change the shape of the output distribution?* is the
same kernel dependence measure used as a graded sensitivity index rather
than a pass-or-fail screen, and adds total-order indices for interaction.
*Scaling HSIC to large ensembles* covers what to do when the design grows
past the point where an exact Gram matrix is affordable. *Did the
management change shift the simulated distribution?* picks the story up
after calibration, when two forward ensembles are to be compared.

## References

- Da Veiga, S. (2015). Global sensitivity analysis with dependence
  measures. *Journal of Statistical Computation and Simulation*, 85(7),
  1283-1305.
- Gretton, A., Fukumizu, K., Teo, C. H., Song, L., Scholkopf, B., &
  Smola, A. J. (2008). A kernel statistical test of independence.
  *Advances in Neural Information Processing Systems*, 20.
- Hu, R., Sejdinovic, D., & Evans, R. J. (2024). A kernel test for causal
  association via noise contrastive backdoor adjustment. *Journal of
  Machine Learning Research*, 25(160), 1-56.

## Reproduce

Seed `1` for the simulator noise and `seed = 11L` for both the
Latin-hypercube design and the permutation null. Package versions follow.

```{r session-info}
sessionInfo()
```
